### 2013-12-02 inst/doc/hypotheticalExample_vignette.Rmd
generated vignette when we have added noisy genes

<!--
  %\VignetteEngine{knitr}
  %\VignetteIndexEntry{hypothetical example}
  %\VignetteDepends{categoryComparePaperRev, ggplot2}
-->

# Hypothetical Example

To demonstrate the utility of the general `categoryCompare` approach, we will construct an artificial data set based on the Gene Ontology. We will attempt to demonstrate that differences in the **set** or **list** based comparisons are also dependent on the number of items annotated to a particular term.

## GO Terms

```{r nGO}
nGO <- 20
```

We will make use of three sets of `r nGO` GO terms from *H. sapiens*. For each set, we want differing numbers of genes annotated to them.

### Num Annotations Dist

Let's first look at the distribution of number of genes annotated to the GO terms in **biological process**

```{r goAnnotations}
library(org.Hs.eg.db)
library(GO.db)
library(ccPaperRev)
hsGO <- as.list(org.Hs.egGO2ALLEGS)

goOntology <- Ontology(names(hsGO))
goBP <- names(goOntology)[goOntology == "BP"]
hsGO <- hsGO[goBP]
hsGO <- lapply(hsGO, unique)
universeGenes <- unique(unlist(hsGO))
```

```{r goCountDist}
library(ggplot2)
hsGO_count <- sapply(hsGO, length)
hsGO_count <- data.frame(count=hsGO_count)
ggplot(hsGO_count, aes(x=count)) + geom_bar(binwidth=10) + xlim(0, 2000) + ylim(0, 500)
```

### Set up groups

```{r defineMin}
minGO <- 10
```

Let us define a minimum number of genes that need to be annotated to a GO term (`r minGO`), and then divide the GO terms into groups that we can subsequently take random samples from.

```{r defineGroupLimits}
grpLow <- c(10, 100)
grpMed <- c(250, 500)
grpHi <- c(500, 1500)
```

```{r sampleGroups}
set.seed(271113) # for reproducibility
nGO <- 20

GO_low <- limitedRandomSample(hsGO_count, grpLow, nGO)
GO_med <- limitedRandomSample(hsGO_count, grpMed, nGO)
GO_hi <- limitedRandomSample(hsGO_count, grpHi, nGO)
```

### Get Genes

Now, for all of these we need to generate **two** samples of genes from them to comprise our **differentially expressed** (DE) sets from the genome, representing two different expression experiments.

How do we select these two sets of genes?

* define the maximum number of diff genes for the set
* Randomly sort the GO terms
* For each GO term in turn (i.e. iterating over the GO terms):
  * take a random fraction of genes for that GO term (sampling from an exponentially decaying distribution of possibilites), adding the requisite number of genes
  * If all the genes for a GO term get taken, remove that GO term from future consideration
* Continue iterating through until the total number of genes is reached

What we expect from this is that in both datasets, the GO terms with lower numbers of genes annotated will have large fractions of their annotations present, while GO terms with more and more genes will have lower fractions, and there will be less overlap between the two datasets.

These expectations should be checked.

```{r generateSamples}
useGO <- c(GO_low, GO_med, GO_hi)
goList <- list(low=GO_low, med=GO_med, hi=GO_hi)
nGene <- 1000
sample1_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
sample2_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
```

In addition to the genes that are annotated to the GO terms of interest, we will also add in genes that have no relation to the GO terms under consideration, these would be considered **noise** genes.

```{r noiseGenes}
nNoise <- 500
gene2HsGO <- reverseSplit(hsGO)
not_useGO <- sapply(gene2HsGO, function(x){
  sum(x %in% useGO) == 0
})
noiseGenes <- names(not_useGO)[not_useGO]
noiseGenes <- sample(noiseGenes, nNoise)

# check that we did this right, the fraction should not change after adding noise genes
useGO_frac <- sapply(hsGO[useGO], function(x){
  length(intersect(x, sample1_org)) / length(x)
})
sample1 <- c(sample1_org, noiseGenes)

useGO_fracNoise <- sapply(hsGO[useGO], function(x){
  length(intersect(x, sample1)) / length(x)
})
sample2 <- c(sample2_org, noiseGenes)
plot(useGO_frac, useGO_fracNoise)
```

```{r checkAssumptions}
# check assumptions
useGO_counts <- sapply(hsGO[useGO], length)
useGO_frac <- sapply(hsGO[useGO], function(x){
  length(intersect(x, sample1)) / length(x)
})
checkGO <- data.frame(counts=useGO_counts, frac=useGO_frac, sample="1")
useGO_frac2 <- sapply(hsGO[useGO], function(x){
  length(intersect(x, sample2)) / length(x)
})
checkGO <- rbind(checkGO, data.frame(counts=useGO_counts, frac=useGO_frac2, sample="2"))
ggplot(checkGO, aes(x=counts, y=frac, color=sample)) + geom_point()
```

From the plot above, it appears that we meet expectations. In addition, we manage to share only `r length(intersect(sample1, sample2))` genes between the two samples.

What about the degree of overlap between the two samples as a function of size or fraction?

```{r overlapDist}
overlapDist <- sapply(useGO, function(x){
  baseAnn <- hsGO[[x]]
  t1 <- intersect(baseAnn, sample1)
  t2 <- intersect(baseAnn, sample2)
  c(length(baseAnn), length(t1), length(t2), length(intersect(t1,t2)))
})
overlapDist <- t(overlapDist)
overlapFrac <- c(overlapDist[,2] / overlapDist[,1], overlapDist[,3]/overlapDist[,1], overlapDist[,4]/overlapDist[,1])
overlapFrac <- data.frame(count=rep(overlapDist[,1], 3), 
                          frac=overlapFrac, 
                          id=c(rep("s1", 60), rep("s2", 60), rep("s1.s2", 60)))
overlapFrac
ggplot(overlapFrac, aes(x=count, y=frac, color=id)) + geom_point()
```

## Do the comparisons

### categoryCompare method

#### Enrichment

```{r ccInit}
s1GO <- new("GOHyperGParamsCC", geneIds=sample1, universeGeneIds=universeGenes, ontology="BP", fdr=0, annotation="org.Hs.eg.db")
s1_enrich <- hyperGTestCC(s1GO)

s2GO <- new("GOHyperGParamsCC", geneIds=sample2, universeGeneIds=universeGenes, ontology="BP", fdr=0, annotation="org.Hs.eg.db")
s2_enrich <- hyperGTestCC(s2GO)
```

#### Basic Results

```{r pvalueCut}
pval <- 0.05
minP <- -1 * log10(pval)
```

For the two-sample comparison, we take the minimum p-value of the two sets of results, and then compare with `r -1*log10(pval)`, anything above that is considered significant. We are using **log**'s because they are easier to visualize.

```{r compRes}
s1_goP <- -1 * log10((s1_enrich@pvalues)[useGO])
s2_goP <- -1 * log10((s2_enrich@pvalues)[useGO])

s_minP <- apply(cbind(s1_goP, s2_goP), 1, min)
sum(s_minP >= minP)
```

### Intersection Method

```{r intersectMethod}
comGO <- new("GOHyperGParamsCC", geneIds=intersect(sample1, sample2), universeGeneIds=universeGenes, ontology="BP", fdr=0, annotation="org.Hs.eg.db")
com_enrich <- hyperGTestCC(comGO)
com_goP <- -1 * log10((com_enrich@pvalues)[useGO])
```

### Compare Methods

To compare the methods, we will start with the `r length(useGO)` terms in our low, med and hi groups, and compare which ones and how many show up.

```{r numericComparison}
sum(s_minP >= minP)
sum(com_goP >= minP)

invisible(lapply(goList, function(x){
  print(sum(s_minP[x] >= minP))
  print(sum(com_goP[x] >= minP))
}))
```

We will also plot a histogram to make this more apparent.

```{r plotHistogram}
s_count <- sapply(goList, function(x){
  sum(s_minP[x] >= minP)
})
com_count <- sapply(goList, function(x){
  sum(com_goP[x] >= minP)
})

countData <- data.frame(count=c(s_count, com_count), grp=rep(names(goList), 2), type=rep(c("s", "com"), each=3))
countData$grp <- factor(countData$grp, levels=c("low", "med", "hi"), ordered=TRUE)
ggplot(countData, aes(x=grp, y=count, fill=type)) + geom_bar(stat="identity", position="dodge") + ylim(0, 20)
```

The histogram suggests that there may be a relationship between the difference in p-values and the proportion of genes annotated to a GO term. Let's check.

As we are using logged p-values, we will simply take the difference of p-values and plot them. We expect to see higher p-values (i.e. positive differences) for the inter-group comparison *vs* the intersection of genes.

```{r plotPDiffs}
pDiff <- data.frame(pdiff = s_minP - com_goP,
                    annFrac = useGO_frac,
                    grp = rep(names(goList), each=20))
pDiff$grp <- factor(pDiff$grp, levels=c("low", "med", "hi"), ordered=T)

ggplot(pDiff, aes(x=annFrac, y=pdiff)) + geom_point()
```

Doesn't appear to be anything discernible for the fraction alone. What if it is split up across the groups?

```{r plotPDiffsGroup}
ggplot(pDiff, aes(x=annFrac, y=pdiff)) + geom_point() + facet_grid(. ~ grp, scales="free_x")
```

Hmm, well, which ones were actually significant in different cases?

* Sig. using both methods
* Sig in category comparison only
* Sig in gene intersection only

```{r plotPDiffsMethod}
pDiff$sigGroup <- "none"
sigBoth <- names(s_minP)[(s_minP >= 1.3) & (com_goP >= 1.3)]
pDiff[sigBoth, "sigGroup"] <- "both"
sigS <- names(s_minP)[(s_minP >= 1.3) & (com_goP < 1.3)]
pDiff[sigS, "sigGroup"] <- "s"

ggplot(pDiff, aes(x=annFrac, y=pdiff, color=sigGroup)) + geom_point() + facet_grid(. ~ grp, scales="free_x")
```

Now we can see that:

1. There is nothing from combining gene lists that is significant while combining enrichment results is not.
2. In general, there is a much better p-value from the combined enrichment than from combined gene lists.

### 2013-12-02 inst/doc/hypotheticalExample_vignette.R

@@ -1,0 +1,177 @@
+
+## ----nGO-----------------------------------------------------------------
+nGO <- 20
+
+
+## ----goAnnotations-------------------------------------------------------
+library(org.Hs.eg.db)
+library(GO.db)
+library(ccPaperRev)
+hsGO <- as.list(org.Hs.egGO2ALLEGS)
+
+goOntology <- Ontology(names(hsGO))
+goBP <- names(goOntology)[goOntology == "BP"]
+hsGO <- hsGO[goBP]
+hsGO <- lapply(hsGO, unique)
+universeGenes <- unique(unlist(hsGO))
+
+
+## ----goCountDist---------------------------------------------------------
+library(ggplot2)
+hsGO_count <- sapply(hsGO, length)
+hsGO_count <- data.frame(count=hsGO_count)
+ggplot(hsGO_count, aes(x=count)) + geom_bar(binwidth=10) + xlim(0, 2000) + ylim(0, 500)
+
+
+## ----defineMin-----------------------------------------------------------
+minGO <- 10
+
+
+## ----defineGroupLimits---------------------------------------------------
+grpLow <- c(10, 100)
+grpMed <- c(250, 500)
+grpHi <- c(500, 1500)
+
+
+## ----sampleGroups--------------------------------------------------------
+set.seed(271113) # for reproducibility
+nGO <- 20
+
+GO_low <- limitedRandomSample(hsGO_count, grpLow, nGO)
+GO_med <- limitedRandomSample(hsGO_count, grpMed, nGO)
+GO_hi <- limitedRandomSample(hsGO_count, grpHi, nGO)
+
+
+## ----generateSamples-----------------------------------------------------
+useGO <- c(GO_low, GO_med, GO_hi)
+goList <- list(low=GO_low, med=GO_med, hi=GO_hi)
+nGene <- 1000
+sample1_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
+sample2_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
+
+
+## ----noiseGenes----------------------------------------------------------
+nNoise <- 500
+gene2HsGO <- reverseSplit(hsGO)
+not_useGO <- sapply(gene2HsGO, function(x){
+  sum(x %in% useGO) == 0
+})
+noiseGenes <- names(not_useGO)[not_useGO]
+noiseGenes <- sample(noiseGenes, nNoise)
+
[…]

### 2013-12-02 R/hypotheticalDataGeneration.R
new functions for doing fractions and multiple (simple) go enrichments

@@ -79,3 +79,61 @@ sampleTerms <- function(terms, term2gene, nGenes=2000, expRate=4){
   }
   return(sampleGenes)
 }
+
+#' calculate fraction of genes annotated to GO terms
+#' 
+#' For the supplied list of GO 2 gene annotations, and lists of genes, calculate the what fraction of a GO term is in the gene list.
+#' 
+#' @details returns a \code{data.frame} with the total count, and fraction for each of the gene lists supplied, suitable for plotting in \code{ggplot2}.
+#' @param go2gene list of GO terms and genes they annotate
+#' @param geneList list of character vectors of genes
+#' @export
+#' @return data.frame
+calcFraction <- function(go2gene, geneList){
+  nGO <- length(go2gene)
+  
+  goSizes <- sapply(go2gene, length)
+  goNames <- names(go2gene)
+  
+  nList <- length(geneList)
+  
+  goFrac <- lapply(geneList, function(inGenes){
+    sapply(go2gene, function(inGO){
+      length(intersect(inGenes, inGO)) / length(inGO)
+    })
+  })
+  goFrac <- stack(goFrac)
+  names(goFrac) <- c("frac", "genelist")
+  goFrac$goid <- rep(goNames, nList)
+  goFrac$size <- rep(goSizes, nList)
+  return(goFrac)
+}
+
+#' hyperGOMultiEnrichment
+#' 
+#' hypergeometric (Fisher's Exact Test) enrichment calculations for multiple lists of genes, with common background, as well as the intersection of the gene lists.
+#' 
+#' @param geneList \code{list} where each entry is a gene list
+#' @param universe \code{character} vector of the background (assumed to be common across all)
+#' @param ontology which GO ontology to use (default is "BP")
+#' @param annotation what is the source of annotation (default is "org.Hs.eg.db")
+#' @importClassesFrom categoryComparePaperRev GOHyperGParamsCC
+#' @importFrom categoryComparePaperRev hyperGTestCC
+#' @return \code{list}, see Details
+#' @details the object returned will have an object for each sample list, as well as the one generated from the intersection
+#' @export
+hyperGOMultiEnrichment <- function(geneList, universe, ontology="BP", annotation="org.Hs.eg.db"){
+  tmpIntersect <- geneList[[1]]
+  
+  nList <- length(geneList)
+  for (iList in 1:nList){
[…]

### 2013-12-03 R/hypotheticalDataGeneration.R
added function for getting out p-values

@@ -137,3 +137,23 @@ hyperGOMultiEnrichment <- function(geneList, universe, ontology="BP", annotation
     outHyper <- hyperGTestCC(inHyper)
   })
 }
+
+#' get p-values from multiple enrichments
+#' 
+#' @param geneListNames the names to combine and take the minimum
+#' @param hyperEnrichList the set of hypergeometric enrichment results
+#' @param useTerms
+#' @export
+#' @return \code{data.frame}, see details
+#' @details the \code{data.frame} returned has the p-values for each GO term from both the set based and list based results. The set-based value is the minimum value of all the set enrichments for that GO term, log transformed.
+pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList){
+  allValues <- lapply(hyperEnrichList, function(inEnrich){
+    -1 * log10(inEnrich@pvalues[useTerms])
+  })
+  
+  setValues <- do.call(cbind, allValues[geneListNames])
+  setValues <- apply(setValues, 1, min)
+  
+  outValues <- data.frame(set = setValues, list = allValues$intersect)
+  return(outValues)
+}

### 2013-12-03 R/hypotheticalDataGeneration.R
modifying pvaluesMultiEnrich to do log optionally

@@ -143,12 +143,17 @@ hyperGOMultiEnrichment <- function(geneList, universe, ontology="BP", annotation
 #' @param geneListNames the names to combine and take the minimum
 #' @param hyperEnrichList the set of hypergeometric enrichment results
 #' @param useTerms
+#' @param log return original or log-transformed values
 #' @export
 #' @return \code{data.frame}, see details
 #' @details the \code{data.frame} returned has the p-values for each GO term from both the set based and list based results. The set-based value is the minimum value of all the set enrichments for that GO term, log transformed.
-pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList){
+pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRUE){
   allValues <- lapply(hyperEnrichList, function(inEnrich){
-    -1 * log10(inEnrich@pvalues[useTerms])
+    tmpVal <- inEnrich@pvalues[useTerms]
+    if (log){
+      tmpVal <- -1 * log10(tmpVal)
+    }
+    tmpVal
   })
   
   setValues <- do.call(cbind, allValues[geneListNames])
@@ -157,3 +162,9 @@ pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList){
   outValues <- data.frame(set = setValues, list = allValues$intersect)
   return(outValues)
 }
+
+#' calculate difference and significance
+#' 
+#' @param pvalueData the \code{data.frame} of values
+#' @param pCutoff the p-value cutoff to use for significance
+#' @param log

### 2013-12-03 R/hypotheticalDataGeneration.R
added function for doing differences and determining significant list

@@ -167,4 +167,25 @@ pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRU
 #' 
 #' @param pvalueData the \code{data.frame} of values
 #' @param pCutoff the p-value cutoff to use for significance
-#' @param log
+#' @param log the values are logged, and p-value should be too
+#' @return \code{data.frame}
+#' @export
+pvalueDiffSig <- function(pvalueData, pCutoff=0.05, log=TRUE){
+  if (log){
+    pCutoff <- -1 * log10(pCutoff)
+  }
+  
+  pvalueData$diff <- pvalueData$set - pvalueData$list
+  pvalueData$sigState <- "none"
+  
+  sigSets <- list(both = (pvalueData$set >= pCutoff) & (pvalueData$list >= pCutoff),
+                  set = (pvalueData$set >= pCutoff) & (pvalueData$list < pCutoff),
+                  list = (pvalueData$set < pCutoff) & (pvalueData$list >= pCutoff))
+  
+  invisible(lapply(names(sigSets), function(inSig){
+    if (sum(sigSets[[inSig]]) > 0){
+      pvalueData$sigState[sigSets[[inSig]]] <<- inSig
+    }
+  }))
+  return(pvalueData)
+}

### 2013-12-04 inst/doc/hypotheticalExample_vignette.Rmd
new generated version of the vignette

@@
 We will attempt to demonstrate that differences in the **set** or **list** based comparisons are also dependent on the number of items annotated to a particular term.
+
+## Definitions
+
+**Set** based: calculations performed independently on each *set* of features, and then the results are combined.
+
+**List** based: calculations performed on the intersected *list* of two sets of features.
 
@@
 
+### Checking our expectations about the fractions of GO terms.
+
+```{r checkFractions}
+goFractions <- calcFraction(hsGO[useGO], list(sample1=sample1_org, sample2=sample2_org))
+ggplot(goFractions, aes(x=size, y=frac, color=genelist)) + geom_point()
+```
+
+### Do calculations
+
+For each `geneList`, we will do the GO enrichment calculations, and then get the results. We also do the intersection of all the `geneList`'s, and do the calculations again.
+
+```{r runCalcs}
+samples_noNoise <- list(sample1 = sample1_org, sample2 = sample2_org)
+go_noNoise <- hyperGOMultiEnrichment(samples_noNoise, universe=universeGenes)
+```
+
+Now get the p-values out (`-1 * log10(pvalue)`).
+For each of our test GO terms, we take the minimum p-value from the results of the **set** calculations, and also the values from the **list** calculations.
+
+```{r noNoisePvalues}
+noNoise_pvalues <- pvaluesMultiEnrich(c("sample1", "sample2"), useGO, go_noNoise)
+
+noNoise_pvalues <- pvalueDiffSig(noNoise_pvalues, pCutoff=0.05, log=TRUE)
+```
+
+Finally, lets add in our values of the fraction, total genes, and the class (**low**, **med**, **hi**) so we can do some fancy plotting stuff with it.
+
+```{r noNoiseAddInfo}
+sizeClass <- rep(c("low", "med", "hi"), each=20)
+names(sizeClass) <- useGO
+noNoise_pvalues$sizeClass <- sizeClass
+noNoise_pvalues$size <- goFractions[1:60,4]
+noNoise_pvalues$frac <- goFractions$frac[1:60]
+```
+
+Now we will plot the difference in log-p-values (where we have **set** - **list**), so that positive values imply that **set** had lower p-values than **list**, based on the *fraction* of genes annotated.
+
+```{r noNoisePlotStuff}
+noNoise_pvalues$sizeClass <- factor(noNoise_pvalues$sizeClass, levels=c("low", "med", "hi"), ordered=TRUE)
+ggplot(noNoise_pvalues, aes(x=frac, y=diff, color=sigState)) + geom_point() + facet_grid(. ~ sizeClass, scales="free_x")
+ggplot(noNoise_pvalues, aes(x=sizeClass, y=diff)) + geom_boxplot() + geom_point()
+```
+
+We can see that it appears in general that **set** has better p-values than **list**, especially in the *med* and *hi* groups. Also, both methods get pretty much the same GO terms as significant.
+
+How many of the GO terms have better p-values in **set** compared to **list**?
+
+```{r noNoiseBetterPVals}
+tapply(noNoise_pvalues$diff, noNoise_pvalues$sizeClass, function(x){sum(x > 0)})
+```
+
+What is the average difference among each group?
+
+```{r noNoiseBetterMean}
+tapply(noNoise_pvalues$diff, noNoise_pvalues$sizeClass, mean)
+```
+
+So ideally, the mean of the full set of differences, combined with how many pass in each case, should adequately quantify how well each of the methods is doing.
+
+## Noise
+
 In addition to the genes that are annotated to the GO terms of interest, we will also add in genes that have no relation to the GO terms under consideration, these would be considered **noise** genes.
+To make the model simple, we will have the **same** genes as noise in **both** samples.
 
@@
 # check that we did this right, the fraction should not change after adding noise genes
-useGO_frac <- sapply(hsGO[useGO], function(x){
-  length(intersect(x, sample1_org)) / length(x)
-})
 sample1 <- c(sample1_org, noiseGenes)
+sample2 <- c(sample2_org, noiseGenes)
 
-useGO_fracNoise <- sapply(hsGO[useGO], function(x){
-  length(intersect(x, sample1)) / length(x)
-})
-sample2 <- c(sample2_org, noiseGenes)
-plot(useGO_frac, useGO_fracNoise)
+samplesNoise <- list(sample1=sample1, sample2=sample2)
+
+goFracNoise <- calcFraction(hsGO[useGO], samplesNoise)
+plot(goFracNoise$frac, goFractions$frac)
 ```
 
-```{r checkAssumptions}
-# check assumptions
-useGO_counts <- sapply(hsGO[useGO], length)
-useGO_frac <- sapply(hsGO[useGO], function(x){
-  length(intersect(x, sample1)) / length(x)
-})
-checkGO <- data.frame(counts=useGO_counts, frac=useGO_frac, sample="1")
-useGO_frac2 <- sapply(hsGO[useGO], function(x){
-  length(intersect(x, sample2)) / length(x)
-})
-checkGO <- rbind(checkGO, data.frame(counts=useGO_counts, frac=useGO_frac2, sample="2"))
-ggplot(checkGO, aes(x=counts, y=frac, color=sample)) + geom_point()
+```{r noisyEnrichment}
+go_noise <- hyperGOMultiEnrichment(samplesNoise, universeGenes)
 ```
 
-From the plot above, it appears that we meet expectations. In addition, we manage to share only `r length(intersect(sample1, sample2))` genes between the two samples.
+```{r noisePvalues}
+noise_pvalues <- pvaluesMultiEnrich(c("sample1", "sample2"), useGO, go_noise)
 
-What about the degree of overlap between the two samples as a function of size or fraction?
+noise_pvalues <- pvalueDiffSig(noise_pvalues, pCutoff=0.05, log=TRUE)
 
-```{r overlapDist}
-overlapDist <- sapply(useGO, function(x){
-  baseAnn <- hsGO[[x]]
-  t1 <- intersect(baseAnn, sample1)
-  t2 <- intersect(baseAnn, sample2)
-  c(length(baseAnn), length(t1), length(t2), length(intersect(t1,t2)))
-})
-overlapDist <- t(overlapDist)
-overlapFrac <- c(overlapDist[,2] / overlapDist[,1], overlapDist[,3]/overlapDist[,1], overlapDist[,4]/overlapDist[,1])
-overlapFrac <- data.frame(count=rep(overlapDist[,1], 3), 
-                          frac=overlapFrac, 
-                          id=c(rep("s1", 60), rep("s2", 60), rep("s1.s2", 60)))
-overlapFrac
-ggplot(overlapFrac, aes(x=count, y=frac, color=id)) + geom_point()
+noise_pvalues$sizeClass <- sizeClass
+noise_pvalues$size <- goFracNoise[1:60,4]
+noise_pvalues$frac <- goFracNoise$frac[1:60]
 ```
 
-## Do the comparisons
-
-### categoryCompare method
-
-#### Enrichment
-
-```{r ccInit}
-s1GO <- new("GOHyperGParamsCC", geneIds=sample1, universeGeneIds=universeGenes, ontology="BP", fdr=0, annotation="org.Hs.eg.db")
-s1_enrich <- hyperGTestCC(s1GO)
-
-s2GO <- new("GOHyperGParamsCC", geneIds=sample2, universeGeneIds=universeGenes, ontology="BP", fdr=0, annotation="org.Hs.eg.db")
-s2_enrich <- hyperGTestCC(s2GO)
+```{r plotSummarizeNoise}
+noise_pvalues$sizeClass <- factor(noise_pvalues$sizeClass, c("low", "med", "hi"), ordered=TRUE)
+ggplot(noise_pvalues, aes(x=frac, y=diff, color=sigState)) + geom_point() + facet_grid(. ~ sizeClass, scales="free_x")
 ```
 
-#### Basic Results
-
-```{r pvalueCut}
-pval <- 0.05
-minP <- -1 * log10(pval)
+```{r summarizeCountsNoise}
+tapply(noise_pvalues$diff, noise_pvalues$sizeClass, function(x){sum(x > 0)})
+tapply(noise_pvalues$diff, noise_pvalues$sizeClass, mean)
 ```
 
-For the two-sample comparison, we take the minimum p-value of the two sets of results, and then compare with `r -1*log10(pval)`, anything above that is considered significant.
-We are using **log**'s because they are easier to visualize.
+Although using gene lists that were perfect (i.e.
+no noise) everything in the **list** method was pretty much still significant, on average **sets** gave much better p-values.
+However, as we add noise to the system in the form of genes that are not annotated to our test GO terms, more terms are significant only in the **sets** method (i.e.
+show up in **both** sets).
 
-```{r compRes}
-s1_goP <- -1 * log10((s1_enrich@pvalues)[useGO])
-s2_goP <- -1 * log10((s2_enrich@pvalues)[useGO])
+## Expectations
 
-s_minP <- apply(cbind(s1_goP, s2_goP), 1, min)
-sum(s_minP >= minP)
-```
+Based on the above results, what I expect we will see if we do a programmatic search over the space is that there is a dependence on both the fraction of DE genes that are noisy, and the fraction of noisy genes that are shared (which really amounts to the same thing, but is a subtle point).
 
-### Intersection Method
+# To Do:
 
-```{r intersectMethod}
-comGO <- new("GOHyperGParamsCC", geneIds=intersect(sample1, sample2), universeGeneIds=universeGenes, ontology="BP", fdr=0, annotation="org.Hs.eg.db")
-com_enrich <- hyperGTestCC(comGO)
-com_goP <- -1 * log10((com_enrich@pvalues)[useGO])
-```
-
-### Compare Methods
-
-To compare the methods, we will start with the `r length(useGO)` terms in our low, med and hi groups, and compare which ones and how many show up.
-
-```{r numericComparison}
-sum(s_minP >= minP)
-sum(com_goP >= minP)
-
-invisible(lapply(goList, function(x){
-  print(sum(s_minP[x] >= minP))
-  print(sum(com_goP[x] >= minP))
-}))
-```
-
-We will also plot a histogram to make this more apparent.
-
-```{r plotHistogram}
-s_count <- sapply(goList, function(x){
-  sum(s_minP[x] >= minP)
-})
-com_count <- sapply(goList, function(x){
-  sum(com_goP[x] >= minP)
-})
-
-countData <- data.frame(count=c(s_count, com_count), grp=rep(names(goList), 2), type=rep(c("s", "com"), each=3))
-countData$grp <- factor(countData$grp, levels=c("low", "med", "hi"), ordered=TRUE)
-ggplot(countData, aes(x=grp, y=count, fill=type)) + geom_bar(stat="identity", position="dodge") + ylim(0, 20)
-```
-
-The histogram suggests that there may be a relationship between the difference in p-values and the proportion of genes annotated to a GO term. Let's check.
-
-As we are using logged p-values, we will simply take the difference of p-values and plot them.
-We expect to see higher p-values (i.e.
-positive differences) for the inter-group comparison *vs* the intersection of genes.
-
-```{r plotPDiffs}
-pDiff <- data.frame(pdiff = s_minP - com_goP,
-                    annFrac = useGO_frac,
-                    grp = rep(names(goList), each=20))
-pDiff$grp <- factor(pDiff$grp, levels=c("low", "med", "hi"), ordered=T)
-
-ggplot(pDiff, aes(x=annFrac, y=pdiff)) + geom_point()
-```
-
-Doesn't appear to be anything discernible for the fraction alone. What if it is split up across the groups?
-
-```{r plotPDiffsGroup}
-ggplot(pDiff, aes(x=annFrac, y=pdiff)) + geom_point() + facet_grid(. ~ grp, scales="free_x")
-```
-
-Hmm, well, which ones were actually significant in different cases?
-
-* Sig. using both methods
-* Sig in category comparison only
-* Sig in gene intersection only
-
-```{r plotPDiffsMethod}
-pDiff$sigGroup <- "none"
-sigBoth <- names(s_minP)[(s_minP >= 1.3) & (com_goP >= 1.3)]
-pDiff[sigBoth, "sigGroup"] <- "both"
-sigS <- names(s_minP)[(s_minP >= 1.3) & (com_goP < 1.3)]
-pDiff[sigS, "sigGroup"] <- "s"
-
-ggplot(pDiff, aes(x=annFrac, y=pdiff, color=sigGroup)) + geom_point() + facet_grid(. ~ grp, scales="free_x")
-```
-
-Now we can see that:
-
-1. There is nothing from combining gene lists that is significant while combining enrichment results is not.
-2. In general, there is a much better p-value from the combined enrichment than from combined gene lists.
+* Check expectations of the gene to GO distribution from a couple of real human data sets
+* Run both the **noNoise** and **noise** 100 times with different sets of GO terms and make sure that this holds.
+* Vary the fraction of noise genes, and also vary the fraction that is common, and find a way to quantify the effect.

### 2013-12-04 inst/doc/hypotheticalExample_vignette.R

@@ -50,6 +50,44 @@ sample1_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
 sample2_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
 
 
+## ----checkFractions------------------------------------------------------
+goFractions <- calcFraction(hsGO[useGO], list(sample1=sample1_org, sample2=sample2_org))
+ggplot(goFractions, aes(x=size, y=frac, color=genelist)) + geom_point()
+
+
+## ----runCalcs------------------------------------------------------------
+samples_noNoise <- list(sample1 = sample1_org, sample2 = sample2_org)
+go_noNoise <- hyperGOMultiEnrichment(samples_noNoise, universe=universeGenes)
+
+
+## ----noNoisePvalues------------------------------------------------------
+noNoise_pvalues <- pvaluesMultiEnrich(c("sample1", "sample2"), useGO, go_noNoise)
+
+noNoise_pvalues <- pvalueDiffSig(noNoise_pvalues, pCutoff=0.05, log=TRUE)
+
+
+## ----noNoiseAddInfo------------------------------------------------------
+sizeClass <- rep(c("low", "med", "hi"), each=20)
+names(sizeClass) <- useGO
+noNoise_pvalues$sizeClass <- sizeClass
+noNoise_pvalues$size <- goFractions[1:60,4]
+noNoise_pvalues$frac <- goFractions$frac[1:60]
+
+
+## ----noNoisePlotStuff----------------------------------------------------
+noNoise_pvalues$sizeClass <- factor(noNoise_pvalues$sizeClass, levels=c("low", "med", "hi"), ordered=TRUE)
+ggplot(noNoise_pvalues, aes(x=frac, y=diff, color=sigState)) + geom_point() + facet_grid(. ~ sizeClass, scales="free_x")
+ggplot(noNoise_pvalues, aes(x=sizeClass, y=diff)) + geom_boxplot() + geom_point()
+
+
+## ----noNoiseBetterPVals--------------------------------------------------
+tapply(noNoise_pvalues$diff, noNoise_pvalues$sizeClass, function(x){sum(x > 0)})
+
+
+## ----noNoiseBetterMean---------------------------------------------------
+tapply(noNoise_pvalues$diff, noNoise_pvalues$sizeClass, mean)
+
+
[…]

### 2013-12-05 R/hypotheticalDataGeneration.R
added new function to get fractions for all go terms annotated to de genes

@@ -189,3 +189,31 @@ pvalueDiffSig <- function(pvalueData, pCutoff=0.05, log=TRUE){
   }))
   return(pvalueData)
 }
+
+#' trim go2gene and get fractions
+#' 
+#' for the cases of examining the "real" data, we need to take a GO 2 gene mapping, get the gene 2 GO mapping and find just those GO terms that actually have genes in the DE list, and finally calculate the fractions.
+#' 
+#' @param go2gene list of GO 2 gene mappings
+#' @param deList the differentially expressed gene list as a \code{character} vector
+#' @importFrom Biobase reverseSplit
+#' @return \code{data.frame} of fractions and counts
+#' @export
+trimGOFrac <- function(go2gene, deList){
+  stopifnot(is.list(go2gene))
+  stopifnot(is.character(deList))
+  
+  go2gene <- lapply(go2gene, unique)
+  
+  gene2go <- reverseSplit(go2gene)
+  gene2go <- gene2go[deList]
+  keepGO <- unique(unlist(gene2go, use.names=F))
+  
+  go2gene <- go2gene[keepGO]
+  
+  goSize <- sapply(go2gene, length)
+  goFrac <- sapply(go2gene, function(x){
+    length(intersect(x, deList)) / length(x)
+  })
+  return(data.frame(size=goSize, frac=goFrac))
+}

### 2013-12-06 R/hypotheticalDataGeneration.R
adding in the lung data as something saved with the package

@@ -217,3 +217,10 @@ trimGOFrac <- function(go2gene, deList){
   })
   return(data.frame(size=goSize, frac=goFrac))
 }
+
+#' @name lung.RData
+#' @title lung.RData
+#' @docType data
+#' @source ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE18nnn/GSE18842/matrix/GSE18842_series_matrix.txt.gz
+#' @details Downloaded, created an \code{ExpressionSet} object using \code{getGEO} on Dec 6, 2013.
+NULL

### 2013-12-06 R/hypotheticalDataGeneration.R
changing any NA values to 0

@@ -158,6 +158,9 @@ pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRU
   
   setValues <- do.call(cbind, allValues[geneListNames])
   setValues <- apply(setValues, 1, min)
+  setValues[is.na(setValues)] <- 0
+  
+  allValues$intersect[is.na(allValues$intersect)] <- 0
   
   outValues <- data.frame(set = setValues, list = allValues$intersect)
   return(outValues)
@@ -217,6 +220,7 @@ trimGOFrac <- function(go2gene, deList){
   })
   return(data.frame(size=goSize, frac=goFrac))
 }
+
 
 #' @name lung.RData
 #' @title lung.RData

### 2013-12-06 R/hypotheticalDataGeneration.R
smarter handling of NA values depending on log or not

@@ -156,11 +156,16 @@ pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRU
     tmpVal
   })
   
+  naVal <- 1
+  if (log){
+    naVal <- 0
+  }
+  
   setValues <- do.call(cbind, allValues[geneListNames])
   setValues <- apply(setValues, 1, min)
-  setValues[is.na(setValues)] <- 0
+  setValues[is.na(setValues)] <- naVal
   
-  allValues$intersect[is.na(allValues$intersect)] <- 0
+  allValues$intersect[is.na(allValues$intersect)] <- naVal
   
   outValues <- data.frame(set = setValues, list = allValues$intersect)
   return(outValues)

### 2013-12-09 R/hypotheticalDataGeneration.R
added names for the case where the GO term was not even tested

@@ -153,6 +153,7 @@ pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRU
     if (log){
       tmpVal <- -1 * log10(tmpVal)
     }
+    names(tmpVal) <- useTerms
     tmpVal
   })
   

### 2013-12-10 R/hypotheticalDataGeneration.R
modified pvaluesMultiEnrich to return both the pvalue and whether or not the go term was even measured

@@ -148,6 +148,12 @@ hyperGOMultiEnrichment <- function(geneList, universe, ontology="BP", annotation
 #' @return \code{data.frame}, see details
 #' @details the \code{data.frame} returned has the p-values for each GO term from both the set based and list based results. The set-based value is the minimum value of all the set enrichments for that GO term, log transformed.
 pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRUE){
+  naVal <- 1
+  if (log){
+    naVal <- 0
+  }
+  
+  # get the p-values
   allValues <- lapply(hyperEnrichList, function(inEnrich){
     tmpVal <- inEnrich@pvalues[useTerms]
     if (log){
@@ -157,11 +163,6 @@ pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRU
     tmpVal
   })
   
-  naVal <- 1
-  if (log){
-    naVal <- 0
-  }
-  
   setValues <- do.call(cbind, allValues[geneListNames])
   setValues <- apply(setValues, 1, min)
   setValues[is.na(setValues)] <- naVal
@@ -169,7 +170,21 @@ pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRU
   allValues$intersect[is.na(allValues$intersect)] <- naVal
   
   outValues <- data.frame(set = setValues, list = allValues$intersect)
-  return(outValues)
+  
+  # get whether it was even measured
+  presentValues <- lapply(hyperEnrichList, function(inEnrich){
+    measuredGO <- names(inEnrich@pvalues)
+    hasGO <- useTerms %in% measuredGO
+    names(hasGO) <- useTerms
+  })
+  
+  initPresent <- logical(length(useTerms))
+  setPresent <- do.call(cbind, presentValues[geneListNames])
+  setPresent <- apply(setPresent, 1, function(x){sum(x) > 0})
+  
+  outPresent <- data.frame(set = setPresent, list = presentValues$intersect)
+  
+  return(list(values = outValues, present = outPresent))
 }
 
 #' calculate difference and significance

### 2013-12-10 R/hypotheticalDataGeneration.R
added proper return value, and made sure to remove na values in the min function call

@@ -164,7 +164,7 @@ pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRU
   })
   
   setValues <- do.call(cbind, allValues[geneListNames])
-  setValues <- apply(setValues, 1, min)
+  setValues <- apply(setValues, 1, min, na.rm=TRUE)
   setValues[is.na(setValues)] <- naVal
   
   allValues$intersect[is.na(allValues$intersect)] <- naVal
@@ -176,6 +176,7 @@ pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRU
     measuredGO <- names(inEnrich@pvalues)
     hasGO <- useTerms %in% measuredGO
     names(hasGO) <- useTerms
+    hasGO
   })
   
   initPresent <- logical(length(useTerms))

### 2013-12-10 R/hypotheticalDataGeneration.R
added proper condition if not using logs

@@ -164,7 +164,13 @@ pvaluesMultiEnrich <- function(geneListNames, useTerms, hyperEnrichList, log=TRU
   })
   
   setValues <- do.call(cbind, allValues[geneListNames])
-  setValues <- apply(setValues, 1, min, na.rm=TRUE)
+  
+  if (log){
+    setValues <- apply(setValues, 1, min, na.rm=TRUE)
+  } else {
+    setValues <- apply(setValues, 1, max, na.rm=TRUE)
+  }
+  
   setValues[is.na(setValues)] <- naVal
   
   allValues$intersect[is.na(allValues$intersect)] <- naVal

### 2013-12-10 R/hypotheticalDataGeneration.R
added function for doing stats on multiple of the same GO term

@@ -249,6 +249,28 @@ trimGOFrac <- function(go2gene, deList){
   return(data.frame(size=goSize, frac=goFrac))
 }
 
+#' calculate statistics for all the same GO terms across multiple samples
+#' 
+#' For the case where we have calculated p-values and differences for random samples of genes, we need to calculate the mean and standard deviation for each GO term.
+#' 
+#' @param inList the list of results to work with
+#' @details expects each entry in \code{inList} to have a data.frame named \code{values}, and takes the column "diff" and does the statistics on it.
+#' @return data.frame
+#' @export
+sameGOStats <- function(inList){
+  getDiff <- function(inVar){
+    return(inVar$values$diff)
+  }
+  
+  allVals <- lapply(inList, getDiff)
+  allVals <- do.call(cbind, allVals)
+  meanVals <- apply(allVals, 1, mean, na.rm=T)
+  sdVals <- apply(allVals, 1, sd, na.rm=T)
+  minCI <- meanVals - 1.96 * (sdVals / sqrt(ncol(allVals)))
+  maxCI <- meanVals + 1.96 * (sdVals / sqrt(ncol(allVals)))
+  outVals <- data.frame(mean = meanVals, min = minCI, max = maxCI)
+}
+
 
 #' @name lung.RData
 #' @title lung.RData

### 2013-12-10 R/hypotheticalDataGeneration.R
added explicit return to the gostats calculation, and modified namespace

@@ -269,6 +269,7 @@ sameGOStats <- function(inList){
   minCI <- meanVals - 1.96 * (sdVals / sqrt(ncol(allVals)))
   maxCI <- meanVals + 1.96 * (sdVals / sqrt(ncol(allVals)))
   outVals <- data.frame(mean = meanVals, min = minCI, max = maxCI)
+  return(outVals)
 }
 
 

### 2013-12-11 R/hypotheticalDataGeneration.R
added function for generating noise genes

@@ -272,6 +272,19 @@ sameGOStats <- function(inList){
   return(outVals)
 }
 
+#' possible noise genes to sample from
+#' 
+#' @param go2gene list of GO term to gene mapping
+#' @param goSample character list of GO terms already sampled
+#' @export
+#' @return noise genes to sample from
+possibleNoise <- function(go2gene, goSample){
+  allGO <- names(go2gene)
+  notSample <- allGO[!(allGO %in% goSample)]
+  remGO <- go2gene[notSample]
+  posGenes <- unique(unlist(remGO, use.names=FALSE))
+  return(posGenes)
+}
 
 #' @name lung.RData
 #' @title lung.RData

### 2013-12-11 R/hypotheticalDataGeneration.R
changed the noise gene function

@@ -279,10 +279,9 @@ sameGOStats <- function(inList){
 #' @export
 #' @return noise genes to sample from
 possibleNoise <- function(go2gene, goSample){
-  allGO <- names(go2gene)
-  notSample <- allGO[!(allGO %in% goSample)]
-  remGO <- go2gene[notSample]
-  posGenes <- unique(unlist(remGO, use.names=FALSE))
+  allGenes <- unique(unlist(go2gene))
+  sampleGenes <- unique(unlist(go2gene[goSample]))
+  posGenes <- allGenes[!(allGenes %in% sampleGenes)]
   return(posGenes)
 }
 

### 2013-12-12 R/hypotheticalDataGeneration.R
go sample function being added

@@ -272,6 +272,53 @@ sameGOStats <- function(inList){
   return(outVals)
 }
 
+#' generate GO sample with genes
+#' 
+#' To test reproducibility of GO term sampling, we need to be able to sample GO terms and the genes annotated to them, and possibly noise genes as well.
+#' 
+#' @param go2gene \code{list} of go terms with associated genes
+#' @param nGO the number of go terms. see \emph{Details} for having different numbers of terms
+#' @param goLimits ranges of gene counts to restrict sampling. If multiple use a list
+#' @param nSample how many samples of genes do we want
+#' @param nGene the number of genes to sample that are annotated to the GO terms
+#' @param nNoise the number of noise genes
+#' @param sharedNoiseFrac how much of the noisy genes should be shared
+#' @export
+#' @return list
+#' @details \code{nGO} and \code{goLimits} should be the same length
+goSample <- function(go2gene, nGO, goLimits, nSample=2, nGene=1000, nNoise=0, sharedNoiseFrac=1){
+  stopifnot(names(nGO) == names(goLimits))
+  goCount <- data.frame(count=sapply(go2gene, length))
+  
+  goSamples <- unlist(lapply(names(nGO), function(inName){
+    limitedRandomSample(goCount, goLimits[[inName]], nGO[inName])
+  }))
+  
+  goClass <- unlist(lapply(names(nGO), function(inName){rep(inName, nGO[inName])}))
+  goSize <- goCount[goSamples, 'count']
+  
+  outSamples <- lapply(seq(1, nSample), function(iSample){
+    sampleTerms(goSamples, go2gene, nGene)[1:nGene]
+  })
+  
+  goSampleAnn <- sapply(goSamples, function(inGO){
+    min(sapply(outSamples, function(inSample){
+      length(intersect(go2gene[[inGO]], inSample))
+    }))
+  })
+  
+  goFrac <- goSampleAnn / goSize
+  
+  # and now the noise genes
+  if (nNoise != 0){
+    noiseGenes <- possibleNoise(go2gene, goSamples)
+    
+    if (sharedNoiseFrac)
+  }
+  
+}
+
+
 #' possible noise genes to sample from
 #' 
 #' @param go2gene list of GO term to gene mapping

### 2013-12-12 R/hypotheticalDataGeneration.R
running GO samples

@@ -309,7 +309,7 @@ goSample <- function(go2gene, nGO, goLimits, nSample=2, nGene=1000, nNoise=0, sh
   
   goFrac <- goSampleAnn / goSize
   
-  outGOData <- data.frame(id = goSamples, class = goClass, size = goSize, frac = goFrac)
+  outGOData <- data.frame(id = as.character(goSamples), class = goClass, size = goSize, frac = goFrac)
   
   noiseSample <- vector("list", nSample)
   

### 2013-12-18 R/hypotheticalDataGeneration.R
changed function to use the data frame directly , and modified the vignette appropriately

@@ -254,12 +254,12 @@ trimGOFrac <- function(go2gene, deList){
 #' For the case where we have calculated p-values and differences for random samples of genes, we need to calculate the mean and standard deviation for each GO term.
 #' 
 #' @param inList the list of results to work with
-#' @details expects each entry in \code{inList} to have a data.frame named \code{values}, and takes the column "diff" and does the statistics on it.
+#' @details expects each entry in \code{inList} to be a data.frame, and takes the column "diff" and does the statistics on it.
 #' @return data.frame
 #' @export
 sameGOStats <- function(inList){
   getDiff <- function(inVar){
-    return(inVar$values$diff)
+    return(inVar$diff)
   }
   
   allVals <- lapply(inList, getDiff)

### 2013-12-19 R/hypotheticalDataGeneration.R
new function for doing diff GO stats

@@ -272,6 +272,28 @@ sameGOStats <- function(inList){
   return(outVals)
 }
 
+
+#' calculate difference between noise and clean differences for 100 gene samples
+#' 
+#' @param inList the list of results to work with
+#' @return data.frame
+#' @export
+diffGOStats <- function(inList){
+  getDiff <- function(inVar){
+    diffDiff <- inVar$noise$diff - inVar$clean$diff
+    return(diffDiff)
+  }
+  
+  allvals <- lapply(inList, getDiff)
+  allVals <- do.call(cbind, allVals)
+  meanVals <- apply(allVals, 1, mean, na.rm=T)
+  sdVals <- apply(allVals, 1, sd, na.rm=T)
+  minCI <- meanVals - 1.96 * (sdVals / sqrt(ncol(allVals)))
+  maxCI <- meanVals + 1.96 * (sdVals / sqrt(ncol(allVals)))
+  outVals <- data.frame(mean = meanVals, min = minCI, max = maxCI)
+  return(outVals)
+}
+
 #' generate GO sample with genes
 #' 
 #' To test reproducibility of GO term sampling, we need to be able to sample GO terms and the genes annotated to them, and possibly noise genes as well.

### 2013-12-19 R/hypotheticalDataGeneration.R
found a typo in the diffGOStats function

@@ -284,7 +284,7 @@ diffGOStats <- function(inList){
     return(diffDiff)
   }
   
-  allvals <- lapply(inList, getDiff)
+  allVals <- lapply(inList, getDiff)
   allVals <- do.call(cbind, allVals)
   meanVals <- apply(allVals, 1, mean, na.rm=T)
   sdVals <- apply(allVals, 1, sd, na.rm=T)

### 2013-12-19 inst/doc/process_lung_data.R
updated vignettes

@@ -1,0 +1,18 @@
+
+## ----downloadData, eval=FALSE--------------------------------------------
+## download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE18nnn/GSE18842/matrix/GSE18842_series_matrix.txt.gz", "/mlab/data/rmflight/Documents/projects/work/categoryComparePaperRev/otherData/GSE18842_series_matrix.txt.gz")
+
+
+## ----loadResave, eval=FALSE----------------------------------------------
+## library(GEOquery)
+## lung <- getGEO(filename="otherData/GSE18842_series_matrix.txt.gz")
+## tmpP <- pData(lung)
+## tmpP <- tmpP[,c("title", "geo_accession", "platform_id", "data_processing", "supplementary_file", "source_name_ch1", "characteristics_ch1", "characteristics_ch1.1")]
+## pData(lung) <- tmpP
+## 
+## tmpF <- fData(lung)
+## tmpF <- tmpF[, c("ID", "GB_ACC", "SPOT_ID", "Annotation Date", "Sequence Type", "Sequence Source", "Target Description", "Representative Public ID", "Gene Title", "Gene Symbol", "ENTREZ_GENE_ID", "RefSeq Transcript ID")]
+## fData(lung) <- tmpF
+## save(lung, file="inst/data/lung.RData", compress="xz")
+
+

### 2013-12-20 R/hypotheticalDataGeneration.R
initial skeleton of the sweepNoiseSample function

@@ -374,6 +374,18 @@ possibleNoise <- function(go2gene, goSample){
   return(posGenes)
 }
 
+#' Different values of noise
+#' 
+#' @param noiseGenes all possible noise genes we can sample from
+#' @param sizeNoise a numeric vector of how many noise genes we want to add (\code{integer})
+#' @param fracShared a numeric vector of the fraction of shared noise genes to sweep over (\code{decimal})
+#' @return \code{list} of \code{lists}
+#' @details each entry in the list corresponds to a particular number of noise genes, and each list therein corresponds to the noise genes for each sample
+#' @export
+sweepNoiseSample <- function(noiseGenes, sizeNoise = seq(10, 1000, 10), fracShared = seq(0, 1, 0.01)){
+  
+}
+
 #' @name lung.RData
 #' @title lung.RData
 #' @docType data

### 2013-12-20 R/hypotheticalDataGeneration.R
able to generate noise samples with different characteristics over a range of values, with testing confirmed

@@ -377,13 +377,51 @@ possibleNoise <- function(go2gene, goSample){
 #' Different values of noise
 #' 
 #' @param noiseGenes all possible noise genes we can sample from
+#' @param nSamples how many samples do we need?
 #' @param sizeNoise a numeric vector of how many noise genes we want to add (\code{integer})
 #' @param fracShared a numeric vector of the fraction of shared noise genes to sweep over (\code{decimal})
 #' @return \code{list} of \code{lists}
 #' @details each entry in the list corresponds to a particular number of noise genes, and each list therein corresponds to the noise genes for each sample
 #' @export
-sweepNoiseSample <- function(noiseGenes, sizeNoise = seq(10, 1000, 10), fracShared = seq(0, 1, 0.01)){
+sweepNoiseSample <- function(noiseGenes, nSamples = 2, sizeNoise = seq(0, 1000, 10), fracShared = seq(0, 1, 0.01)){
+  outSamples <- vector('list', length(sizeNoise))
   
+  sizeLoc <- 1
+  # program this in a for loop first to get it right
+  for (inSize in sizeNoise){
+    #print(inSize)
+        
+    sizeSamples <- vector('list', length(fracShared))
+    
+    fracLoc <- 1
+    
+    for (inFrac in fracShared){
+      #print(inFrac)
+      
+      sampleGenes <- vector('list', nSamples)
+      
+      tmpGenes <- noiseGenes # a gene variable we can modify as needed
+      nShared <- round(inSize * inFrac)
+      nIndependent <- inSize - nShared
+      
+      shareGenes <- sample(tmpGenes, nShared)
+      
+      tmpGenes <- tmpGenes[!(tmpGenes %in% shareGenes)]
+      
+      for (iSample in seq(1, nSamples)){
+        #print(iSample)
+        uniqGenes <- sample(tmpGenes, nIndependent)
+        sampleGenes[[iSample]] <- c(shareGenes, uniqGenes)
+        tmpGenes <- tmpGenes[!(tmpGenes %in% uniqGenes)]
+      }
+      
+      sizeSamples[[fracLoc]] <- sampleGenes
+      fracLoc <- fracLoc + 1
+    }
+    outSamples[[sizeLoc]] <- sizeSamples
+    sizeLoc <- sizeLoc + 1
+  }
+  return(outSamples)
 }
 
 #' @name lung.RData

### 2013-12-20 inst/tests/test_noiseSweep.R

@@ -1,0 +1,29 @@
+context("Noise Sweep")
+
+test_that("noise sweep works properly", {
+  noiseGenes <- as.character(seq(1, 5000))
+  nSamples <- 2
+  sizeNoise <- seq(0, 500, 20)
+  fracShared <- seq(0, 1, .1)
+  
+  expLenNoise <- length(sizeNoise)
+  expLenShared <- length(fracShared)
+  expNumNoise <- sizeNoise
+  expNumShared <- round(rep(sizeNoise, each=expLenShared) * rep(fracShared, expLenNoise) )
+  
+  outNoise <- sweepNoiseSample(noiseGenes, nSamples, sizeNoise, fracShared)
+  
+  hasLenNoise <- length(outNoise)
+  expect_equal(hasLenNoise, expLenNoise)
+  
+  hasNumNoise <- sapply(outNoise, function(x){length(x[[1]][[1]])})
+  expect_equal(hasNumNoise, expNumNoise)
+  
+  hasNumShared <- lapply(outNoise, function(x){
+    lapply(x, function(y){
+      length(intersect(y[[1]], y[[2]]))
+    })
+  })
+  
+  expect_equal(unlist(hasNumShared), expNumShared)
+})

### 2013-12-20 R/hypotheticalDataGeneration.R
refactored sweepNoiseSample to use lapply on the two first loops to speed it up

@@ -384,20 +384,8 @@ possibleNoise <- function(go2gene, goSample){
 #' @details each entry in the list corresponds to a particular number of noise genes, and each list therein corresponds to the noise genes for each sample
 #' @export
 sweepNoiseSample <- function(noiseGenes, nSamples = 2, sizeNoise = seq(0, 1000, 10), fracShared = seq(0, 1, 0.01)){
-  outSamples <- vector('list', length(sizeNoise))
-  
-  sizeLoc <- 1
-  # program this in a for loop first to get it right
-  for (inSize in sizeNoise){
-    #print(inSize)
-        
-    sizeSamples <- vector('list', length(fracShared))
-    
-    fracLoc <- 1
-    
-    for (inFrac in fracShared){
-      #print(inFrac)
-      
+  outSamples <- lapply(sizeNoise, function(inSize){
+    sizeSamples <- lapply(fracShared, function(inFrac){
       sampleGenes <- vector('list', nSamples)
       
       tmpGenes <- noiseGenes # a gene variable we can modify as needed
@@ -414,14 +402,11 @@ sweepNoiseSample <- function(noiseGenes, nSamples = 2, sizeNoise = seq(0, 1000, 
         sampleGenes[[iSample]] <- c(shareGenes, uniqGenes)
         tmpGenes <- tmpGenes[!(tmpGenes %in% uniqGenes)]
       }
-      
-      sizeSamples[[fracLoc]] <- sampleGenes
-      fracLoc <- fracLoc + 1
-    }
-    outSamples[[sizeLoc]] <- sizeSamples
-    sizeLoc <- sizeLoc + 1
-  }
+      return(sampleGenes)
+    })
+  })
   return(outSamples)
+  
 }
 
 #' @name lung.RData

### 2014-01-06 vignettes/findingData.R
added file with R code for finding some possible data

@@ -1,0 +1,30 @@
+library(GEOmetadb)
+con <- dbConnect(SQLite(), 'GEOmetadb.sqlite')
+
+obQuery <- paste("SELECT DISTINCT gse.title,gse.gse",
+                 "FROM",
+                 "gse",
+                 "WHERE",
+                 "gse.type LIKE '%Expression profiling by array%' AND",
+                 "gse.title LIKE '%obes%'", sep=" ")
+
+obRes <- dbGetQuery(con, obQuery)
+
+
+crQuery <- paste("SELECT DISTINCT gse.title,gse.gse",
+                 "FROM",
+                 "gse",
+                 "WHERE",
+                 "gse.type LIKE '%Expression profiling by array%' AND",
+                 "gse.title LIKE '%crohn%'", sep=" ")
+
+crRes <- dbGetQuery(con, crQuery)
+
+metabQuery <- paste("SELECT DISTINCT gse.title,gse.gse",
+                    "FROM",
+                    "gse",
+                    "WHERE",
+                    "gse.type LIKE '%Expression profiling by array%' AND",
+                    "gse.title LIKE '%metabol%'", sep=" ")
+
+metabRes <- dbGetQuery(con, metabQuery)

### 2014-01-07 R/hypotheticalDataGeneration.R
added fishersMethod to our data

@@ -409,6 +409,16 @@ sweepNoiseSample <- function(noiseGenes, nSamples = 2, sizeNoise = seq(0, 1000, 
   
 }
 
+#' fishers method for combining p-values
+#' 
+#' @param x set of p-values to combine
+#' @export
+#' @return combined p-values
+#' @details function from \url{http://mikelove.wordpress.com/2012/03/12/combining-p-values-fishers-method-sum-of-p-values-binomial/}
+fishersMethod <- function(x){
+  pchisq(-2 * sum(log(x)),df=2*length(x),lower=FALSE)
+} 
+
 #' @name lung.RData
 #' @title lung.RData
 #' @docType data

### 2014-01-08 R/hypotheticalDataGeneration.R
starting to generate code for GSEA samples

@@ -419,6 +419,12 @@ fishersMethod <- function(x){
   pchisq(-2 * sum(log(x)),df=2*length(x),lower=FALSE)
 } 
 
+#' change set of p-values for GSEA
+#' 
+#' Given a list of sample genes, the full list of named p-values, and a mean and sd
+#' 
+#' @param sampleList 
+
 #' @name lung.RData
 #' @title lung.RData
 #' @docType data

### 2014-01-08 R/hypotheticalDataGeneration.R
added function for generating sample p-values for GSEA

@@ -423,7 +423,21 @@ fishersMethod <- function(x){
 #' 
 #' Given a list of sample genes, the full list of named p-values, and a mean and sd
 #' 
-#' @param sampleList 
+#' @param sampleList the sample genes we will use
+#' @param initPValues initial p-values for all the genes (named vector)
+#' @param distPValues a distribution of p-values to take samples from for the sample genes
+#' @return list where for each sample the appropriate entries have modified p-values
+#' @export
+samplePValueGenGSEA <- function(sampleList, initPValues, distPValues){
+  outValues <- lapply(sampleList, function(inList){
+    toMod <- which(names(initPValues) %in% inList)
+    nMod <- length(inList)
+    usePValues <- initPValues
+    usePValues[toMod] <- sample(distPValues, nMod)
+    return(usePValues)
+  })
+  return(outValues)
+}
 
 #' @name lung.RData
 #' @title lung.RData

### 2014-01-08 R/hypotheticalDataGeneration.R
generating gene set results

@@ -439,6 +439,49 @@ samplePValueGenGSEA <- function(sampleList, initPValues, distPValues){
   return(outValues)
 }
 
+#' run multiple geneSetTest's
+#' 
+#' @param statistics named vector of statistics
+#' @param genesets list of the gene sets to test
+#' @param alternative the alternative hypothesis to test
+#' @param ranks.only use the ranks of the entries only 
+#' @seealso geneSetTest
+#' @importFrom limma geneSetTest
+#' @export
+#' @return set of p-values for the genesets
+multiGeneSetTest <- function(statistics, genesets, alternative="down", ranks.only=FALSE){
+  genesetStatistics <- sapply(genesets, function(inSet){
+    index <- which(names(statistics) %in% inSet)
+    geneSetTest(index, statistics, alternative=alternative, ranks.only=ranks.only)
+  })
+  return(genesetStatistics)
+}
+
+#' run geneSetTest for multiple samples
+#' 
+#' For a list of samples and list of gene sets, do testing of the gene sets for each sample, as well
+#' as a meta-sample where there p-values for the samples are combined using Fisher's method
+#' 
+#' @param samplePValues list of p-values for each sample
+#' @param genesets list of gene sets to test
+#' @param alternative the alternative hypothesis to test
+#' @param ranks.only use the ranks only
+#' @seealso geneSetTest
+#' @return list with the gene set p-values for each sample, as well as the combined
+#' @export
+multiSampleGeneSetTest <- function(samplePValues, genesets, alternative="mixed", ranks.only=FALSE, transform2Log=TRUE){
+  pvalMatrix <- do.call(cbind, samplePValues)
+  combPValues <- apply(pvalMatrix, 1, fishersMethod)
+  samplePValues$comb <- combPValues
+  if (transform2Log){
+    samplePValues <- lapply(samplePValues, function(inVal){
+      -1 * log(inVal)
+    })
+  }
+  sampleSetValues <- lapply(samplePValues, multiGeneSetTest, genesets, alternative, ranks.only)
+  return(sampleSetValues)
+}
+
 #' @name lung.RData
 #' @title lung.RData
 #' @docType data

### 2014-01-13 R/hypotheticalDataGeneration.R
added new function for doing the genewise statistics using limma

@@ -482,6 +482,44 @@ multiSampleGeneSetTest <- function(samplePValues, genesets, alternative="mixed",
   return(sampleSetValues)
 }
 
+
+#' rank genes using limma
+#'
+#' @param exprData log-expression values from an ExpressionSet
+#' @param pData 
+#' @param sampleStatus character vector describing the samples
+#' @param doComps character vector of which comparisons to make
+#' @param dupStrategy how to resolve duplicates, default is to take the smallest p-value
+#' @param correction which multiple testing correction to apply to the results
+#' @return
+#' @export
+#' @importFrom limma makeContrasts lmFit contrasts.fit eBayes topTable
+rankGenes <- function(exprData, sampleStatus, doComps, dupStrategy="minP", aggregateIndex, adjust.method="BH"){
+  f <- factor(sampleStatus)
+  design <- model.matrix(~0 + f)
+  colnames(design) <- levels(f)
+  
+  contrast.matrix <- makeContrasts(contrasts=doComps, levels=design)
+  fit <- lmFit(exprData, design)
+  fit2 <- contrasts.fit(fit, contrast.matrix)
+  fit2 <- eBayes(fit2)
+  
+  outData <- lapply(doComps, function(inComp){
+    compData <- topTable(fit2, coef=inComp, number=Inf, adjust.method=adjust.method, sort.by="none")
+    naIndex <- is.na(aggregateIndex)
+    compData$aggregateBy <- aggregateIndex
+    compData <- compData[!naIndex,]
+    compData <- by(compData, compData$aggregateBy, function(inData){
+      if (dupStrategy == "minP"){
+        inData[which.min(inData$P.Value),]
+      }
+      
+    })
+    compData <- do.call(rbind, compData)
+  })
+  return(outData)
+}
+
 #' @name lung.RData
 #' @title lung.RData
 #' @docType data

### 2014-01-13 R/hypotheticalDataGeneration.R
calculating logFC and p-values using limma

@@ -491,7 +491,7 @@ multiSampleGeneSetTest <- function(samplePValues, genesets, alternative="mixed",
 #' @param doComps character vector of which comparisons to make
 #' @param dupStrategy how to resolve duplicates, default is to take the smallest p-value
 #' @param correction which multiple testing correction to apply to the results
-#' @return
+#' @return list of data frames
 #' @export
 #' @importFrom limma makeContrasts lmFit contrasts.fit eBayes topTable
 rankGenes <- function(exprData, sampleStatus, doComps, dupStrategy="minP", aggregateIndex, adjust.method="BH"){

### 2014-01-14 R/hypotheticalDataGeneration.R
new function to generate differentially expressed lists

@@ -520,6 +520,41 @@ rankGenes <- function(exprData, sampleStatus, doComps, dupStrategy="minP", aggre
   return(outData)
 }
 
+#' get differentially expressed genes in a data.frame
+#' 
+#' @param rankedData data.frame of p-values and fold changes
+#' @param id the column to use as a returned id
+#' @param useP the column with the p-values
+#' @param pcutoff maximum p-value to use
+#' @param lfc column identifier for log-fold-change
+#' @param lfcCutoff should a minumum fold-change be required
+#' @param splitDir should significant results be split by up and down?
+#' @export
+#' @return list
+#' @details if \code{splitDir=TRUE} (default), then the returned list will have \code{up}, \code{dn}, and \code{universe}. If \code{splitDir=FALSE}, then it will be \code{sig} and \code{universe}.
+getDiffGenes <- function(rankedData, id="aggregateBy", useP="adj.P.Val", pcutoff=0.05, lfc="logFC", lfcCutoff=NA, splitDir=TRUE){
+  sigEntry <- rankedData[, useP] <= pcutoff
+  
+  if (!(is.na(lfcCutoff))){
+    sigEntry <- sigEntry & abs(rankedData[, lfc] >= lfcCutoff)
+  }
+  
+  if (splitDir){
+    upEntry <- rankedData[, lfc] > 0
+    upID <- rankedData[(upEntry & sigEntry), id]
+    dnEntry <- rankedData[, lfc] < 0
+    dnID <- rankedData[(dnEntry & sigEntry), id]
+    
+    outData <- list(up=upID, dn=dnID, universe=rankedData[, id])
+  } else {
+    sigID <- rankedData[sigEntry, id]
+    outData <- list(sig=sigID, universe=rankedData[, id])
+  }
+  
+  return(outData)
+  
+}
+
 #' @name lung.RData
 #' @title lung.RData
 #' @docType data

### 2014-01-14 R/hypotheticalDataGeneration.R
added the original id (i.e. rownames) to the output data object so we can use it for other things.

@@ -508,6 +508,7 @@ rankGenes <- function(exprData, sampleStatus, doComps, dupStrategy="minP", aggre
     compData <- topTable(fit2, coef=inComp, number=Inf, adjust.method=adjust.method, sort.by="none")
     naIndex <- is.na(aggregateIndex)
     compData$aggregateBy <- aggregateIndex
+    compData$orgID <- rownames(compData)
     compData <- compData[!naIndex,]
     compData <- by(compData, compData$aggregateBy, function(inData){
       if (dupStrategy == "minP"){

### 2014-01-14 R/hypotheticalDataGeneration.R
added function for collapsing probe sets to single genes

@@ -483,6 +483,33 @@ multiSampleGeneSetTest <- function(samplePValues, genesets, alternative="mixed",
 }
 
 
+#' collapse probes to genes
+#' 
+#' take the median value of a set of probes to the associated genes in each sample. Make sure that \code{collapseBy} is in the same order as the \code{exprData} rows, and any rows you want removed should have \code{NA}.
+#' 
+#' @param exprData an expressionSet
+#' @param collapseBy character vector of associated gene ids
+#' @return exprData a matrix of median intensities
+#' @export
+collapseProbes <- function(exprData, collapseBy){
+  naLoc <- is.na(collapseBy)
+  exprData <- exprData[!naLoc,]
+  collapseBy <- collapseBy[!naLoc]
+  
+  doMedian <- function(inValues){
+    if (nrow(inValues) > 1){
+      outValues <- apply(inValues, 2, median)
+    } else {
+      outValues <- inValues
+    }
+    return(outValues)
+  }
+  
+  collapseData <- by(exprData, collapseBy, doMedian)
+  collapseData <- do.call(rbind, collapseData)
+  return(collapseData)
+}
+
 #' rank genes using limma
 #'
 #' @param exprData log-expression values from an ExpressionSet

### 2014-01-14 R/hypotheticalDataGeneration.R
add variable doAggregation and put into it's own logic location for rankGenes

@@ -521,7 +521,7 @@ collapseProbes <- function(exprData, collapseBy){
 #' @return list of data frames
 #' @export
 #' @importFrom limma makeContrasts lmFit contrasts.fit eBayes topTable
-rankGenes <- function(exprData, sampleStatus, doComps, dupStrategy="minP", aggregateIndex, adjust.method="BH"){
+rankGenes <- function(exprData, sampleStatus, doComps, dupStrategy="minP", doAggregation=FALSE, aggregateIndex, adjust.method="BH"){
   f <- factor(sampleStatus)
   design <- model.matrix(~0 + f)
   colnames(design) <- levels(f)
@@ -533,17 +533,20 @@ rankGenes <- function(exprData, sampleStatus, doComps, dupStrategy="minP", aggre
   
   outData <- lapply(doComps, function(inComp){
     compData <- topTable(fit2, coef=inComp, number=Inf, adjust.method=adjust.method, sort.by="none")
-    naIndex <- is.na(aggregateIndex)
-    compData$aggregateBy <- aggregateIndex
-    compData$orgID <- rownames(compData)
-    compData <- compData[!naIndex,]
-    compData <- by(compData, compData$aggregateBy, function(inData){
-      if (dupStrategy == "minP"){
-        inData[which.min(inData$P.Value),]
-      }
-      
-    })
-    compData <- do.call(rbind, compData)
+    if (doAggregation){
+      naIndex <- is.na(aggregateIndex)
+      compData$aggregateBy <- aggregateIndex
+      compData$orgID <- rownames(compData)
+      compData <- compData[!naIndex,]
+      compData <- by(compData, compData$aggregateBy, function(inData){
+        if (dupStrategy == "minP"){
+          inData[which.min(inData$P.Value),]
+        }
+        
+      })
+      compData <- do.call(rbind, compData)
+    }
+    
   })
   return(outData)
 }

### 2014-01-14 R/hypotheticalDataGeneration.R
forgot to add a return compData

@@ -546,7 +546,7 @@ rankGenes <- function(exprData, sampleStatus, doComps, dupStrategy="minP", doAgg
       })
       compData <- do.call(rbind, compData)
     }
-    
+    return(compData)
   })
   return(outData)
 }

### 2014-01-14 R/hypotheticalDataGeneration.R
added ID to the compdata data.frame

@@ -533,6 +533,7 @@ rankGenes <- function(exprData, sampleStatus, doComps, dupStrategy="minP", doAgg
   
   outData <- lapply(doComps, function(inComp){
     compData <- topTable(fit2, coef=inComp, number=Inf, adjust.method=adjust.method, sort.by="none")
+    compData$id <- rownames(compData)
     if (doAggregation){
       naIndex <- is.na(aggregateIndex)
       compData$aggregateBy <- aggregateIndex

### 2014-01-14 R/hypotheticalDataGeneration.R
added a new function to do Romer on multiple lists

@@ -587,6 +587,29 @@ getDiffGenes <- function(rankedData, id="aggregateBy", useP="adj.P.Val", pcutoff
   
 }
 
+
+#' multicontrast romer
+#' 
+#' given a contrast.matrix with multiple columns, do \code{romer} for those
+#' 
+#' @param index list of indices specifying how y maps to gene sets
+#' @param y numeric matrix of log-expression values
+#' @param design design matrix
+#' @param contrastMatrix the contrast.matrix, each column will be used
+#' @param nrot number of rotations
+#' @seealso romer
+#' @export
+#' @return list of results for each contrast performed
+#' @importFrom limma romer
+multicontrastRomer <- function(index, y, design, contrastMatrix, nrot=9999){
+  nContrast <- ncol(contrastMatrix)
+  
+  outData <- lapply(seq(1, nContrast), function(inCol){
+    useContrast <- contrastMatrix[, inCol]
+    romer(index, y, design, contrast=contrastMatrix[, inCol], nrot=nrot)
+  })
+}
+
 #' @name lung.RData
 #' @title lung.RData
 #' @docType data

### 2014-01-14 R/hypotheticalDataGeneration.R
setting up mmGeneSetTests

@@ -483,6 +483,10 @@ multiSampleGeneSetTest <- function(samplePValues, genesets, alternative="mixed",
 }
 
 
+#' multiple-multiple geneSetTests
+#' 
+#' The default \code{geneSetTest} in \code{limma} will only do a single index and a single test. This function allows one to pass in a set of indices, and do all three tests (up, down, mixed).
+
 #' collapse probes to genes
 #' 
 #' take the median value of a set of probes to the associated genes in each sample. Make sure that \code{collapseBy} is in the same order as the \code{exprData} rows, and any rows you want removed should have \code{NA}.

### 2014-01-15 R/hypotheticalDataGeneration.R
adding new function for doing multiple gene set tests (limma alternative to GSEA)

@@ -486,6 +486,33 @@ multiSampleGeneSetTest <- function(samplePValues, genesets, alternative="mixed",
 #' multiple-multiple geneSetTests
 #' 
 #' The default \code{geneSetTest} in \code{limma} will only do a single index and a single test. This function allows one to pass in a set of indices, and do all three tests (up, down, mixed).
+#' 
+#' @param index list of indices generated by \code{symbols2indices}
+#' @param statistics the statistics to use
+#' @param ranks.only whether to rank only, default is FALSE
+#' @param ... other parameters for \code{geneSetTest}
+#' 
+#' @details Note that the function will try to do all three of up, down and mixed, but if there is no sign (i.e. F-statistic instead of t-statistic) the function will only do a test for "mixed"
+#' @export
+#' @return data.frame of p-values for each alternative test
+#' @importFrom limma geneSetTest
+mmGeneSetTest <- function(index, statistics, ranks.only=FALSE, ...){
+  alternatives <- c("up", "down", "mixed")
+  
+  allsamesign <- all(statistics >= 0) || all(statistics <= 0)
+  if (allsamesign){
+    alternatives <- "mixed"
+  }
+  
+  byIndex <- lapply(index, function(inIndices){
+    byAlternative <- sapply(alternatives, function(inAlternative){
+      geneSetTest(inIndices, statistics, inAlternative, ranks.only=ranks.only)
+    })
+  })
+  
+  pvalues <- do.call(rbind, byIndex)
+  return(pvalues)
+}
 
 #' collapse probes to genes
 #' 

### 2014-01-16 R/hypotheticalDataGeneration.R
completely rewrote the function for generating the ranked list of t-values for doing GSEA

@@ -419,24 +419,39 @@ fishersMethod <- function(x){
   pchisq(-2 * sum(log(x)),df=2*length(x),lower=FALSE)
 } 
 
-#' change set of p-values for GSEA
+#' change set of t-values for GSEA
 #' 
-#' Given a list of sample genes, the full list of named p-values, and a mean and sd
+#' Given a list of sample genes, the gene universe, and a set of t-values
 #' 
 #' @param sampleList the sample genes we will use
-#' @param initPValues initial p-values for all the genes (named vector)
-#' @param distPValues a distribution of p-values to take samples from for the sample genes
-#' @return list where for each sample the appropriate entries have modified p-values
+#' @param universeGenes all the genes to use
+#' @param tValues a distribution of t-values
+#' @return list of new t-values that should put genes from sampleList at the top
 #' @export
-samplePValueGenGSEA <- function(sampleList, initPValues, distPValues){
-  outValues <- lapply(sampleList, function(inList){
-    toMod <- which(names(initPValues) %in% inList)
-    nMod <- length(inList)
-    usePValues <- initPValues
-    usePValues[toMod] <- sample(distPValues, nMod)
-    return(usePValues)
+sampletValueGenGSEA <- function(sampleList, universeGenes, tValues){
+  tValues <- sort(tValues)
+  nUniverse <- length(universeGenes)
+  outT <- numeric(nUniverse)
+  names(outT) <- universeGenes
+  newTSample <- lapply(sampleList, function(inList){
+    outTSample <- outT
+    nGene <- length(inList)
+    
+    inList <- sample(inList, nGene)
+    useT <- sample(tValues[1:nGene], nGene)
+    
+    outTSample[inList] <- useT
+    
+    otherT <- sample(tValues[seq(nGene+1, nUniverse)], nUniverse-nGene)
+    otherGenes <- universeGenes[!(universeGenes %in% inList)]
+    outTSample[otherGenes] <- otherT
+    return(outTSample)
   })
-  return(outValues)
+  
+  combSample <- do.call(cbind, newTSample)
+  combSample <- rowMeans(combSample)
+  newTSample$comb <- combSample
+  return(newTSample)
 }
 
 #' run multiple geneSetTest's

### 2014-01-16 R/hypotheticalDataGeneration.R
added function to get gsea diffs

@@ -571,6 +571,27 @@ getDiffGenes <- function(rankedData, id="aggregateBy", useP="adj.P.Val", pcutoff
   
 }
 
+
+#' results and diffs following genesettest
+#' 
+#' @param gseaRes a list of results from doing geneSetTests
+#' @param contrast which alternative hypothesis to extract (for fake data only one should be valid)
+#' @param orgLists the names of the original samples to extract, will be compared with "comb"
+#' @export
+#' @return data.frame of -1*log of p-values, and difference
+gseaDiffs <- function(gseaRes, contrast="down", orgLists=c("s1", "s2")){
+  orgRes <- do.call(cbind, lapply(gseaRes[orgLists], function(x){-1*log(x[,contrast])}))
+  
+  useNames <- rownames(orgRes)
+  
+  orgRes <- rowMin(orgRes)
+  
+  outRes <- data.frame(org=orgRes, comb=-1*log(gseaRes$comb[,contrast]))
+  outRes$diff <- outRes$org - outRes$comb
+  return(outRes)
+}
+
+
 #' @name lung.RData
 #' @title lung.RData
 #' @docType data

### 2014-01-21 R/hypotheticalDataGeneration.R
moved gseaSizeRange to the R/code file

@@ -676,6 +676,16 @@ multicontrastRomer <- function(index, y, design, contrastMatrix, nrot=9999){
   })
 }
 
+#' generate entries within a distribution
+#' 
+#' @param inLimits the range to use
+#' @param nSample how many entries within the range
+#' 
+#' @return set of sizes
+#' @export
+gseaSizeRange <- function(inLimits, nSample){
+  round(runif(nSample, inLimits[1], inLimits[2]))
+}
 
 #' @name lung.RData
 #' @title lung.RData

### 2014-01-21 R/hypotheticalDataGeneration.R
moved gseaSizeRanges and gseaProportion to actual built-in functions

@@ -687,6 +687,50 @@ gseaSizeRange <- function(inLimits, nSample){
   round(runif(nSample, inLimits[1], inLimits[2]))
 }
 
+#' generates proportion of genes in top 50
+#' 
+#' @param inSize vector of sizes to sample
+#' @param sigProp vector of significant proportion, i.e. fraction above 50 rank
+#' @param statistics the set of statistics to use (will be ranked by function)
+#' @param nSample number of samples to use
+#' 
+#' @details For each supplied inSize and sigProp, random indices of \code{inSize} are taken from the statistics. Based on \code{sigProp}, for each sample different numbers of indices take values from the upper 50 of ranked values and the lower 50 of ranked values. In this case each entry in \code{inSize} and \code{sigProp} denotes a different term, and they are treated independently. The \emph{list} combined method is based on averaging the statistics in this case.
+#' 
+#' @export
+#' @return list of lists for each \code{term}, with \emph{index} and \emph{stats}, \code{stats} contains a column for each sample, as well as a combined column. 
+gseaProportion <- function(inSize, sigProp, statistics, nSample=2){
+  statistics <- sort(statistics)
+  nPosGene <- length(statistics)
+  grpLoc <- seq(1, nPosGene, round(nPosGene / 4))
+  topStats <- statistics[seq(1, grpLoc[3])]
+  botStats <- statistics[seq(grpLoc[3], nPosGene)]
+  gseaStatistics <- lapply(seq(1, length(inSize)), function(inIndex){
+    nGene <- inSize[inIndex]
+    outIndex <- sample(nPosGene, nGene)
+    useProp <- sigProp[inIndex]
+    tmpStat <- statistics
+    
+    nTop <- round(nGene*useProp)
+    nBot <- nGene - nTop
+    
+    outStatistics <- lapply(seq(1, nSample), function(inSample){
+      topIndex <- sample(outIndex, nTop)
+      botIndex <- outIndex[!(outIndex %in% topIndex)]
+      tmpT <- tmpStat
+      tmpT[topIndex] <- sample(topStats, nTop)
+      tmpT[botIndex] <- sample(botStats, nBot)
+      return(tmpT)
+    })
+    
+    outStatistics <- do.call(cbind,outStatistics)
+    outStatistics <- cbind(outStatistics, rowMeans(outStatistics))
+    colnames(outStatistics) <- c(paste("s", seq(1, nSample), sep=""), "comb")
+    
[…]

### 2014-01-21 R/hypotheticalDataGeneration.R
move gseaProportion2 to gseaProportionVary and in the function file

@@ -731,6 +731,53 @@ gseaProportion <- function(inSize, sigProp, statistics, nSample=2){
   return(gseaStatistics)
 }
 
+
+
+#' vary gsea proportion
+#' 
+#' @param inSize vector of sizes
+#' @param sigProp list of proportions, one for each sample
+#' @param statistics the statistics to use
+#' 
+#' @details For each supplied inSize and sigProp, random indices of \code{inSize} are taken from the statistics. Based on \code{sigProp}, for each sample different numbers of indices take values from the upper 50 of ranked values and the lower 50 of ranked values. In this case each entry in \code{inSize} and denotes a different term, and they are treated independently. Each \emph{list} entry of \code{sigProp} is a sample, and each vector should be the same length as \code{inSize}. The \emph{list} combined method is based on averaging the statistics in this case.
+#' 
+#' @export
+#' @return list of lists for each \code{term}, with \emph{index} and \emph{stats}, \code{stats} contains a column for each sample, as well as a combined column. 
+gseaProportionVary <- function(inSize, sigProp, statistics){
+  statistics <- sort(statistics)
+  nPosGene <- length(statistics)
+  nSample <- length(sigProp)
+  grpLoc <- seq(1, nPosGene, round(nPosGene / 4))
+  topStats <- statistics[seq(1, grpLoc[4])]
+  botStats <- statistics[seq(grpLoc[4], nPosGene)]
+  gseaStatistics <- lapply(seq(1, length(inSize)), function(inIndex){
+    
+    nGene <- inSize[inIndex]
+    outIndex <- sample(nPosGene, nGene)
+    tmpStat <- statistics
+    
+    outStatistics <- lapply(seq(1, nSample), function(inSample){
+      # get significance proportion for each sample individually
+      nTop <- round(nGene * sigProp[[inSample]][inIndex])
+      nBot <- nGene - nTop
+      topIndex <- sample(outIndex, nTop)
+      botIndex <- outIndex[!(outIndex %in% topIndex)]
+      tmpT <- tmpStat
+      tmpT[topIndex] <- sample(topStats, nTop)
+      tmpT[botIndex] <- sample(botStats, nBot)
+      return(tmpT)
+    })
+    
+    outStatistics <- do.call(cbind,outStatistics)
+    outStatistics <- cbind(outStatistics, rowMeans(outStatistics))
[…]

### 2014-01-21 R/hypotheticalDataGeneration.R
adding p2signed function to function code

@@ -777,6 +777,15 @@ gseaProportionVary <- function(inSize, sigProp, statistics){
   return(gseaStatistics)
 }
 
+#' apply regression model to pvalues
+#' 
+#' To transform p-values to a signed range, we will apply the results of a regression model to the p-values
+#' 
+#' @param inModel result from \code{lm}
+#' @param pvalues set of pvalues to transform
+p2signed <- function(inModel, pvalues){
+  outSigned <- inModel$coefficients[1] + inModel$coefficients[2]*pvalues
+}
 
 #' @name lung.RData
 #' @title lung.RData

### 2014-01-21 R/hypotheticalDataGeneration.R
missed export directive

@@ -783,6 +783,8 @@ gseaProportionVary <- function(inSize, sigProp, statistics){
 #' 
 #' @param inModel result from \code{lm}
 #' @param pvalues set of pvalues to transform
+#' @export 
+#' @return transformed values
 p2signed <- function(inModel, pvalues){
   outSigned <- inModel$coefficients[1] + inModel$coefficients[2]*pvalues
 }

### 2014-01-21 R/hypotheticalDataGeneration.R
move gseaProportionFisher to code file

@@ -789,6 +789,51 @@ p2signed <- function(inModel, pvalues){
   outSigned <- inModel$coefficients[1] + inModel$coefficients[2]*pvalues
 }
 
+#' gseaProportion, combined using Fisher's method
+#' 
+#' @param inSize vector of sizes to sample
+#' @param sigProp vector of significant proportion, i.e. fraction above 50 rank
+#' @param statistics the set of statistics to use (will be ranked by function)
+#' @param nSample number of samples to use
+#' 
+#' @details For each supplied inSize and sigProp, random indices of \code{inSize} are taken from the statistics. Based on \code{sigProp}, for each sample different numbers of indices take values from the upper 50 of ranked values and the lower 50 of ranked values. In this case each entry in \code{inSize} and \code{sigProp} denotes a different term, and they are treated independently. The \emph{list} combined method is based on averaging the statistics in this case.
+#' 
+#' @export
+#' @return list of lists for each \code{term}, with \emph{index} and \emph{stats}, \code{stats} contains a column for each sample, as well as a combined column. 
+gseaProportionFisher <- function(inSize, sigProp, statistics, nSample=2){
+  statistics <- sort(statistics)
+  nPosGene <- length(statistics)
+  grpLoc <- seq(1, nPosGene, round(nPosGene / 4))
+  topStats <- statistics[seq(1, grpLoc[3])]
+  botStats <- statistics[seq(grpLoc[3]+1, nPosGene)]
+  gseaStatistics <- lapply(seq(1, length(inSize)), function(inIndex){
+    
+    nGene <- inSize[inIndex]
+    outIndex <- sample(nPosGene, nGene)
+    tmpStat <- statistics
+    
+    outStatistics <- lapply(seq(1, nSample), function(inSample){
+      # get significance proportion for each sample individually
+      nTop <- round(nGene * sigProp[inIndex])
+      nBot <- nGene - nTop
+      topIndex <- sample(outIndex, nTop)
+      botIndex <- outIndex[!(outIndex %in% topIndex)]
+      tmpT <- tmpStat
+      tmpT[topIndex] <- sample(topStats, nTop)
+      tmpT[botIndex] <- sample(botStats, nBot)
+      return(tmpT)
+    })
+    
+    outStatistics <- do.call(cbind,outStatistics)
+    combStat <- apply(outStatistics, 1, fishersMethod)
[…]

### 2014-01-21 R/hypotheticalDataGeneration.R
moving gseaProportionMax to built-in functions

@@ -834,6 +834,51 @@ gseaProportionFisher <- function(inSize, sigProp, statistics, nSample=2){
   return(gseaStatistics)
 }
 
+
+#' gseaProportion combining using max p-value
+#' 
+#' @param inSize vector of sizes to sample
+#' @param sigProp vector of significant proportion, i.e. fraction above 50 rank
+#' @param statistics the set of statistics to use (will be ranked by function)
+#' @param nSample number of samples to use
+#' 
+#' @details For each supplied inSize and sigProp, random indices of \code{inSize} are taken from the statistics. Based on \code{sigProp}, for each sample different numbers of indices take values from the upper 50 of ranked values and the lower 50 of ranked values. In this case each entry in \code{inSize} and \code{sigProp} denotes a different term, and they are treated independently. The \emph{list} combined method is based on taking the maximum p-value among the samples.
+#' 
+#' @export
+#' @return list of lists for each \code{term}, with \emph{index} and \emph{stats}, \code{stats} contains a column for each sample, as well as a combined column. 
+gseaProportionMax <- function(inSize, sigProp, statistics, nSample=2){
+  statistics <- sort(statistics)
+  nPosGene <- length(statistics)
+  grpLoc <- seq(1, nPosGene, round(nPosGene / 4))
+  topStats <- statistics[seq(1, grpLoc[3])]
+  botStats <- statistics[seq(grpLoc[3]+1, nPosGene)]
+  gseaStatistics <- lapply(seq(1, length(inSize)), function(inIndex){
+    
+    nGene <- inSize[inIndex]
+    outIndex <- sample(nPosGene, nGene)
+    tmpStat <- statistics
+    
+    outStatistics <- lapply(seq(1, nSample), function(inSample){
+      # get significance proportion for each sample individually
+      nTop <- round(nGene * sigProp[inIndex])
+      nBot <- nGene - nTop
+      topIndex <- sample(outIndex, nTop)
+      botIndex <- outIndex[!(outIndex %in% topIndex)]
+      tmpT <- tmpStat
+      tmpT[topIndex] <- sample(topStats, nTop)
+      tmpT[botIndex] <- sample(botStats, nBot)
+      return(tmpT)
+    })
+    
+    outStatistics <- do.call(cbind,outStatistics)
+    outStatistics <- cbind(outStatistics, rowMax(outStatistics))
[…]

### 2014-01-23 R/hypotheticalDataGeneration.R
made mmGeneSetTest parallel

@@ -505,13 +505,26 @@ multiSampleGeneSetTest <- function(samplePValues, genesets, alternative="mixed",
 #' @param index list of indices generated by \code{symbols2indices}
 #' @param statistics the statistics to use
 #' @param ranks.only whether to rank only, default is FALSE
+#' @param do.parallel use multiple cores to do everything
 #' @param ... other parameters for \code{geneSetTest}
 #' 
 #' @details Note that the function will try to do all three of up, down and mixed, but if there is no sign (i.e. F-statistic instead of t-statistic) the function will only do a test for "mixed"
 #' @export
 #' @return data.frame of p-values for each alternative test
 #' @importFrom limma geneSetTest
-mmGeneSetTest <- function(index, statistics, ranks.only=FALSE, ...){
+#' @importFrom parallel mclapply
+mmGeneSetTest <- function(index, statistics, ranks.only=FALSE, do.parallel=TRUE, ...){
+  
+  currCores <- getOption("mc.cores")
+  if (is.null(currCores)){
+    currCores <- 1
+  }
+  
+  if (!do.parallel){
+    currCores <- 1
+  }
+  
+  
   alternatives <- c("up", "down", "mixed")
   
   allsamesign <- all(statistics >= 0) || all(statistics <= 0)
@@ -519,11 +532,11 @@ mmGeneSetTest <- function(index, statistics, ranks.only=FALSE, ...){
     alternatives <- "mixed"
   }
   
-  byIndex <- lapply(index, function(inIndices){
+  byIndex <- mclapply(index, function(inIndices){
     byAlternative <- sapply(alternatives, function(inAlternative){
       geneSetTest(inIndices, statistics, inAlternative, ranks.only=ranks.only)
     })
-  })
+  }, mc.preschedule=FALSE, mc.cores=currCores)
   
   pvalues <- do.call(rbind, byIndex)
   return(pvalues)
@@ -667,13 +680,23 @@ gseaDiffs <- function(gseaRes, contrast="down", orgLists=c("s1", "s2")){
 #' @export
 #' @return list of results for each contrast performed
 #' @importFrom limma romer
-multicontrastRomer <- function(index, y, design, contrastMatrix, nrot=9999){
+#' @importFrom parallel mclapply
+multicontrastRomer <- function(index, y, design, contrastMatrix, nrot=9999, do.parallel=TRUE){
[…]

### 2014-01-24 R/datasets.R
added datasets from the uc-crohns and muscle skin comparisons

@@ -1,0 +1,60 @@
+#' @name uc_crohns_rawData.RData
+#' @title uc_crohns_rawData.RData
+#' @docType data
+#' @source http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE36807
+#' @details Used getGEO to get data on January 13, 2014
+NULL
+
+#' @name cucRomer.RData
+#' @title cucRomer.RData
+#' @docType data
+#' @source generated by the UC vs Crohn's analysis vignette
+#' @details used \code{limma} \code{romer} to generate results
+NULL
+
+#' @name skin_rma_data.RData
+#' @title skin_rma_data.RData
+#' @docType data
+#' @source generated using raw data files from Jeffrey Petruska.
+#' @details RMA normalized Affymetrix data.
+NULL
+
+
+#' @name muscle_data.RData
+#' @title muscle_data.RData
+#' @docType data
+#' @source http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE4411
+#' @details RMA normalized intensities downloaded from GEO using \code{getGEO}, on January 13, 2014; data was trimmed down to only the denervated and innervated controls
+NULL
+
+
+#' @name skinRomer.RData
+#' @title skinRomer.RData
+#' @docType data
+#' @source running \code{romer} on skin t7 vs t0 and t14 vs t0 comparisons
+#' @details did two sets of contrasts, t7 vs t0 and t14 vs t0 using the \code{multicontrastRomer} function
+NULL
+
+
+
+#' @name skinGeneSetTest.RData
+#' @title skinGeneSetTest.RData
+#' @docType data
+#' @source running \code{geneSetTest} on skin t7 vs t0 and t14 vs t0 comparisons
+#' @details did two sets of contrasts, t7 vs t0 and t14 vs t0, and up, down, and mixed alternative tests using \code{mmGeneSetTest}
+NULL
+
+#' @name muscleRomer.RData
+#' @title muscleRomer.RData
+#' @docType data
+#' @source running \code{romer} on muscle denervated over innervated
+#' @details did one contrast, denervated vs innervated using the \code{multicontrastRomer} function
+NULL
+
+
+#' @name muscleGeneSetTest.RData
+#' @title muscleGeneSetTest.RData
+#' @docType data
+#' @source running \code{mmGeneSetTest} on muscle denervated over innervated
+#' @details did one contrast, denervated vs innervated using the \code{mmGeneSetTest} function
+NULL

### 2014-01-24 R/hypotheticalDataGeneration.R
using data(dataSet), and added documentation

@@ -885,3 +885,43 @@ gseaProportionMax <- function(inSize, sigProp, statistics, nSample=2){
 #' @source ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE18nnn/GSE18842/matrix/GSE18842_series_matrix.txt.gz
 #' @details Downloaded, created an \code{ExpressionSet} object using \code{getGEO} on Dec 6, 2013.
 NULL
+
+#' @name 100GeneSamples.RData
+#' @title 100GeneSamples.RData
+#' @docType data
+#' @source generated in the hypothetical example vignette
+#' @details for the supplied set of GO terms, take 100 different 2 samples of 1000 genes.
+NULL
+
+#' @name 100GeneResults.RData
+#' @title 100GeneResults.RData
+#' @docType data
+#' @source running hypergeometric test on the samples in \code{100GeneSamples}
+#' @details did enrichment on the samples in \code{100GeneSamples}
+NULL
+
+#' @name 100GOSamples.RData
+#' @title 100GOSamples.RData
+#' @docType data
+#' @source for 100 different GO samples of size 100, generate 2 samples with 1000 genes
+NULL
+
+#' @name 100GOResults.RData
+#' @title 100GOResults.RData
+#' @docType data
+#' @source running enrichment and diffs on \code{100GOSamples}
+NULL
+
+#' @name noiseSamples.RData
+#' @title noiseSamples.RData
+#' @docType data
+#' @source hypothetical vignette from \package{ccPaperRev}
+#' @details For two 1000 gene samples, different amounts of noise (0 - 1000) genes (genes not annotated to 100 GO terms defining the samples) were added, with different fractions of overlap between the noise gene lists (0 - 1). 
+NULL
+
+#' @name noiseSweepRes.RData
+#' @title noiseSweepRes.RData
+#' @docType data
+#' @source running hypergeometric calculations and differences on \code{noiseSamples}
+NULL
+

### 2014-01-29 inst/doc/hypotheticalExample_vignette.R
updating vignette outputs with custom css included

@@ -1,133 +1,0 @@
-
-## ----nGO-----------------------------------------------------------------
-nGO <- 20
-
-
-## ----goAnnotations-------------------------------------------------------
-library(org.Hs.eg.db)
-library(GO.db)
-library(ccPaperRev)
-hsGO <- as.list(org.Hs.egGO2ALLEGS)
-
-goOntology <- Ontology(names(hsGO))
-goBP <- names(goOntology)[goOntology == "BP"]
-hsGO <- hsGO[goBP]
-hsGO <- lapply(hsGO, unique)
-universeGenes <- unique(unlist(hsGO))
-
-
-## ----goCountDist---------------------------------------------------------
-library(ggplot2)
-hsGO_count <- sapply(hsGO, length)
-hsGO_count <- data.frame(count=hsGO_count)
-ggplot(hsGO_count, aes(x=count)) + geom_bar(binwidth=10) + xlim(0, 2000) + ylim(0, 500)
-
-
-## ----defineMin-----------------------------------------------------------
-minGO <- 10
-
-
-## ----defineGroupLimits---------------------------------------------------
-grpLow <- c(10, 100)
-grpMed <- c(250, 500)
-grpHi <- c(500, 1500)
-
-
-## ----sampleGroups--------------------------------------------------------
-set.seed(271113) # for reproducibility
-nGO <- 20
-
-GO_low <- limitedRandomSample(hsGO_count, grpLow, nGO)
-GO_med <- limitedRandomSample(hsGO_count, grpMed, nGO)
-GO_hi <- limitedRandomSample(hsGO_count, grpHi, nGO)
-
-
-## ----generateSamples-----------------------------------------------------
-useGO <- c(GO_low, GO_med, GO_hi)
-goList <- list(low=GO_low, med=GO_med, hi=GO_hi)
-nGene <- 1000
-sample1_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
-sample2_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
-
-
-## ----checkFractions------------------------------------------------------
-goFractions <- calcFraction(hsGO[useGO], list(sample1=sample1_org, sample2=sample2_org))
-ggplot(goFractions, aes(x=size, y=frac, color=genelist)) + geom_point()
-
-
-## ----runCalcs------------------------------------------------------------
[…]

### 2014-01-29 inst/doc/process_lung_data.R
removing old vignettes from inst/doc

@@ -1,18 +1,0 @@
-
-## ----downloadData, eval=FALSE--------------------------------------------
-## download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE18nnn/GSE18842/matrix/GSE18842_series_matrix.txt.gz", "/mlab/data/rmflight/Documents/projects/work/categoryComparePaperRev/otherData/GSE18842_series_matrix.txt.gz")
-
-
-## ----loadResave, eval=FALSE----------------------------------------------
-## library(GEOquery)
-## lung <- getGEO(filename="otherData/GSE18842_series_matrix.txt.gz")
-## tmpP <- pData(lung)
-## tmpP <- tmpP[,c("title", "geo_accession", "platform_id", "data_processing", "supplementary_file", "source_name_ch1", "characteristics_ch1", "characteristics_ch1.1")]
-## pData(lung) <- tmpP
-## 
-## tmpF <- fData(lung)
-## tmpF <- tmpF[, c("ID", "GB_ACC", "SPOT_ID", "Annotation Date", "Sequence Type", "Sequence Source", "Target Description", "Representative Public ID", "Gene Title", "Gene Symbol", "ENTREZ_GENE_ID", "RefSeq Transcript ID")]
-## fData(lung) <- tmpF
-## save(lung, file="inst/data/lung.RData", compress="xz")
-
-

### 2014-01-30 inst/doc/responses2reviews.Rmd
generated vignettes

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{reviews}
-->

```{r customCSS, include=FALSE}
cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
options(markdown.HTML.stylesheet = cssFile)
```

# Introduction Problems

**Rev 3**:
* The definition of FLE and FSE is confusing. It is a good rule to use acronyms with more than one letter difference. The terms "feature-list" and "feature-set" do not provide an intuitive idea of the difference. Set/threshold-based (e.g. Fisher's Exact) and rank-based/threshold-free (GSEA) would have been better depictions of the two categories of (annotation) gene-set enrichment tests (GSE, or GSET) or over-representation analysis (ORA). Another major distinction the author may consider is between competitive and self-contained, based on the type of null hypothesis to be rejected, although for microarray and gene expression analysis the competitive methods (or hybrid) are far more popular than self-contained (which is more suitable for rare genetic variants, or specific applications to expression data).
* FLA: another poor acronym choice, too similar to FLE; suggest feature intersection analysis..? (FIA)

**Response**: We agree that the definitions were not truly useful, and appreciate the comments on how to improve them. The original and new text is below.

**Original Text:** Methods commonly used to summarize and infer relationships from high-throughput experiments include feature-list enrichment (FLE) and feature-set enrichment (FSE). FLE seeks to find statistically significant over-represented feature annotations in the differentially expressed feature list compared to all of the measured features (i.e. they are enriched), whereas FSE attempts to find sets of features (defined by common annotations) that appear significantly at the extremes of a ranked feature list.

**New Text:** Methods commonly used to summarize and infer relationships from high-throughput experiments include the set-based threshold based (SBTB) and rank-based threshold-free (RBTF) methods. SBTB methods typically seek to find statistically significant over-represented feature annotations (sets) in the differentially expressed (threshold) feature list compared to all of the measured features (i.e. they are enriched), whereas RBTF attempts to find sets of features (defined by common annotations) that appear significantly at the extremes of a ranked feature list.

# Methods Problems

**Rev 1**: The use of a single approach (or two similar approaches if you include the supplemental data) is not enough.  There is time course data here, some kind of time course analysis would have been nice, using maSigPro or STEM or similar.  Something like WGCNA for module analysis could also have been employed.  Also lacking is any biological follow-up on the in-silico inferences.  As mentioned in the paper, there is probably a profound tissue effect in the data, which could have been addressed by removal of tissue specific genes.

**Response**: As mentioned by yourself, we could not decide whether we were writing a methods or biology paper. In principle, if this was a single analysis of the skin denervation time course, then *maybe* a time course method would be appropriate. The rewrite of the manuscript (in progress) focuses on the **categoryCompare** method, and therefore we did not include an actual time-course specific method analysis of the skin-denervation data. In addition, as the goal was the comparison between tissues, and there is no equivalent dataset for muscle (that we have been able to find), such an analysis would have likely been criticized for not matching the muscle dataset.

**Rev 2**: 
* 1.Page 6, line 192. Reference to Bioconductor (Gentleman et al, 2004)  is misleading here. It would be appropriate to give reference or URL to CategoryCompare here, otherwise it may seem that CategoryCompare was developed in 2004. 
* 2. Page 7, line 237, Muscle Denervation: It should be mentioned here that it is mouse (not rat) data, otherwise further discussions are not clear. 
* Fig 5 requires more detailed explanations in the figure caption. Currently it is hard to understand.

**Response**
* 1. Thank you for pointing out the confusion here. As was noted, the current wording implies a reference to the package itself, whereas the intention was to reference the Bioconductor software project. Therefore, we have moved the Bioconductor reference directly adjacent to the mention of Bioconductor itself, and included the URL to the categoryCompare package site on the Bioconductor project.
* 2. Mouse is mentioned in the description of the data set acquisition.
* 3. Additional information about the figure has been included, including how Table 2 relates to Figure 5, and what the legend signifies.

**Rev 3**:
* the overlap and Jaccard coefficient used in Enrichment Map and cites are also available in combined form (combined overlap coefficient), which was to work better by the authors, so it should be implemented as well
* the authors should provide for input from any gene-set enrichment test producing a gene-set level significance score; restricting to Fisher's Exact test (aka hypergeometric) is over-simplistic, and disregards all the method development that happened in the last decade; being able to process GSEA output would be a good start
* the way the color scheme is used is suboptimal for visualization (the colors do not help very much distinguishing the various condition combinations in an intuitive way); the authors should make a better effort on that side; I specifically suggest trying the following methodology: pick one color for each comparison with an ssociated list of annotation gene-sets enriched (e.g. skin t7 up: yellow, skin t14 up: orange, skin t7 dw: olive green, skin 14t dw: water green, muscle up: magenta, muscle dw: blue) and then represent the significance in more than one comparison using Cytoscape pie chart node coloring (so, a node representing an annotation gene-set significantly over-represented in skin t7 up and skin t14 up would be half yellow half orange).

**Responses**:

* regarding similarity/overlap coefficients, these implementations came from the original Erichment Map (EM) paper (Merico D, Isserlin R, Stueker O, Emili A, Bader GD (2010) Enrichment Map: A Network-Based Method for Gene-Set Enrichment Visualization and Interpretation. PLoS ONE 5(11): e13984. doi:10.1371/journal.pone.0013984). The combined coefficient is mentioned on the EM user guide page, but there is no mention of it actually being found to be better.

One publications Supporting Materials (http://www.nature.com/nature/journal/v466/n7304/extref/nature09146-s1.pdf) mentions that "We find that the arithmetic average of these two behaves better, and that is what we use here and refer to as the weighted overlap", however no evidence of the improvement is provided outside of this statement. Therefore, the combined coefficient has been included, but without more evidence from the authors of EM, I am not inclined to make it the default.

The combined coefficient has been implemented, see https://github.com/rmflight/categoryCompare/commit/7d00dcbcfb4c1fdae3401e45e5c4dccbe3637822 for the actual commit adding the function, and commits https://github.com/rmflight/categoryCompare/commit/7930b0925d31fa2d7691ce3d182a2ae7a89c174d https://github.com/rmflight/categoryCompare/commit/4bf13e21ed174dc73c4d360f92fb5aa75d40d72c for where the slot is added to the classes so it can be used.

* You are right, we should be able to handle just about any input. Some support was available using the "GENcc" methods, however they have been significantly updatedto allow the use of any results as long as they are supplied with significant annotations from each sample, and a mapping of annotations to genes. This was used to enable GSEA type analysis of both experimental data-sets discussed in the publication.

* This is a good point. Although for small numbers of experiments (see for example the Crohn's vs UC comparison, which has only 4 distinct lists, and a total of 7 meaningful comparisons) solid colors may not be too bad, beyond that it gets rather complicated. However, using pie-graphs where the pie is split only by the significant experiments may lead to perception problems, as solid (only significant in one experiment) or a few slices (significant in 2 or 3) may be percieved as more important. Based on this thinking, we have provided a method that splits nodes into pie-graphs with a slice for each condition / sample, and non-significance is denoted by complete desaturation of the color for that condition. In addition, a **generateLegend** function is included that plots a legend in an R plot window for reference. Note that we use png graphics that replace the node in Cytoscape, because we are using RCytoscape to communicate with Cytoscape, we cannot interact with other plugins in Cytoscape from R.

* Piegraph visualization addition in commit: https://github.com/rmflight/categoryCompare/commit/40168a0d55578e96fae9f0e6f3d1f4ef9b42c028

# Discussion Problems

**Rev 1**: 
* There is little tie-in of the results to the underlying biological question either independently or compared to prior art, it is not clear how much, if any, the field has been advanced by the findings.  The graph relationships between the factors as presented in figure 5 are not addressed at all.

* The sections on the general characteristics of skin/muscle denervation are entirely background and do not belong in the discussion. The paper can also not decide if its primary focus is the algorithm or the biology.  There is some mention in various parts of the paper including the discussion how CategoryCompare would be useful for the integration of heterogeneous datasets, yet this potentially useful application is not exemplified or demonstrated in any way.

* line 582 claims that there are "a number of new testable hypotheses" generated by the algorithm, but none are listed.
* line 598 claims that the biological data from the paper is a necessary benchmark for screening assessment of tissues from disease/trauma, but how that applies is not clear.
* the paragraph beginning line 521 even suggests the approach taken in the paper might not be productive.

**Rev 2**:
* 1. Authors discussed the advantages of the CategoryCompare approach, but have not mentioned possible concerns or limitations. It would help to list the assumptions of the approach.
*  2. While using CategoryCompare for mice and rats, authors preserved only the genes that have homologs in both species. How many genes (%) where discarded as a result of such selection. How much information is lost? Why was it necessary if you perform comparison at the level of pathways and biological processes? Have you tried the same comparison without discarding the genes that had no homologs?

**Responses:**
* 1. This was an oversight. We hope that an assessment of the method using hypothetical data demonstrates possible limitations.

* 2. The question of necessity is a good one, and we could not think of any particular reason to enforce that outside of other previous studies doing it, and therefore used all the genes in both cases.

**Rev 3**:
* besides the well thought point about angiogenesis, there is a bit too much speculation about biological results; unless results are more convincing, this just burdens the paper, which is more about the tool and its demonstratation of use; the only point that may deserve a bit more attention is the presence / absence of an inflammatory / immune response signature

# Results Problems

**Rev 1**:
* The figures in the paper are inappropriate.  Figures 1 and 2 are unnecessary, they are simplistic and only referenced in the introduction.  Figure 4 is relatively uninformative and also not needed.  Figure 5 is a wall of graphs which is very hard to interpret. In addition, there is no evidence of the edge weighting mentioned in the paper.

* I do not find the results particularly compelling or novel compared to other approaches.  It is clear there is alot of overlap between the FLA approach and other approaches with respect to the findings.  In addition two of the three novel annotations highlighted in table 2 are not really novel:  There is little difference between "reactive oxygen metabolism" and "response to H2O2"; and similarly little difference between "Blood vessel development" and "VEGF signalling", the latter being a primary driver of angiogenesis.

**Rev 3**:
* GSEA (pre-ranked using the limma moderated t statistic) should be tried out for the analysis, results may get better  I am disappointed there is no clue of axonogenesis or other developmental gradients other than angiogenesis and VEGF
* the authors should compare to the use of Enrichment Map (EM), with the following two-comparison maps: muscle up + skin 7t up, muscle up + skin 14t up, muscle down + skin 7t dw, muscle down + skin 14t dw. Yes , EM was not designed for (wildly) multi-condition studies, but in these cases comparison can still be broken down and analyzed -- the authors should strive to make a convincing point about the increased intuitiveness of their representation, and convince the reader. Note this would probably work better with GSEA NES for coloring nodes.

**Reponse**:

* Re: GSEA: we have used GSEA for both data-sets. Please see the revised manuscript and the "UC vs Crohn's" and "Skin vs Muscle" vignettes in the ccPaperRev package. https://github.com/rmflight/ccPaperRev

* There are many similarities between categoryCompare and other tools, and we admit that. However, we are not seeking to demonstrate that our visualization is better than EM's, but the goal is to provide a tool that is for the most part self-contained in R, from analysis to visualization. Although the data can be visualized in EM, that is merely a convenience, and we were wrong to try and place so much emphasis on the visualization in the original manuscript. In addition, all of the graph information is accessible in the original R instance.

# Submission Petruska of data to GEO

**Rev 1**: The biology in the paper is a microarray experiment, this should be submitted to GEO.

**Response**:

* This was an oversight on the first author's part on not checking the status of getting the data submitted. We have started the submission process, and are still waiting on protocol details to complete the submission.

# Test on multiple datasets

* Suggestion from Editor, and **Rev 1** that if we are going to be a tool paper, we should do multiple datasets.
  * Problem: what datasets to use that exist, that don't take forever to process?
  * Simulated data: can we come up with a simulated dataset?

# Drop mapping to orthologues

Argue that this removes the benefit of using annotations instead of genes.

* How would this change the results? Is it that important? Might be, given the low annotation of Rat genes overall

# Support GSEA

Use GSEA results too

* We should be able to write a method that takes GSEA results and does the same analysis, and demonstrate it.

# Claim of Novelty

**Rev 1**: It is not clear if this is a biology paper that uses a bioinformatic tool or a bioinformatic paper using biology as an example.  From a bioinformatics perspective, there is no clear evidence that the novel algorithm is any better than established methods for performing this kind of analysis, and not enough comparison with other methods was presented in the paper.  From a biological perspective, there are few novel findings presented in this paper and no clear indication of how they impact the field.

If it were to be refactored, the authors should decide whether it is the biology or the algorithm they want to present and prioritize the paper accordingly.  If the former then a more detailed analysis of the array should be undertaken, preferably with further wet bench validation.  If the latter than a wider range of examples of its application should be included, with emphasis on its ability to integrate heterogeneous data and the utility of visualizing the results in cytoscape

**Response**: Nowhere do we claim the algorithm is novel. We simply lay it out. Seriously, a lot of other people have done this, as we list in our references. This is simply packaging it up in a nice little application that we feel is useful.

# What we really have done

Took a simple method that had been previously done (see Shen & Tseng 2010, Merico et al 2010), generalized Shen & Tseng's work (i.e. don't really need to have same items for the MAPE-P part), and then put the equivalent of Merico 2010 into R, that get's visualized in Cytoscape.

Advantages: all the annotation available from Bioconductor, as well as the various enrichment methods available, with the visualization capabilities of R. This means we get the statistical manipulation abilities of R for our data.

# Discussion with Eric

**Trim down the amount of focus on Jeff's data.** Can we trim down the biology and background and just focus on it as an example data set.

Hypothetical situation, theoretical data set: Start with the GO-Terms, have 2000 diff exp. genes, from the 2000, choose some Gene Ontologies (perhaps 20), how many genes need to be rep. ?

Test data set: 60 GO Terms from BP, 20 with large number of genes, 20 with medium, 20 with low (>= 10 genes annotated), and choose large fraction from each.

Set 1500 diff expressed genes from the genes annotated with GO Terms chosen above, and apply the two methods, and see how they are different.

### 2014-01-30 inst/doc/datasets.Rmd
generated vignettes

# ReAnalysis based on reviewer comments

Because we have re-worked the publication as a paper about the tool, we have decided to do a re-analysis of the original data presented in the original manuscript based on the reviewer comments, as well as include other data sets.

# Muscle VS Skin Denervation

The original data consisted of muscle denervation data from GSE4411, and skin denervation data from Dr. Jeffrey Petruska.

# UC VS Crohn's

From GEO there is a ulcerative colitis vs crohn's disease data set available, GSE36807. This would be a nice one to try, as there is already a publication on it.

# Metabolomics VS Transcriptomics

The NCI 60 database has both metabolomics and transcriptomics data available. Unfortunately, there is no **control** data set.

Metabolomics File:

  * File Format: SAMPLENAME, TITLE, PANELNBR, CELLNBR, pname, cellname, VALUE

Ideas for analysis: slow vs fast growing, sensitive to compound vs insensitive to compound

Transcriptomics File:

  * File Format: MOLTID (NCI pattern #), ACC, Gene, TITLE, MOLTNBR (NCI exp. id #), PANELNBR, CELLNBR, pname, cellname, CHIP, FEATURE_ID, UniGene, GeneID, VALUE Gene assignments are based on Unigene Build #U225 (August 2010)

### 2014-01-30 inst/doc/hypotheticalExample_vignette.Rmd
generated vignettes

@@
   %\VignetteIndexEntry{hypothetical example}
-  %\VignetteDepends{categoryComparePaperRev, ggplot2}
+  %\VignetteDepends{categoryComparePaperRev, ggplot2, ALL, GO.db, hgu95av.db, hgu133plus2.db, limma}
 -->
+
+```{r customCSS, include=FALSE}
+cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+options(markdown.HTML.stylesheet = cssFile)
+```
 
@@
 
-```{r nGO}
-nGO <- 20
+```{r setupPlot}
+library(ggplot2)
+library(ggthemes)
+useTheme <- theme_bw() + theme(legend.key=element_blank(), legend.title = element_text(size=25), axis.title = element_text(size=25), axis.text = element_text(size=15), legend.text = element_text(size=20), strip.text.y = element_text(size=20), strip.text.x = element_text(size=20))
+
+savePlot <- function(plotObj, plotFile, figPath="/mlab/data/rmflight/Documents/projects/work/categoryComparePaperRev/savedPlots"){
+  if (file.exists(figPath)){
+    plotObj <- plotObj + ggtitle(NULL)
+    png(filename=file.path(figPath, plotFile), width=854, height=628)
+    print(plotObj)
+    dev.off()
+  }
+}
 ```
 
-We will make use of three sets of `r nGO` GO terms from *H. sapiens*. For each set, we want differing numbers of genes annotated to them.
+We will make use of three sets of GO terms from *H. sapiens*. For each set, we want differing numbers of genes annotated to them.
 
-### Num Annotations Dist
+### Number of Genes Annotated Distribution
 
@@
 
-```{r goAnnotations}
+```{r goAnnotations, message=FALSE}
 library(org.Hs.eg.db)
@@
 ```{r goCountDist}
-library(ggplot2)
 hsGO_count <- sapply(hsGO, length)
-hsGO_count <- data.frame(count=hsGO_count)
-ggplot(hsGO_count, aes(x=count)) + geom_bar(binwidth=10) + xlim(0, 2000) + ylim(0, 500)
+hsGO_count <- data.frame(size=hsGO_count)
+p <- ggplot(hsGO_count, aes(x=size)) + geom_bar(binwidth=10) + xlim(0, 2000) + ylim(0, 500) + useTheme + ggtitle("Number of genes annotated to GO terms") + xlab('Num. of Genes') + ylab('Count')
+p
+
+savePlot(p, "go2gene_distribution.png")
 ```
@@
 
-```{r defineMin}
-minGO <- 10
-```
+Based on the distribution in the previous figure, we will set up 3 different groups of GO terms, with a different number from each group:
 
-Let us define a minimum number of genes that need to be annotated to a GO term (`r minGO`), and then divide the GO terms into groups that we can subsequently take random samples from.
+  * low: 10-100, 50 terms
+  * med: 250-500, 30 terms
+  * hi: 500-1500, 20 terms
+
+This gives us a total of 100 GO terms, with a good sample from each group. We will generate random samples that are defined by these limits, and use this set of GO terms going forward.
 
@@
 set.seed(271113) # for reproducibility
-nGO <- 20
+nGO <- c(50, 30, 20)
+names(nGO) <- c("low", "med", "hi")
 
-GO_low <- limitedRandomSample(hsGO_count, grpLow, nGO)
-GO_med <- limitedRandomSample(hsGO_count, grpMed, nGO)
-GO_hi <- limitedRandomSample(hsGO_count, grpHi, nGO)
+GO_low <- limitedRandomSample(hsGO_count, grpLow, nGO["low"])
+GO_med <- limitedRandomSample(hsGO_count, grpMed, nGO["med"])
+GO_hi <- limitedRandomSample(hsGO_count, grpHi, nGO["hi"])
+
+useGO <- c(GO_low, GO_med, GO_hi)
+useGOSizeClass <- c(rep("low", nGO["low"]), rep("med", nGO["med"]), rep("hi", nGO["hi"]))
+names(useGOSizeClass) <- useGO
+```
+
+### Examine Gene-GO Distribution for Real Data
+
+Before we go and generate our samples, we should actually process some real experimental data and examine the fraction of annotated genes in the experiment compared to the total number of genes annotated.
+We will use the [`ALL`](http://master.bioconductor.org/packages/release/data/experiment/html/ALL.html) data set from `Bioconductor` and a [paired lung cancer dataset](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE18842) available from GEO (available as part of this package).
+
+#### ALL
+
+The `ALL` data set has two classes of tumor, **B** and **T**-cell lymphoblastoma. We will do a simple differential expression between the two cell types.
+
+```{r deALL}
+library(ALL)
+library(limma)
+data(ALL)
+
+grpALL <- factor(strtrim(pData(ALL)$BT, 1))
+designALL <- model.matrix(~0+grpALL)
+colnames(designALL) <- c("B", "T")
+fitALL <- lmFit(ALL, designALL)
+contMatrixALL <- makeContrasts(BvT=B-T, levels=designALL)
+fitALL2 <- contrasts.fit(fitALL, contMatrixALL)
+fitALL2 <- eBayes(fitALL2)
+deALL <- rownames(topTable(fitALL2, adjust="BH", p.value=0.05, number=Inf))
+```
+
+Now for all the GO terms that map to the probes in the `ALL` differentially expressed genes, how does the proportion of differentially expressed genes change as a function of the number of genes annotated to the GO term?
+
+```{r goDistALL}
+library(hgu95av2.db)
+ALLgo <- as.list(hgu95av2GO2ALLPROBES)
+ALL_sizeFrac <- trimGOFrac(ALLgo, deALL)
+ggplot(ALL_sizeFrac, aes(x=size, y=frac)) + geom_point() + xlim(0, 1500) + useTheme + ggtitle("DE ALL fraction as a function of size") + xlab("Number of Genes Annotated to Term") + ylab("Fraction of Term Genes in DE Sample")
+```
+
+#### Lung Cancer
+
+The lung cancer data is a paired sample design from GEO.
+
+```{r deLung}
+data(lung)
+sampleStat <- strsplit2(pData(lung)$title, " ")
+
+notNP <- grep("^NP", sampleStat[,2], invert=T)
+lung <- lung[, notNP]
+sampleStat <- sampleStat[notNP,]
+
+sampleID <- factor(sampleStat[,2])
+cancerID <- factor(sampleStat[,1])
+
+lungDesign <- model.matrix(~sampleID+cancerID)
+fitLung <- lmFit(lung, lungDesign)
+fitLung <- eBayes(fitLung)
+deLung <- rownames(topTable(fitLung, coef="cancerIDTumor", number=2000, adjust.method="BY", p.value=0.00001))
+```
+
+```{r goDistLung}
+library(hgu133plus2.db)
+lung2go <- as.list(hgu133plus2GO2ALLPROBES)
+lung_sizeFrac <- trimGOFrac(lung2go, deLung)
+lung_sizeFrac$genelist <- "lung"
+ggplot(lung_sizeFrac, aes(x=size, y=frac)) + geom_point() + xlim(0, 1500) + useTheme + ggtitle("DE Lung fraction as a function of size") + xlab("Num. of Genes") + ylab("Fraction")
 ```
@@
 ```{r generateSamples}
-useGO <- c(GO_low, GO_med, GO_hi)
 goList <- list(low=GO_low, med=GO_med, hi=GO_hi)
@@
 sample2_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
+
+#check our expectations
+goFractions <- calcFraction(hsGO[useGO], list(sample1=sample1_org, sample2=sample2_org))
+ggplot(goFractions, aes(x=size, y=frac, color=genelist)) + geom_point() + useTheme + ggtitle("Sample GO fractions") + xlab("Num. of Genes") + ylab("Fraction")
 ```
 
-### Checking our expectations about the fractions of GO terms.
+### Compare to experimental
 
-```{r checkFractions}
-goFractions <- calcFraction(hsGO[useGO], list(sample1=sample1_org, sample2=sample2_org))
-ggplot(goFractions, aes(x=size, y=frac, color=genelist)) + geom_point()
+Lets actually compare this to the distribution of fractions *vs* sizes from the experimental data sets.
+Because those used **all** the GO terms, we will redo them with **all** GO terms as well, not just those from our sample.
+
+```{r compareFracs}
+sampleGOFracs <- calcFraction(hsGO, list(sample1=sample1_org, sample2=sample2_org))
+ALL_sizeFrac$genelist <- "all"
+ALL_tmp <- rbind(ALL_sizeFrac, lung_sizeFrac, sampleGOFracs[,c("frac", "genelist", "size")])
+ggplot(ALL_tmp, aes(x=size, y=frac)) + geom_point() + xlim(0, 1500) + facet_grid(genelist ~ .) + useTheme + ggtitle("Sample GO fractions for all") + xlab("Num. of Genes") + ylab("Fraction")
+```
+
+```{r cleanupData}
+keepItems <- c("goFractions", "sample1_org", "sample2_org", "hsGO", "useGO", "useGOSizeClass", "GO_hi", "GO_low", "GO_med", "grpHi", "grpLow", "grpMed", "minGO", "nGO", "nGene", "universeGenes", "savePlot", "useTheme", "fitALL2")
+allItems <- ls()
+delItems <- allItems[!(allItems %in% keepItems)]
+rm(list=delItems)
 ```
@@
 ```{r noNoisePvalues}
-noNoise_pvalues <- pvaluesMultiEnrich(c("sample1", "sample2"), useGO, go_noNoise)
+noNoise_results <- pvaluesMultiEnrich(c("sample1", "sample2"), useGO, go_noNoise)
 
-noNoise_pvalues <- pvalueDiffSig(noNoise_pvalues, pCutoff=0.05, log=TRUE)
+noNoise_pvalues <- pvalueDiffSig(noNoise_results$values, pCutoff=0.05, log=TRUE)
 ```
@@
 ```{r noNoiseAddInfo}
-sizeClass <- rep(c("low", "med", "hi"), each=20)
-names(sizeClass) <- useGO
-noNoise_pvalues$sizeClass <- sizeClass
-noNoise_pvalues$size <- goFractions[1:60,4]
-noNoise_pvalues$frac <- goFractions$frac[1:60]
+noNoise_pvalues$sizeClass <- useGOSizeClass
+noNoise_pvalues$size <- goFractions[1:length(useGOSizeClass),4]
+noNoise_pvalues$frac <- goFractions$frac[1:length(useGOSizeClass)]
 ```
@@
 noNoise_pvalues$sizeClass <- factor(noNoise_pvalues$sizeClass, levels=c("low", "med", "hi"), ordered=TRUE)
-ggplot(noNoise_pvalues, aes(x=frac, y=diff, color=sigState)) + geom_point() + facet_grid(. ~ sizeClass, scales="free_x")
-ggplot(noNoise_pvalues, aes(x=sizeClass, y=diff)) + geom_boxplot() + geom_point()
+p <- ggplot(noNoise_pvalues, aes(x=frac, y=diff)) + geom_point() + facet_grid(.
+~ sizeClass, scales="free_x") + xlab("fraction") + useTheme + ggtitle("Sample - Combined Differences") + xlab("Fraction") + ylab("Diff.
+log p-values")
+p
+savePlot(p, "nonoise_sample_combineddiff.png")
+ggplot(noNoise_pvalues, aes(x=sizeClass, y=diff)) + geom_boxplot() + geom_point() + ggtitle("Sample - Combined Differences") + xlab("Size Class") + ylab("Diff. log p-values")
 ```
@@
 nNoise <- 500
-gene2HsGO <- reverseSplit(hsGO)
-not_useGO <- sapply(gene2HsGO, function(x){
-  sum(x %in% useGO) == 0
-})
-noiseGenes <- names(not_useGO)[not_useGO]
+noiseGenes <- possibleNoise(hsGO, useGO)
 noiseGenes <- sample(noiseGenes, nNoise)
@@
 ```{r noisePvalues}
-noise_pvalues <- pvaluesMultiEnrich(c("sample1", "sample2"), useGO, go_noise)
+noise_results <- pvaluesMultiEnrich(c("sample1", "sample2"), useGO, go_noise)
 
-noise_pvalues <- pvalueDiffSig(noise_pvalues, pCutoff=0.05, log=TRUE)
+noise_pvalues <- pvalueDiffSig(noise_results$values, pCutoff=0.05, log=TRUE)
 
-noise_pvalues$sizeClass <- sizeClass
-noise_pvalues$size <- goFracNoise[1:60,4]
-noise_pvalues$frac <- goFracNoise$frac[1:60]
+noise_pvalues$sizeClass <- useGOSizeClass
+noise_pvalues$size <- goFracNoise[1:length(useGOSizeClass),4]
+noise_pvalues$frac <- goFracNoise$frac[1:length(useGOSizeClass)]
 ```
@@
 noise_pvalues$sizeClass <- factor(noise_pvalues$sizeClass, c("low", "med", "hi"), ordered=TRUE)
-ggplot(noise_pvalues, aes(x=frac, y=diff, color=sigState)) + geom_point() + facet_grid(. ~ sizeClass, scales="free_x")
+ggplot(noise_pvalues, aes(x=frac, y=diff, color=sigState)) + geom_point() + facet_grid(.
+~ sizeClass, scales="free_x") + useTheme + ggtitle("Sample - Combined with Noise") + xlab("Fraction") + ylab("Diff.
+log p-values")
 ```
@@
 
-## Expectations
+### Difference of Differences
 
-Based on the above results, what I expect we will see if we do a programmatic search over the space is that there is a dependence on both the fraction of DE genes that are noisy, and the fraction of noisy genes that are shared (which really amounts to the same thing, but is a subtle point).
+Is there any **difference** in the *difference of p-values* between **noise** vs **noNoise** (comparing **set** vs **list**)?
 
-# To Do:
+```{r diffDiffs}
+diff_pvalues <- noise_pvalues
+diff_pvalues$diff <- noise_pvalues$diff - noNoise_pvalues$diff
+ggplot(diff_pvalues, aes(x=frac, y=diff)) + geom_point() + facet_grid(.
+~ sizeClass, scales="free_x") + ggtitle("difference of differences noise vs noNoise") + useTheme + xlab("Fraction") + ylab("Diff.
+noise / noNoise")
+tapply(diff_pvalues$diff, diff_pvalues$sizeClass, mean)
+```
 
-* Check expectations of the gene to GO distribution from a couple of real human data sets
-* Run both the **noNoise** and **noise** 100 times with different sets of GO terms and make sure that this holds.
-* Vary the fraction of noise genes, and also vary the fraction that is common, and find a way to quantify the effect.
+From this plot, it appears that not only do the **set** values return lower p-values, but return lower p-values when noise is added overall.
+
+## Check Consistency
+
+To make sure that the results are consistent, we will do several iterations of the above calculations.
+
+* For this set of GO terms, do the above calculations 100 times
+* Generate 100 different sets of GO terms, and run calculations again.
+
+### Gene Sample Level Consistency
+
+```{r geneSamples, eval=FALSE}
+nTimes <- 100
+nGene <- 1000
+
+testSamples <- lapply(seq(1, nTimes), function(x){
+  list(s1 = sampleTerms(useGO, hsGO, 1000, 4)[1:nGene],
+       s2 = sampleTerms(useGO, hsGO, 1000, 4)[1:nGene],
+       noise = sample(noiseGenes, nNoise))
+})
+
+save(useGO, testSamples, universeGenes, useGOSizeClass, file="inst/data/100GeneSamples.RData")
+```
+
+```{r runSamples, eval=FALSE}
+data("100GeneSamples")
+library(snowfall)
+# load("inst/data/100GeneSamples.RData")
+testFun <- function(x){
+  nnList <- x[c("s1", "s2")]
+  nnRes <- hyperGOMultiEnrichment(nnList, universe=universeGenes)
+  nn_res <- pvaluesMultiEnrich(c("s1", "s2"), useGO, nnRes)
+  nnP <- pvalueDiffSig(nn_res$values, pCutoff=0.05, log=TRUE)
+  nnP$class <- useGOSizeClass
+
+  noiList <- list(s1 = c(x$s1, x$noise), s2 = c(x$s2, x$noise))
+  noiRes <- hyperGOMultiEnrichment(noiList, universe=universeGenes)
+  noi_res <- pvaluesMultiEnrich(c("s1", "s2"), useGO, noiRes)
+  noiP <- pvalueDiffSig(noi_res$values, pCutoff=0.05, log=TRUE)
+  noiP$class <- useGOSizeClass
+
+  return(list(clean = nnP, noise = noiP))
+}
+sfInit(parallel=TRUE, cpus=10)
+sfExport("useGOSizeClass", "universeGenes", "useGO")
+sfLibrary(ccPaperRev)
+sfLibrary(GO.db)
+sfLibrary(org.Hs.eg.db)
+testRes <- sfLapply(testSamples, testFun)
+sfStop()
+save(testRes, file="inst/data/100GeneResults.RData")
+```
+
+### Errors from 100 Samples of the initial set of GO terms
+
+For each of the GO terms, we want to calculate the mean, standard deviation, and then a 95% confidence interval to represent graphically.
+
+```{r noNoiseStats}
+data("100GeneResults")
+cleanRes <- lapply(testRes, function(x){x$clean})
+testStats_clean <- sameGOStats(cleanRes)
+testStats_clean$class <- useGOSizeClass
+testStats_clean$frac <- goFractions$frac[1:length(useGOSizeClass)]
+testStats_clean$class <- factor(testStats_clean$class, levels=c("low", "med", "hi"), ordered=TRUE)
+p <- ggplot(testStats_clean, aes(x=frac, y=mean, ymax=max, ymin=min)) + geom_point() + geom_errorbar() + facet_grid(.
+~ class, scales="free_x") + useTheme + ggtitle("Sample - List Differences, 100 Gene Samples") + xlab("Fraction") + ylab("Mean diff.
+log p-values")
+p
+savePlot(p, "genesamples100_sampleListDiffs.png")
+```
+
+```{r noiseStats}
+noiseRes <- lapply(testRes, function(x){x$noise})
+testStats_noise <- sameGOStats(noiseRes)
+testStats_noise$class <- useGOSizeClass
+testStats_noise$frac <- goFracNoise$frac[1:length(useGOSizeClass)]
+testStats_noise$class <- factor(testStats_noise$class, levels=c("low", "med", "hi"), ordered=TRUE)
+ggplot(testStats_noise, aes(x=frac, y=mean, ymax=max, ymin=min)) + geom_point() + geom_errorbar() + facet_grid(.
+~ class, scales="free_x") + ggtitle("Sample - List Differences, 100 Gene Samples, with noise") + xlab("Fraction") + ylab("Mean diff.
+log p-values") + useTheme
+```
+
+These show that if there is a difference in the p-values from **set** and **list**, they are real.
+
+What about differences between noise and clean?
+
+```{r 100GeneNoiseClean}
+noiseCleanDiff <- diffGOStats(testRes)
+noiseCleanDiff$class <- useGOSizeClass
+noiseCleanDiff$frac <- goFracNoise$frac[1:length(useGOSizeClass)]
+noiseCleanDiff$class <- factor(noiseCleanDiff$class, levels=c("low", "med", "hi"), ordered=TRUE)
+ggplot(noiseCleanDiff, aes(x=frac, y=mean, ymax=max, ymin=min)) + geom_point() + geom_errorbar() + facet_grid(.
+~ class, scales="free_x") + ggtitle("Difference of Differences, 100 Gene Samples") + useTheme + xlab("Fraction") + ylab("Diff noise / noNoise")
+```
+
+Here we see that the noisy case also tends to generate higher differences in p-values between **set** and **list** than the clean case.
+
+**Caveats**: Currently the *fraction* of genes annotated to GO terms is assumed to be the same for each sample.
+This may not actually be the case, and ideally this should be repeated with independent fraction estimates for each sampling.
+
+## GO Samples
+
+OK, the error bars are not too bad. How about sampling multiple sets of GO terms??
+
+```{r genSamples, eval=FALSE}
+goLimits <- list(low = c(10, 100), med = c(250, 500), hi = c(500, 1500))
+
+nSample <- 100
+
+sfInit(parallel=TRUE, cpus=10)
+sfLibrary(ccPaperRev)
+sfExport("hsGO", "nGO", "goLimits")
+
+testGOSample <- sfLapply(seq(1, nSample), function(x){
+  t1 <- goSample(hsGO, nGO, goLimits, nSample=2, nGene=1000, nNoise=500)
+  return(t1)
+})
+
+sfStop()
+save(testGOSample, file="inst/data/100GOSamples.RData")
+```
+
+Running those samples:
+
+```{r runGOSamples, eval=FALSE}
+data("100GOSamples")
+runGOSamples <- function(x){
+  cleanSample <- x$geneSample
+  cleanRes <- hyperGOMultiEnrichment(cleanSample, universe=universeGenes)
+
+  cleanP <- pvaluesMultiEnrich(names(cleanSample), x$goData$id, cleanRes)
+  cleanP_dif <- pvalueDiffSig(cleanP$values, pCutoff=0.05, log=TRUE)
+  cleanP_dif <- cbind(cleanP_dif, x$goData)
+
+  noiseSample <- lapply(names(cleanSample), function(inSample){
+    c(x$geneSample[[inSample]], x$noiseSample[[inSample]])
+  })
+  names(noiseSample) <- names(cleanSample)
+  noiseRes <- hyperGOMultiEnrichment(noiseSample, universe=universeGenes)
+  noiseP <- pvaluesMultiEnrich(names(cleanSample), x$goData$id, noiseRes)
+  noiseP_dif <- pvalueDiffSig(noiseP$values, pCutoff=0.05, log=TRUE)
+  noiseP_dif <- cbind(noiseP_dif, x$goData)
+  return(list(clean = cleanP_dif, noise = noiseP_dif))
+}
+
+library(snowfall)
+sfInit(parallel=TRUE, cpus=10)
+sfLibrary(ccPaperRev)
+sfLibrary(org.Hs.eg.db)
+sfLibrary(GO.db)
+sfExport("universeGenes")
+
+testGORes <- sfLapply(testGOSample, runGOSamples)
+sfStop()
+save(testGORes, file="inst/data/100GOResults.RData")
+```
+
+```{r lookDiffs}
+data("100GOResults")
+
+grab_noNoise <- function(x){
+  x$clean[,c("diff", "class", "frac")]
+}
+
+go_noNoise <- lapply(testGORes, grab_noNoise)
+go_noNoise <- do.call(rbind, go_noNoise)
+go_noNoise$class <- factor(go_noNoise$class, levels=c("low", "med", "hi"), ordered=TRUE) 
+go_noNoiseT <- go_noNoise[(!is.nan(go_noNoise$diff)),]
+p <- ggplot(go_noNoiseT, aes(x = frac, y = diff)) + geom_point() + facet_grid(.
+~ class, scales="free_x") + useTheme + stat_density2d(aes(fill = ..level..), geom="polygon") + ggtitle("Sample - List Differences, 100 GO Samples, noNoise") + xlab("Fraction") + ylab("Diff.
+log p-values") + labs(fill="density")
+p
+savePlot(p, "gosamples100_samplelistDiff.png")
+
+grab_noise <- function(x){
+  x$noise[, c("diff", "class", "frac")]
+}
+go_noise <- lapply(testGORes, grab_noise)
+go_noise <- do.call(rbind, go_noise)
+go_noise$class <- factor(go_noise$class, levels=c("low", "med", "hi"), ordered=TRUE)
+ggplot(go_noise, aes(x = frac, y = diff)) + geom_point() + facet_grid(.
+~ class, scales="free_x") + ggtitle("Sample - List Differences, 100 GO Samples, Noise") + xlab("Fraction") + ylab("Diff.
+log p-values") + useTheme
+
+grab_noiseDiff <- function(x){
+  outRes <- x$clean[, c("diff", "class", "frac")]
+  outRes$diff <- x$noise$diff - x$clean$diff
+  outRes
+}
+go_diff <- lapply(testGORes, grab_noiseDiff)
+go_diff <- do.call(rbind, go_diff)
+go_diff$class <- factor(go_diff$class, levels=c("low", "med", "hi"), ordered=TRUE)
+ggplot(go_diff, aes(x = frac, y = diff)) + geom_point() + facet_grid(.
+~ class, scales="free_x") + ggtitle("Difference of Differences, 100 GO Samples") + xlab("Fraction") + ylab("Diff.
+log p-values") + useTheme
+ggplot(go_diff, aes(x = frac, y = diff, fill=class)) + geom_boxplot(position="identity", alpha=0.4) + ggtitle("Difference of Differences, 100 GO Samples") + xlab("Fraction") + ylab("Diff.
+noise / noNoise") + useTheme
+```
+
+How about we compare the distribution of noise and no-noise using violin plots?
+
+```{r compNoiseNoNoise}
+go_noise$noise <- "noise"
+go_noNoise$noise <- "none"
+
+go_all <- rbind(go_noise, go_noNoise)
+rownames(go_all) <- NULL
+ggplot(go_all, aes(x = frac, y = diff, fill = noise)) + geom_violin(trim=TRUE, alpha=0.5, position="identity") + facet_grid(.
+~ class, scales="free_x") + ggtitle("Difference of Differences, 100 GO Samples") + xlab("Fraction") + ylab("Diff.
+log p-values") + useTheme
+```
+
+What if we aggregate things into bins first, based on their class and fraction?
+
+```{r aggregate}
+go_diff$frac10 <- round(go_diff$frac * 10) / 10
+ggplot(go_diff, aes(x = factor(frac10), y = diff)) + geom_boxplot() + facet_grid(.
+~ class, scales="free_x") + ggtitle("Difference of Differences, 100 GO Samples") + xlab("Fraction") + ylab("Diff.
+log p-values") + useTheme
+
+probPoint <- is.infinite(go_diff$diff) | is.na(go_diff$diff) | is.nan(go_diff$diff)
+go_diff <- go_diff[!probPoint, ]
+ggplot(go_diff, aes(x=frac, y=diff)) + stat_smooth() + facet_grid(.
+~ class, scales="free_x") + ggtitle("Difference of Differences, 100 GO Samples") + xlab("Fraction") + ylab("Diff.
+log p-values") + useTheme
+```
+
+Although the improvement of noise over clean data dips below 0 at medium fractions, it does tend to improve rather significantly as the fraction of genes annotated to a GO term increases.
+
+## As a function of noisy and shared noise genes
+
+For a given set of genes and GO terms, lets now keep adding more and more noise genes, and for the set amount of noise genes, go from 0% to 100% shared noise genes.
+And then see how things change.
+For this, we are going to use our initial set of GO terms (`useGO`), and a single original set of genes, simply sampling noise genes as we go along.
+Note that in this case **noise** genes are defined as genes that are not annotated to any of the GO terms that the DE genes are annotated to.
+
+As in the previous examples, we will generate the samples first, and then run the calculations on them. This is because it is easier to test sample generation and enrichment separately.
+
+```{r sweepNoise, eval=FALSE}
+noiseGenes <- possibleNoise(hsGO, useGO)
+sizeNoise <- seq(0, 1000, 10)
+fracShared <- seq(0, 1, 0.01)
+noiseSamples <- sweepNoiseSample(noiseGenes, nSamples=2, sizeNoise, fracShared)
+
+nGene <- 1000
+cleanSample <- list(s1 = sampleTerms(useGO, hsGO, nGene, 4)[1:nGene],
+                    s2 = sampleTerms(useGO, hsGO, nGene, 4)[1:nGene])
+
+save(noiseSamples, cleanSample, sizeNoise, fracShared, universeGenes, useGO, useGOSizeClass, file=file.path(useDir, 'inst/data/noiseSamples.RData'))
+```
+
+```{r runSweptNoise, eval=FALSE}
+data(noiseSamples)
+
+useNoise <- unlist(noiseSamples, recursive=FALSE)
+sizeNoiseAll <- rep(sizeNoise, each=length(fracShared))
+fracSharedAll <- rep(fracShared, length(sizeNoise))
+
+runSweepNoise <- function(noiseGenes){
+  inNoise <- lapply(seq(1, length(cleanSample)), function(x){
+    c(cleanSample[[x]], noiseGenes[[x]])
+  })
+  names(inNoise) <- names(cleanSample)
+
+  noiseRes <- hyperGOMultiEnrichment(inNoise, universe=universeGenes)
+  noiseP <- pvaluesMultiEnrich(names(inNoise), useGO, noiseRes)
+  noisePComp <- pvalueDiffSig(noiseP$values, pCutoff=0.05, log=TRUE)
+  noisePComp$class <- useGOSizeClass
+  return(noisePComp)
+}
+
+library(snowfall)
+sfInit(parallel=TRUE, cpus=10)
+sfLibrary(ccPaperRev)
+sfLibrary(org.Hs.eg.db)
+sfLibrary(GO.db)
+sfExport("universeGenes", "useGO", "useGOSizeClass", "cleanSample")
+noiseSweepRes <- sfLapply(useNoise, runSweepNoise)
+sfStop()
+save(noiseSweepRes, file.path(useDir, "inst/data/noiseSweepRes.RData"))
+```
+
+Examine the noise results.
+
+```{r loadSweptNoise}
+data(noiseSamples)
+data(noiseSweepRes)
+
+noiseFrac <- data.frame(size=rep(sizeNoise, each=length(fracShared)),
+                        shared=rep(fracShared, length(sizeNoise)))
+```
+
+```{r examineSweptNoise}
+outMedians <- lapply(noiseSweepRes, function(inRes){
+  tapply(inRes$diff, inRes$class, median)
+})
+outMedians <- do.call(rbind, outMedians)
+outMedians <- cbind(outMedians, noiseFrac)
+
+ggplot(outMedians, aes(x=size, y=hi, color=shared)) + geom_point() + useTheme + ggtitle("Hi Median Difference") + xlab("Noise Added") + ylab("Median Diff.")
+
+ggplot(outMedians, aes(x=size, y=med, color=shared)) + geom_point() + useTheme + ggtitle("Medium Median Difference") + xlab("Noise Added") + ylab("Median Diff.")
+
+ggplot(outMedians, aes(x=size, y=low, color=shared)) + geom_point() + useTheme + ggtitle("Low Median Difference") + xlab("Noise Added") + ylab("Median Diff.")
+```
+
+```{r examine2}
+outMedians2 <- data.frame(diff=c(outMedians$low, outMedians$med, outMedians$hi), class=c(rep("low", nrow(outMedians)), rep("med", nrow(outMedians)), rep("hi", nrow(outMedians))), size=rep(outMedians$size, 3), shared=rep(outMedians$shared, 3))
+outMedians2$class <- factor(outMedians2$class, c("low", "med", "hi"), ordered=TRUE)
+
+p <- ggplot(outMedians2, aes(x=size, y=diff, color=shared)) + geom_point() + useTheme + facet_grid(class ~ ., scales="free_y") + ggtitle("Sample - Combined as a function of size and shared fraction")
+p
+savePlot(p, "samplecombined_functionSizeFraction.png")
+```
+
+# GSEA
+
+Another way to do the analysis is using a Gene Set Enrichment Analysis, or GSEA.
+In contrast to over-representation analysis, where we count if a group is more highly represented in a group *vs* the background; GSEA tests if a **set** of genes are more highly ranked than random sets of genes.
+
+For a hypothetical example, we are going to use the `geneSetTest` from `limma`, because it allows us to test on a set of gene statistics themselves.
+
+## Data set up
+
+The sets will be our same GO terms, and we will change the proportion of genes from each GO term that are enhanced above a base value.
+All genes will be assigned a value from a uniform distribution with a value of 0.5.
+For each GO term set, some genes will be selected (with an exponential decaying probability for more genes), and assigned p-values from a uniform distribution with a value of 0.01.
+
+For the **list** version, the p-values of genes will be combined using **Fishers** method, see the function `fishersMethod` provided in this package.
+
+## Generate Data
+
+We will use the t-values that `limma` has generated for the ALL data set, and just sample from those.
+
+```{r generateSampleRanks}
+tALL <- topTable(fitALL2, number=Inf)
+tALL <- tALL$t[(tALL$t < 10) & (tALL$t > -10)]
+tValues <- runif(length(universeGenes), -6, 6)
+
+gseaSamples <- list(s1 = sampleTerms(useGO, hsGO, 1000, 4),
+                    s2 = sampleTerms(useGO, hsGO, 1000, 4))
+
+gseaGOFracS1 <- calcFraction(hsGO[useGO], list(s1=gseaSamples[[1]]))
+
+gseaTValues <- sampletValueGenGSEA(gseaSamples, universeGenes, tValues)
+```
+
+Check that we did what we thought. If done right, all of the **sample** genes should be on one side of the a histogram.
+
+```{r checkValues}
+tmpDat <- data.frame(x=gseaTValues[[1]], type="all", stringsAsFactors=FALSE)
+tmpDat$type[(universeGenes %in% gseaSamples[[1]])] <- "sample"
+ggplot(tmpDat, aes(x=x, fill=type)) + geom_histogram(binwidth=0.1, position="identity") + ggtitle("location of samples") + useTheme
+
+tmpComb <- data.frame(x=gseaTValues[["comb"]], type="all", stringsAsFactors=FALSE)
+tmpComb$type[(universeGenes %in% gseaSamples[[1]])] <- "sample"
+ggplot(tmpComb, aes(x=x, fill=type)) + geom_histogram(binwidth=0.1, position="identity") + ggtitle("location of samples") + useTheme
+```
+
+Now lets actually do some calculations.
+
+```{runSomeGSEA}
+go2index <- symbols2indices(hsGO[useGO], universeGenes)
+gseaSetResults <- lapply(gseaTValues, function(x){
+  mmGeneSetTest(go2index, x)
+})
+```
+
+And figure out what is doing better.
+
+```{r gseaBetterGO}
+cleanDiff <- gseaDiffs(gseaSetResults)
+cleanDiff$sizeClass <- useGOSizeClass
+cleanDiff$sizeClass <- factor(cleanDiff$sizeClass, c("low", "med", "hi"), ordered=TRUE)
+cleanDiff$frac <- gseaGOFracS1$frac
+
+ggplot(cleanDiff, aes(x=frac, y=diff)) + geom_point() + facet_grid(. ~ sizeClass) + ggtitle("Sample - List for GSEA, using GO") + useTheme + xlab("Fraction") + ylab("Diff. log p-values")
+```
+
+So this appears to be a different trend than for the hypergeometric case, in that a **meta** sample (achieved by averaging the t-statistics) actually does no worse, and in many cases much better than the worst p-value.
+This effect is less-pronounced as the number of genes in a class become larger, and as the fraction covered increases.
+
+So what happens if we add some noisy genes into the top 1000? Stuff that isn't annotated at all to the GO terms of interest?
+
+```{r gseaNoise}
+nNoise <- 2000
+noiseGenes <- possibleNoise(hsGO, useGO)
+noiseGenes <- sample(noiseGenes, nNoise)
+
+gseaNoise <- list(s1=c(gseaSamples[["s1"]], noiseGenes),
+                  s2=c(gseaSamples[["s2"]], noiseGenes))
+
+gseaTValues_noise <- sampletValueGenGSEA(gseaNoise, universeGenes, tValues)
+
+tmpDat <- data.frame(x=gseaTValues_noise[[1]], type="all", stringsAsFactors=FALSE)
+tmpDat$type[(universeGenes %in% gseaSamples[[1]])] <- "sample"
+ggplot(tmpDat, aes(x=x, fill=type)) + geom_histogram(binwidth=0.1, position="identity", alpha=0.75) + ggtitle("Sample - List GSEA GO, adding noise genes") + useTheme
+```
+
+```{r runGSEANoise}
+gseaSetResults_noise <- lapply(gseaTValues_noise, function(x){
+  mmGeneSetTest(go2index, x)
+})
+```
+
+```{r gseaBetter}
+noiseDiff <- gseaDiffs(gseaSetResults_noise)
+noiseDiff$sizeClass <- useGOSizeClass
+noiseDiff$sizeClass <- factor(noiseDiff$sizeClass, c("low", "med", "hi"), ordered=TRUE)
+noiseDiff$frac <- gseaGOFracS1$frac
+
+ggplot(noiseDiff, aes(x=frac, y=diff)) + geom_point() + facet_grid(. ~ sizeClass) + ggtitle("Sample - List, GSEA GO with Noise") + useTheme + xlab("Fraction") + ylab("Diff. log p-values")
+
+diffDiff <- data.frame(diff=(noiseDiff$diff - cleanDiff$diff), sizeClass=noiseDiff$sizeClass, frac=noiseDiff$frac)
+ggplot(diffDiff, aes(x=frac, y=diff)) + geom_point() + facet_grid(. ~ sizeClass) + ggtitle("Difference of Differences, GSEA GO") + useTheme + xlab("Fraction") + ylab("Diff. noise / noNoise")
+```
+
+So, it appears that having a uniform distribution does not make things behave the way we might expect.
+What if it is more an issue of independence, that GO is a bad fit for GSEA type analysis because of the issue that many of the terms are heavily overlapped.
+
+### Independence
+
+What if we force independence on our objects?
+i.e.
+each object has independent sets of genes, and treat them appropriately?
+i.e.
+to count as "diff" we will put genes (features) into the top 50% of ranked t-values (therefore reducing coverage means pushing annotated genes down into the bottom 50% of ranked values), and adding noise means to push un-annotated (i.e.
+random) above in rank.
+
+Lets try this for a single example.
+
+```{r independentGSEA}
+nGene <- 10000
+indepTValues <- sort(runif(nGene, -6, 6))
+names(indepTValues) <- as.character(seq(1, nGene))
+
+gseaMap <- list(t1=sample(seq(1, nGene/2), 100))
+
+tmpDat <- data.frame(tVal=indepTValues, status="all", stringsAsFactors=F)
+tmpDat$status[gseaMap[["t1"]]] <- "sample"
+ggplot(tmpDat, aes(x=tVal, fill=status)) + geom_histogram(binwidt=0.1, position="identity") + ggtitle("Check independent distribution, 100%") + useTheme + xlab("t-statistic")
+
+geneSetTest(gseaMap$t1, indepTValues, alternative="down", ranks.only=FALSE)
+
+gseaMap80 <- list(t1 = c(sample(seq(1, nGene/2), 80), 
+                         sample(seq((nGene/2) + 1, nGene), 20 )))
+
+geneSetTest(gseaMap80$t1, indepTValues, alternative="down", ranks.only=FALSE)
+
+tmpDat <- data.frame(tVal=indepTValues, status="all", stringsAsFactors=F)
+tmpDat$status[gseaMap80[["t1"]]] <- "sample"
+ggplot(tmpDat, aes(x=tVal, fill=status)) + geom_histogram(binwidt=0.1, position="identity") + ggtitle("check independent distribution, 80%") + useTheme + xlab("t-statistic")
+
+gseaMap50 <- list(t1 = c(sample(seq(1, nGene/2), 50),
+                         sample(seq((nGene/2) + 1, nGene), 50)))
+
+geneSetTest(gseaMap50$t1, indepTValues, alternative="down", ranks.only=FALSE)
+
+gseaMapUnif <- list(t1 = sample(nGene, 100))
+geneSetTest(gseaMapUnif$t1, indepTValues, alternative="down", ranks.only=FALSE)
+```
+
+OK, so at least we verify that as the percentage of genes in the top 50% decreases, then the p-values decrease as well, as we expect they should.
+What about as we increase the number of genes above above the set, while keeping the number in the top 50 the same?
+
+```{r checkNabove}
+gseaAbove100 <- list(t1=sample(seq(100, nGene/2), 100))
+geneSetTest(gseaAbove100$t1, indepTValues, alternative="down", ranks.only=FALSE)
+
+gseaAbove500 <- list(t1=sample(seq(501, nGene/2), 100))
+geneSetTest(gseaAbove500$t1, indepTValues, alternative="down", ranks.only=FALSE)
+
+gseaAbove1000 <- list(t1=sample(seq(1001, nGene/2), 100))
+geneSetTest(gseaAbove1000$t1, indepTValues, alternative="down", ranks.only=FALSE)
+
+gseaAbove2000 <- list(t1=sample(seq(2001, nGene/2), 100))
+geneSetTest(gseaAbove2000$t1, indepTValues, alternative="down", ranks.only=FALSE)
+
+gseaAbove4000 <- list(t1=sample(seq(4001, nGene/2), 100))
+geneSetTest(gseaAbove4000$t1, indepTValues, alternative="down", ranks.only=FALSE)
+```
+
+OK, so it takes a rather large proportion above to change the p-value any amount.
+
+So, what should we do. We need a function that will let us vary the size and proportion of genes that are in the bottom 50% of values.
+
+```{r gseaFunctions, eval=FALSE}
+?gseaSizeRange
+?gseaProportion
+```
+
+Lets verify this function is doing what we think it is.
+
+```{r checkGSEAProp}
+testSize <- gseaSizeRange(c(500,500), 50)
+useStatistics <- runif(10000, -6, 6)
+useSig <- runif(50, .1, 1)
+testStats <- gseaProportion(testSize, useSig, useStatistics)
+
+fullIndex <- seq(1, length(useStatistics))
+useCut <- 5000
+
+realRanks <- lapply(testStats, function(inStat){
+
+  apply(inStat$stats, 2, function(inCol){
+    outRank <- rank(inCol)[inStat$index]
+    sum(outRank < useCut) / length(inStat$index)
+  })
+
+})
+
+realRanks <- do.call(rbind, realRanks)
+plot(realRanks[,1], useSig)
+plot(realRanks[,2], useSig)
+plot(realRanks[,3], useSig)
+```
+
+OK, it looks like it is working. That's good.
+
+```{r testFunctions}
+testSize <- gseaSizeRange(c(500,500), 50)
+useStatistics <- runif(10000, -6, 6)
+useSig <- runif(50, .1, 1)
+testStats <- gseaProportion(testSize, useSig, useStatistics)
+
+outP <- lapply(testStats, function(inStat){
+  tmpP <- lapply(seq(1, ncol(inStat$stats)), function(useColumn){
+    geneSetTest(inStat$index, inStat$stats[,useColumn], "down", ranks.only=FALSE)
+  })
+  tmpP <- do.call(cbind, tmpP)
+  colnames(tmpP) <- colnames(inStat$stats)
+  return(tmpP)
+})
+
+outP2 <- do.call(rbind, outP)
+outP2 <- -1 * log(outP2)
+
+tmpP <- data.frame(set=apply(outP2[, c("s1", "s2")], 1, min), list=outP2[,"comb"])
+tmpP$diff <- tmpP$set - tmpP$list
+tmpP$frac <- useSig
+p <- ggplot(tmpP, aes(x=frac, y=diff)) + geom_point() + useTheme + ggtitle("Sample - Combined average t-statistics") + xlab("Fraction") + ylab("Diff. log p-values")
+p
+savePlot(p, "samplecombined_tstatistics.png")
+```
+
+Yeah, that doesn't seem to really work.
+
+What if GSEA is dependent on the **shared** proportion that is in the up or down state?
+i.e.
+if we allow the proportion "down" to vary between the two samples, and plot the difference in **set** and **list** values based on how **close** the value of the proportion in the top is??
+
+```{r gseaVaryProp, eval=FALSE}
+?gseaProportionVary
+```
+
+Check that we generate different significant proportions for independent samples, and what is the combined.
+
+```{r checkgseaProportionsVary}
+testSize <- gseaSizeRange(c(500,500), 50)
+useStatistics <- runif(10000, -6, 6)
+useSig <- list(s1=runif(50, .1, 1), s2=runif(50, .1, 1))
+testStats <- gseaProportionVary(testSize, useSig, useStatistics)
+
+fullIndex <- seq(1, length(useStatistics))
+useCut <- 5000
+
+realRanks <- lapply(testStats, function(inStat){
+
+  apply(inStat$stats, 2, function(inCol){
+    outRank <- rank(inCol)[inStat$index]
+    sum(outRank < useCut) / length(inStat$index)
+  })
+
+})
+
+realRanks <- do.call(rbind, realRanks)
+plot(realRanks[,1], useSig[[1]])
+plot(realRanks[,2], useSig[[2]])
+plot(realRanks[,3], useSig[[1]])
+plot(realRanks[,3], useSig[[2]])
+```
+
+OK, this seems to be working still.
+
+Now, lets try this on some fake data.
+
+```{r gseaProportions2Data}
+testSize <- gseaSizeRange(c(500,500), 50)
+useStatistics <- runif(10000, -6, 6)
+useSig <- list(s1=runif(50, .1, 1), s2=runif(50, .1, 1))
+testStats <- gseaProportionVary(testSize, useSig, useStatistics)
+
+outP <- lapply(testStats, function(inStat){
+  tmpP <- lapply(seq(1, ncol(inStat$stats)), function(useColumn){
+    geneSetTest(inStat$index, inStat$stats[,useColumn], "down", ranks.only=FALSE)
+  })
+  tmpP <- do.call(cbind, tmpP)
+  colnames(tmpP) <- colnames(inStat$stats)
+  return(tmpP)
+})
+
+outP2 <- do.call(rbind, outP)
+outP2 <- -1 * log(outP2)
+
+tmpP <- data.frame(set=apply(outP2[, c("s1", "s2")], 1, min), list=outP2[,"comb"])
+tmpP$diff <- tmpP$set - tmpP$list
+
+realRanks <- lapply(testStats, function(inStat){
+
+  apply(inStat$stats, 2, function(inCol){
+    outRank <- rank(inCol)[inStat$index]
+    sum(outRank < useCut) / length(inStat$index)
+  })
+
+})
+
+realRanks <- do.call(rbind, realRanks)
+
+tmpP$frac <- realRanks[,3]
+ggplot(tmpP, aes(x=frac, y=diff)) + geom_point() + ggtitle("Sample - Combined vary fraction") + useTheme + xlab("Fraction") + ylab("Diff. log p-values")
+```
+
+So, none of this has really worked, at least not in the way that was hoped.
+
+### GSEA using p-values instead of t-statistics
+
+For our hypothetical comparison, we need to think of how one would combine two studies.
+In this case, one might combine two studies by combining the p-values.
+Therefore, we will use ranked p-values as our metric in the `geneSetTest`.
+
+For this to work with `geneSetTest`, we need to make the range 0, 0.5, 1 conform to -1, 0, 1, because `geneSetTest` works nicer with signed values.
+We will use a linear regression to transform the p-values to a signed range.
+
+```{r generateRegression}
+regData <- data.frame(y=c(-1, 0, 1), x=c(0, 0.5, 1))
+regModel <- lm(y ~ x, regData)
+
+testData <- runif(10000, 0, 1)
+outData <- regModel$coefficients[1] + regModel$coefficients[2]*testData
+hist(outData, 100)
+```
+
+So our transformation works. Lets make sure it will work like the t-statistic.
+
+```{r pvalRanks}
+nGene <- 10000
+indepTValues <- sort(runif(nGene, 0, 1))
+names(indepTValues) <- as.character(seq(1, nGene))
+
+gseaMap <- list(t1=sample(seq(1, nGene/4), 100))
+
+tmpDat <- data.frame(tVal=indepTValues, status="all", stringsAsFactors=F)
+tmpDat$status[gseaMap[["t1"]]] <- "sample"
+ggplot(tmpDat, aes(x=tVal, fill=status)) + geom_histogram(binwidt=0.01, position="identity") + ggtitle("check p-values, 100%") + useTheme + xlab("P-values")
+
+geneSetTest(gseaMap$t1, p2signed(regModel,indepTValues), alternative="down", ranks.only=FALSE)
+
+gseaMap80 <- list(t1 = c(sample(seq(1, nGene/2), 80), 
+                         sample(seq((nGene/2) + 1, nGene), 20 )))
+
+geneSetTest(gseaMap80$t1, p2signed(regModel,indepTValues), alternative="down", ranks.only=FALSE)
+
+tmpDat <- data.frame(tVal=indepTValues, status="all", stringsAsFactors=F)
+tmpDat$status[gseaMap80[["t1"]]] <- "sample"
+ggplot(tmpDat, aes(x=tVal, fill=status)) + geom_histogram(binwidt=0.01, position="identity") + ggtitle("check p-values, 80%") + useTheme + xlab("P-values")
+
+gseaMap50 <- list(t1 = c(sample(seq(1, nGene/2), 50),
+                         sample(seq((nGene/2) + 1, nGene), 50)))
+
+geneSetTest(gseaMap50$t1, p2signed(regModel,indepTValues), alternative="down", ranks.only=FALSE)
+```
+
+Now, lets maintain proportions between samples as we did before, but for the **list** method, combine the p-values using Fisher's method.
+
+```{r gseaPropFisher, eval=FALSE}
+?gseaProportionFisher
+```
+
+```{r checkGSEAPropFisher}
+testSize <- gseaSizeRange(c(500,500), 50)
+useStatistics <- runif(10000, 0, 1)
+useSig <- runif(50, .1, 1)
+testStats <- gseaProportionFisher(testSize, useSig, useStatistics)
+
+fullIndex <- seq(1, length(useStatistics))
+useCut <- 5000
+
+realRanks <- lapply(testStats, function(inStat){
+
+  apply(inStat$stats, 2, function(inCol){
+    outRank <- rank(inCol)[inStat$index]
+    sum(outRank < useCut) / length(inStat$index)
+  })
+
+})
+
+realRanks <- do.call(rbind, realRanks)
+plot(realRanks[,1], useSig)
+plot(realRanks[,2], useSig)
+plot(realRanks[,3], useSig)
+```
+
+```{r testFishers}
+testSize <- gseaSizeRange(c(500,500), 50)
+useStatistics <- runif(10000, 0, 1)
+useSig <- runif(50, .1, 1)
+testStats <- gseaProportionFisher(testSize, useSig, useStatistics)
+
+outP <- lapply(testStats, function(inStat){
+  tmpP <- lapply(seq(1, ncol(inStat$stats)), function(useColumn){
+    geneSetTest(inStat$index, p2signed(regModel, inStat$stats[,useColumn]), "down", ranks.only=FALSE)
+  })
+  tmpP <- do.call(cbind, tmpP)
+  colnames(tmpP) <- colnames(inStat$stats)
+  return(tmpP)
+})
+
+outP2 <- do.call(rbind, outP)
+outP2 <- -1 * log(outP2)
+
+tmpP <- data.frame(set=apply(outP2[, c("s1", "s2")], 1, min), list=outP2[,"comb"])
+tmpP$diff <- tmpP$set - tmpP$list
+tmpP$frac <- useSig
+ggplot(tmpP, aes(x=frac, y=diff)) + geom_point() + ggtitle("Sample - Combined Fishers") + useTheme + xlab("Fraction") + ylab("Diff. log p-values")
+
+# try also using ranks only
+outPRank <- lapply(testStats, function(inStat){
+  tmpP <- lapply(seq(1, ncol(inStat$stats)), function(useColumn){
+    geneSetTest(inStat$index, inStat$stats, ranks.only=TRUE)
+  })
+  tmpP <- do.call(cbind, tmpP)
+  colnames(tmpP) <- colnames(inStat$stats)
+  return(tmpP)
+})
+
+outP2Rank <- do.call(rbind, outPRank)
+outP2Rank <- -1 * log(outP2Rank)
+
+tmpP <- data.frame(set=apply(outP2Rank[, c("s1", "s2")], 1, min), list=outP2Rank[,"comb"])
+tmpP$diff <- tmpP$set - tmpP$list
+tmpP$frac <- useSig
+ggplot(tmpP, aes(x=frac, y=diff)) + geom_point() + ggtitle("Sample - Combined Fishers Ranks") + useTheme + xlab("Fraction") + ylab("Diff. log p-values")
+```
+
+#### Maximum p-value
+
+Finally, we note that [Shen & Tseng from 2010](http://dx.doi.org/10.1093/bioinformatics/btq148) combined independent experiments by taking the **maximum** p-value between samples.
+
+```{r gseaProportionMax}
+?gseaProportionMax
+```
+
+```{r checkGSEAMax}
+testSize <- gseaSizeRange(c(500,500), 50)
+useStatistics <- runif(10000, 0, 1)
+useSig <- runif(50, .1, 1)
+testStats <- gseaProportionMax(testSize, useSig, useStatistics)
+
+fullIndex <- seq(1, length(useStatistics))
+useCut <- 5000
+
+realRanks <- lapply(testStats, function(inStat){
+
+  apply(inStat$stats, 2, function(inCol){
+    outRank <- rank(inCol)[inStat$index]
+    sum(outRank < useCut) / length(inStat$index)
+  })
+
+})
+
+realRanks <- do.call(rbind, realRanks)
+plot(realRanks[,1], useSig)
+plot(realRanks[,2], useSig)
+plot(realRanks[,3], useSig)
+```
+
+```{r runGSEAMax}
+testSize <- gseaSizeRange(c(500,500), 50)
+useStatistics <- runif(10000, 0, 1)
+useSig <- runif(50, .1, 1)
+testStats <- gseaProportionMax(testSize, useSig, useStatistics)
+
+outP <- lapply(testStats, function(inStat){
+  tmpP <- lapply(seq(1, ncol(inStat$stats)), function(useColumn){
+    geneSetTest(inStat$index, p2signed(regModel, inStat$stats[,useColumn]), "down", ranks.only=FALSE)
+  })
+  tmpP <- do.call(cbind, tmpP)
+  colnames(tmpP) <- colnames(inStat$stats)
+  return(tmpP)
+})
+
+outP2 <- do.call(rbind, outP)
+outP2 <- -1 * log(outP2)
+
+tmpP <- data.frame(set=apply(outP2[, c("s1", "s2")], 1, min), list=outP2[,"comb"])
+tmpP$diff <- tmpP$set - tmpP$list
+tmpP$frac <- useSig
+ggplot(tmpP, aes(x=frac, y=diff)) + geom_point() + ggtitle("Sample - Combined Max-P") + useTheme + xlab("Fraction") + ylab("Diff. log p-values")
+```
+
+And wow! It appears to work!
+
+Lets test using different size ranges and the proportions.
+
+```{r varySizesAndProps}
+gseaLo <- gseaSizeRange(c(20, 250), 50)
+gseaMed <- gseaSizeRange(c(250, 500), 30)
+gseaHi <- gseaSizeRange(c(500, 1000), 20)
+
+gseaSizes <- c(gseaLo, gseaMed, gseaHi)
+gseaSizeClass <- c(rep("low", 50), rep("med", 30), rep("hi", 20))
+
+gseaFrac <- runif(100, 0.3, 1)
+gseaStatistics <- runif(10000, 0, 1)
+
+testStats <- gseaProportionMax(gseaSizes, gseaFrac, gseaStatistics)
+
+outP <- lapply(testStats, function(inStat){
+  tmpP <- lapply(seq(1, ncol(inStat$stats)), function(useColumn){
+    geneSetTest(inStat$index, p2signed(regModel, inStat$stats[,useColumn]), "down", ranks.only=FALSE)
+  })
+  tmpP <- do.call(cbind, tmpP)
+  colnames(tmpP) <- colnames(inStat$stats)
+  return(tmpP)
+})
+
+outP2 <- do.call(rbind, outP)
+outP2 <- -1 * log(outP2)
+
+tmpP <- data.frame(set=apply(outP2[, c("s1", "s2")], 1, min), list=outP2[,"comb"])
+tmpP$diff <- tmpP$set - tmpP$list
+tmpP$frac <- gseaFrac
+tmpP$class <- gseaSizeClass
+tmpP$class <- factor(tmpP$class, c("low", "med", "hi"), ordered=TRUE)
+ggplot(tmpP, aes(x=frac, y=diff)) + geom_point() + facet_grid(. ~ class) + useTheme
+
+p <- ggplot(tmpP, aes(x=frac, y=diff)) + geom_point() + useTheme + ggtitle("Sample - Combined Max-P") + xlab("Fraction") + ylab("Diff. log p-values")
+p
+savePlot(p, "samplecombined_maxp.png")
+```
+
+So, this is very important. The method of combining p-values for the **list** method is **extremely** important.
+
+Does this hold using ranks only for the calculation?
+
+```{r maxPRankOnly}
+outPRank <- lapply(testStats, function(inStat){
+  tmpP <- lapply(seq(1, ncol(inStat$stats)), function(useColumn){
+    geneSetTest(inStat$index, inStat$stats[,useColumn], ranks.only=TRUE)
+  })
+  tmpP <- do.call(cbind, tmpP)
+  colnames(tmpP) <- colnames(inStat$stats)
+  return(tmpP)
+})
+
+outP2Rank <- do.call(rbind, outPRank)
+outP2Rank <- -1 * log(outP2Rank)
+
+tmpPRank <- data.frame(set=apply(outP2Rank[, c("s1", "s2")], 1, min), list=outP2Rank[,"comb"])
+tmpPRank$diff <- tmpPRank$set - tmpPRank$list
+tmpPRank$frac <- gseaFrac
+tmpPRank$class <- gseaSizeClass
+tmpPRank$class <- factor(tmpPRank$class, c("low", "med", "hi"), ordered=TRUE)
+ggplot(tmpPRank, aes(x=frac, y=diff)) + geom_point() + facet_grid(. ~ class) + ggtitle("Sample - Combined Max-P Ranks") + xlab("Fraction") + ylab("Diff. log p-values")
+```
+
+No, it doesn't because the `ranks.only=TRUE` will only check for **mixed** values, not at one end or the other.

### 2014-01-30 inst/doc/process_lung_data.Rmd
generated vignettes

<!--
  %\VignetteEngine{knitr}
%\VignetteIndexEntry{process lung data}
%\VignetteDepends{GEOquery}
-->

```{r customCSS, include=FALSE}
cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
options(markdown.HTML.stylesheet = cssFile)
```

# Lung Data

One of the experimental data sets we use is from [GEO][http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE18842]. This documents the actual processing.

```{r downloadData, eval=FALSE}
download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE18nnn/GSE18842/matrix/GSE18842_series_matrix.txt.gz", "/mlab/data/rmflight/Documents/projects/work/categoryComparePaperRev/otherData/GSE18842_series_matrix.txt.gz")
```

```{r loadResave, eval=FALSE}
library(GEOquery)
lung <- getGEO(filename="otherData/GSE18842_series_matrix.txt.gz")
tmpP <- pData(lung)
tmpP <- tmpP[,c("title", "geo_accession", "platform_id", "data_processing", "supplementary_file", "source_name_ch1", "characteristics_ch1", "characteristics_ch1.1")]
pData(lung) <- tmpP

tmpF <- fData(lung)
tmpF <- tmpF[, c("ID", "GB_ACC", "SPOT_ID", "Annotation Date", "Sequence Type", "Sequence Source", "Target Description", "Representative Public ID", "Gene Title", "Gene Symbol", "ENTREZ_GENE_ID", "RefSeq Transcript ID")]
fData(lung) <- tmpF
save(lung, file="inst/data/lung.RData", compress="xz")
```

### 2014-01-30 inst/doc/rcy_pieNodes.Rmd
generated vignettes

<!--
  %\VignetteEngine{knitr}
  %\VignetteIndexEntry{pie nodes in Cytoscape}
  %\VignetteDepends{RCytoscape, colorspace, categoryComparePaperRev}
-->

```{r customCSS, include=FALSE}
cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
options(markdown.HTML.stylesheet = cssFile)
```

# RCy Pie Nodes

If we want to create pie nodes in `RCy` using custom graphics, what do we need to do, and how hard will it be?

```{r exampleGraph, eval=FALSE}
library(RCytoscape)
g <- new ('graphNEL', edgemode='directed')
g <- graph::addNode ('A', g)
g <- graph::addNode ('B', g)
g <- graph::addNode ('C', g)
cw <- new.CytoscapeWindow ('vignette', graph=g)
displayGraph(cw)
layoutNetwork(cw, layout.name="grid")
redraw(cw)
```

OK, so we have now a little example graph to play with. How about we create some pie graphs and save them as png objects?

```{r pieGraphs}
pieData <- list(
  A = list(
    area = c(0.5, 0.5),
    color = c("blue", "green")),
  B = list(
    area = c(0.2, 0.8),
    color = c("red", "yellow")),
  C = list(
    area = c(0.6, 0.4),
    color = c("green", "purple"))
  )

plotPie <- function(inData){
  names(inData$area) <- ""
  pie(inData$area, col=inData$color, clockwise=TRUE)
}

plotPie(pieData[["A"]])
plotPie(pieData[["B"]])
plotPie(pieData[["C"]])
```

Sweet, little pie graphs. Next up, saving the graphics and letting Cytoscape know about them.

```{r savePie, eval=FALSE}
saveDir <- file.path(getwd(), "vignettes/")

outFiles <- sapply(names(pieData), function(pieName){
  pieFile <- paste("pieFile_", pieName, ".png", sep="")
  outFile <- file.path(saveDir, pieFile)
  png(outFile, bg="transparent")
  plotPie(pieData[[pieName]])
  dev.off()
  return(outFile)
})
names(outFiles) <- NULL
outFiles2 <- paste("file://localhost/", outFiles, sep="")
```

Now to try and associate them with the nodes in Cytoscape

```{r showInCy, eval=FALSE}
setNodeImageDirect(cw, names(pieData), outFiles2)
setDefaultNodeColor(cw, 'transparent')
setNodeOpacityDirect(cw, names(pieData), 0)
redraw(cw)
```

Finally, to use it in `categoryCompare`, I think we would always want the same number of **slices** for each node image, but use the saturated vs desaturated color to indicate significant vs not-significant, so the eye does not get fooled, and individuals can quickly scan to see which ones are which.

```{r desaturation}
library(colorspace)
desaturatePercentage <- function(col, percentage=0.5){
  # this code is taken from the desaturate function in colorspace
  if (is.character(col) && (all(substr(col, 1L, 1L) == "#") & 
    all(nchar(col) %in% c(7L, 9L)))) {
    alpha <- substr(col, 8L, 9L)
    col <- substr(col, 1L, 7L)
    col <- hex2RGB(col)
    }
  else {
    col <- col2rgb(col, alpha = TRUE)
    alpha <- format(as.hexmode(col[4L, ]), width = 2L, upper.case = TRUE)
    alpha[alpha == "FF"] <- ""
    col <- RGB(t(col[1L:3L, ])/255)
  }
  col <- as(col, "polarLUV")
  col@coords[, 2L] <- col@coords[, 2L] * percentage
  col <- hex(col)
  return(col)
}

twoColors <- c("red", desaturatePercentage("red", 0.25))
pie(c(0.5, 0.5), col=twoColors)
```

```{r genMultiplePies}
library(categoryComparePaperRev)

ccOpts <- new("ccOptions", listNames=c("T10_UP", "T10_DN", "T48_UP", "T48_DN", "T1_X", "T1_Y"), colorType="pie")

nTerm <- 1000

pieMatrix <- do.call(cbind, lapply(listNames(ccOpts), function(inName){
    baseCol <- rep(ccOpts@unsaturatedColor[inName], nTerm)
    baseCol[sample(nTerm, 500)] <- compareColors(ccOpts)[inName]
    baseCol
  }))

pieArea <- rep(1 / length(listNames(ccOpts)), length(listNames(ccOpts)))
names(pieArea) <- ""

```

### 2014-01-30 inst/doc/uc_crohns_analysis.Rmd
generated vignettes

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{UC vs Crohn's}
%\VignetteDepends{ccPaperRev, GEOquery, limma, hgu133plus2.db, GO.db, org.Hs.eg.db}
-->

```{r customCSS, include=FALSE}
cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
options(markdown.HTML.stylesheet = cssFile)
```

```{r loadLibrary, message=FALSE}
library(ccPaperRev)
```

# Crohn's VS Ulcerative Colitis

GEO dataset **GSE36807** has crohn's, uc, and normal samples.

```{r getData, eval=FALSE}
library(GEOquery)

cucData <- getGEO(GEO="GSE36807")[[1]]

tmpP <- pData(cucData)
tmpP$status <- "control"

cLoc <- grep("Crohn", tmpP$characteristics_ch1)
tmpP$status[cLoc] <- "crohns"

ucLoc <- grep("Ulcerative colitis", tmpP$characteristics_ch1)
tmpP$status[ucLoc] <- "uc"
pData(cucData) <- tmpP

.sessionInfo <- sessionInfo()
.timeDate <- Sys.time()
save(cucData, .sessionInfo, .timeDate, file="inst/data/uc_crohns_rawData.RData")
```

## Differential analysis

Use the `limma` package to rank probes.

```{r rankProbes}
library(limma)
library(hgu133plus2.db)
data(uc_crohns_rawData)

cucComps <- c("uc - control", "crohns - control")
geneID <- unlist(mget(featureNames(cucData), hgu133plus2ENTREZID))

cucExpr <- exprs(cucData)
cucCollapse <- collapseProbes(cucExpr, geneID) # collapse to single genes using median of expression
cucCharacter <- pData(cucData)

cucFC <- rankGenes(cucCollapse, cucCharacter$status, cucComps, doAggregation=FALSE, aggregateIndex=NA)
names(cucFC) <- c("UC", "CROHNS")

cucDiff <- lapply(cucFC, getDiffGenes, id="id")
# this object can be run using base ccEnrich, or can be easily coerced
```

## ORA

Do normal over-representation analysis on the `cucDiff`, those genes differentially expressed in the Crohn's and UC samples.

```{r ORA}
library(GO.db)
cucGeneList <- list(UC_up=list(genes=cucDiff$UC$up, universe=cucDiff$UC$universe, annotation="org.Hs.eg.db"),
                    UC_dn=list(genes=cucDiff$UC$dn, universe=cucDiff$UC$universe, annotation="org.Hs.eg.db"),
                    CR_up=list(genes=cucDiff$CROHNS$up, universe=cucDiff$CROHNS$universe, annotation="org.Hs.eg.db"),
                    CR_dn=list(genes=cucDiff$CROHNS$dn, universe=cucDiff$CROHNS$universe, annotation="org.Hs.eg.db"))
cucGeneList <- new("ccGeneList", cucGeneList, ccType="BP")

cucEnrich <- ccEnrich(cucGeneList)
pvalueType(cucEnrich) <- "pval"
pvalueCutoff(cucEnrich) <- 0.01

cucEnrich

cucOpts <- new("ccOptions", listNames=names(cucGeneList), compareNames=
                c("UC_up", "UC_dn", "CR_up", "CR_dn", "UC_up,CR_up", "UC_up,CR_dn", "UC_dn,CR_up", "UC_dn,CR_dn"))
cucCompare <- ccCompare(cucEnrich, cucOpts)

cucCompare

```

```{r out2Cytoscape, eval=FALSE}
cucCompare <- breakEdges(cucCompare$BP, 0.8)
cucCy <- ccOutCyt(cucCompare, cucOpts)
breakEdges(cucCy, 1)
```

## Run GSEA type analysis using limma

There are two ways to do GSEA using `limma`.

  1 Use `romer` for competitive tests with rotations to do random samples
  2 Use `geneSetTest` for gene-wise competitive tests, *i.e.* each set is tested against random gene samples

We're going to try both and see what we get out.

```{r setupGeneSets}
library(GO.db)
library(org.Hs.eg.db)
library(limma)
hsGO <- as.list(org.Hs.egGO2ALLEGS)
hsGO <- hsGO[(Ontology(names(hsGO)) == "BP")]
hsGO <- lapply(hsGO, unique)

useSet <- symbols2indices(hsGO, rownames(cucCollapse))
```

Run using `romer`

```{r setupContrasts}
sampleStatus <- cucCharacter$status
doComps <- cucComps
f <- factor(sampleStatus)
design <- model.matrix(~0 + f)
colnames(design) <- levels(f)

contrast.matrix <- makeContrasts(contrasts=doComps, levels=design)
```

```{r runRomer, eval=FALSE}
cucRomer <- multicontrastRomer(useSet, cucCollapse, design, contrast.matrix, nrot=10000)
names(cucRomer) <- c("UC", "CROHNS")
save(cucRomer, file="inst/data/cucRomer.RData")

```

```{r examineResults}
data(cucRomer)

geneAnnMapping <- new("namedList", .Data=hsGO, names=names(hsGO))

getSigID <- function(inRomer, pCut=0.05, whichCol=c("Up", "Down")){
  sigID <- lapply(whichCol, function(inCol){
    rownames(inRomer[(inRomer[, inCol] <= pCut),])
  })
  names(sigID) <- whichCol
  return(sigID)
}

cucSigRomer <- lapply(cucRomer, getSigID, pCut=0.01)
cucSigRomer <- unlist(cucSigRomer, recursive=FALSE)

genCCSigList <- function(inSig){
  tmp <- new("ccSigList", sigID=inSig)
}

cucCCSig <- lapply(cucSigRomer, genCCSigList)

cucCCRomer <- new("GENccEnrichResult", cucCCSig, categoryName="GSEAGO", geneAnnMapping=geneAnnMapping, overlapType="overlap", annDescription=Term(names(geneAnnMapping)))

cucRomerOpts <- new("ccOptions", listNames=names(cucCCRomer), compareNames=c(
  "UC.Up", "UC.Down", "CROHNS.Up", "CROHNS.Down", "UC.Up,CROHNS.Up", "UC.Up,CROHNS.Down", "UC.Down,CROHNS.Up", "UC.Down,CROHNS.Down"))
compareCUCRomer <- ccCompare(cucCCRomer, cucRomerOpts)
compareCUCRomer
```

```{r send2Cytoscape, eval=FALSE}
compareCUCRomer <- breakEdges(compareCUCRomer, 0.8)
cucRomerCW <- ccOutCyt(compareCUCRomer, cucRomerOpts, postText="romer", rpcPort=9001)
generateLegend(cucRomerOpts)
```

### 2014-01-30 inst/doc/hypotheticalExample_vignette.R

@@ -1,0 +1,924 @@
+
+## ----customCSS, include=FALSE--------------------------------------------
+cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+options(markdown.HTML.stylesheet = cssFile)
+
+
+## ----setupPlot-----------------------------------------------------------
+library(ggplot2)
+library(ggthemes)
+useTheme <- theme_bw() + theme(legend.key=element_blank(), legend.title = element_text(size=25), axis.title = element_text(size=25), axis.text = element_text(size=15), legend.text = element_text(size=20), strip.text.y = element_text(size=20), strip.text.x = element_text(size=20))
+
+
+savePlot <- function(plotObj, plotFile, figPath="/mlab/data/rmflight/Documents/projects/work/categoryComparePaperRev/savedPlots"){
+  if (file.exists(figPath)){
+    plotObj <- plotObj + ggtitle(NULL)
+    png(filename=file.path(figPath, plotFile), width=854, height=628)
+    print(plotObj)
+    dev.off()
+  }
+}
+
+
+## ----goAnnotations, message=FALSE----------------------------------------
+library(org.Hs.eg.db)
+library(GO.db)
+library(ccPaperRev)
+hsGO <- as.list(org.Hs.egGO2ALLEGS)
+
+goOntology <- Ontology(names(hsGO))
+goBP <- names(goOntology)[goOntology == "BP"]
+hsGO <- hsGO[goBP]
+hsGO <- lapply(hsGO, unique)
+universeGenes <- unique(unlist(hsGO))
+
+
+## ----goCountDist---------------------------------------------------------
+hsGO_count <- sapply(hsGO, length)
+hsGO_count <- data.frame(size=hsGO_count)
+p <- ggplot(hsGO_count, aes(x=size)) + geom_bar(binwidth=10) + xlim(0, 2000) + ylim(0, 500) + useTheme + ggtitle("Number of genes annotated to GO terms") + xlab('Num. of Genes') + ylab('Count')
+p
+
+savePlot(p, "go2gene_distribution.png")
+
+
+## ----defineGroupLimits---------------------------------------------------
+grpLow <- c(10, 100)
+grpMed <- c(250, 500)
+grpHi <- c(500, 1500)
+
+
[…]

### 2014-01-30 inst/doc/process_lung_data.R

@@ -1,0 +1,23 @@
+
+## ----customCSS, include=FALSE--------------------------------------------
+cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+options(markdown.HTML.stylesheet = cssFile)
+
+
+## ----downloadData, eval=FALSE--------------------------------------------
+## download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE18nnn/GSE18842/matrix/GSE18842_series_matrix.txt.gz", "/mlab/data/rmflight/Documents/projects/work/categoryComparePaperRev/otherData/GSE18842_series_matrix.txt.gz")
+
+
+## ----loadResave, eval=FALSE----------------------------------------------
+## library(GEOquery)
+## lung <- getGEO(filename="otherData/GSE18842_series_matrix.txt.gz")
+## tmpP <- pData(lung)
+## tmpP <- tmpP[,c("title", "geo_accession", "platform_id", "data_processing", "supplementary_file", "source_name_ch1", "characteristics_ch1", "characteristics_ch1.1")]
+## pData(lung) <- tmpP
+## 
+## tmpF <- fData(lung)
+## tmpF <- tmpF[, c("ID", "GB_ACC", "SPOT_ID", "Annotation Date", "Sequence Type", "Sequence Source", "Target Description", "Representative Public ID", "Gene Title", "Gene Symbol", "ENTREZ_GENE_ID", "RefSeq Transcript ID")]
+## fData(lung) <- tmpF
+## save(lung, file="inst/data/lung.RData", compress="xz")
+
+

### 2014-01-30 inst/doc/rcy_pieNodes.R

@@ -1,0 +1,107 @@
+
+## ----customCSS, include=FALSE--------------------------------------------
+cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+options(markdown.HTML.stylesheet = cssFile)
+
+
+## ----exampleGraph, eval=FALSE--------------------------------------------
+## library(RCytoscape)
+## g <- new ('graphNEL', edgemode='directed')
+## g <- graph::addNode ('A', g)
+## g <- graph::addNode ('B', g)
+## g <- graph::addNode ('C', g)
+## cw <- new.CytoscapeWindow ('vignette', graph=g)
+## displayGraph(cw)
+## layoutNetwork(cw, layout.name="grid")
+## redraw(cw)
+
+
+## ----pieGraphs-----------------------------------------------------------
+pieData <- list(
+  A = list(
+    area = c(0.5, 0.5),
+    color = c("blue", "green")),
+  B = list(
+    area = c(0.2, 0.8),
+    color = c("red", "yellow")),
+  C = list(
+    area = c(0.6, 0.4),
+    color = c("green", "purple"))
+  )
+
+plotPie <- function(inData){
+  names(inData$area) <- ""
+  pie(inData$area, col=inData$color, clockwise=TRUE)
+}
+
+plotPie(pieData[["A"]])
+plotPie(pieData[["B"]])
+plotPie(pieData[["C"]])
+
+
+## ----savePie, eval=FALSE-------------------------------------------------
+## saveDir <- file.path(getwd(), "vignettes/")
+## 
+## outFiles <- sapply(names(pieData), function(pieName){
+##   pieFile <- paste("pieFile_", pieName, ".png", sep="")
+##   outFile <- file.path(saveDir, pieFile)
+##   png(outFile, bg="transparent")
+##   plotPie(pieData[[pieName]])
+##   dev.off()
+##   return(outFile)
+## })
+## names(outFiles) <- NULL
+## outFiles2 <- paste("file://localhost/", outFiles, sep="")
+
+
+## ----showInCy, eval=FALSE------------------------------------------------
+## setNodeImageDirect(cw, names(pieData), outFiles2)
+## setDefaultNodeColor(cw, 'transparent')
+## setNodeOpacityDirect(cw, names(pieData), 0)
+## redraw(cw)
+
+
[…]

### 2014-01-30 inst/doc/responses2reviews.R

@@ -1,0 +1,6 @@
+
+## ----customCSS, include=FALSE--------------------------------------------
+cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+options(markdown.HTML.stylesheet = cssFile)
+
+

### 2014-01-30 inst/doc/skin_muscle_preprocessing.R

@@ -1,0 +1,246 @@
+
+## ----customCSS, include=FALSE--------------------------------------------
+cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+options(markdown.HTML.stylesheet = cssFile)
+
+
+## ----loadLibrary, message=FALSE------------------------------------------
+library(ccPaperRev)
+
+
+## ----loadLibraries-------------------------------------------------------
+options(stringsAsFactors=TRUE)
+library(affy)
+
+
+## ----skinRaw, eval=FALSE-------------------------------------------------
+## currLoc <- getwd()
+## dataLoc <- "/mlab/data/rmflight/Documents/projects/work/Petruska/drg_sc_skin_denervation/Skin/"
+## setwd(dataLoc)
+## 
+## attFile <- "Sample information file.txt"
+## dataAttributes <- read.table(attFile, sep="\t", header=T)
+## dataAttributes$timePoint <- 0
+## dataAttributes$timePoint[(dataAttributes$Time == "7 day")] <- 7
+## dataAttributes$timePoint[(dataAttributes$Time == "14 day")] <- 14
+## 
+## celData <- ReadAffy(phenoData=dataAttributes)
+## skinData <- rma(celData)
+## 
+## setwd(currLoc)
+## .sessionInfo <- sessionInfo()
+## .timeDate <- Sys.time()
+## save(skinData, .sessionInfo, .timeDate, file="inst/data/skin_rma_data.RData")
+
+
+## ----muscleRaw, eval=FALSE-----------------------------------------------
+## library(GEOquery)
+## muscleData <- getGEO(GEO="GSE4411")[[1]]
+## 
+## controlSamples <- grep("Control", pData(muscleData)$title)
+## muscleData <- muscleData[,controlSamples]
+## 
+## tmpP <- pData(muscleData)
+## tmpP$innervation <- "innervated"
+## denLoc <- grep("Denervated", tmpP$title)
+## tmpP$innervation[denLoc] <- "denervated"
+## pData(muscleData) <- tmpP
+## .sessionInfo <- sessionInfo()
+## .timeDate <- Sys.time()
+## save(muscleData, .sessionInfo, .timeDate, file="inst/data/muscle_data.RData")
+
+
+## ----diffSkin------------------------------------------------------------
+library(limma)
[…]

### 2014-01-30 inst/doc/uc_crohns_analysis.R

@@ -1,0 +1,141 @@
+
+## ----customCSS, include=FALSE--------------------------------------------
+cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+options(markdown.HTML.stylesheet = cssFile)
+
+
+## ----loadLibrary, message=FALSE------------------------------------------
+library(ccPaperRev)
+
+
+## ----getData, eval=FALSE-------------------------------------------------
+## library(GEOquery)
+## 
+## cucData <- getGEO(GEO="GSE36807")[[1]]
+## 
+## tmpP <- pData(cucData)
+## tmpP$status <- "control"
+## 
+## cLoc <- grep("Crohn", tmpP$characteristics_ch1)
+## tmpP$status[cLoc] <- "crohns"
+## 
+## ucLoc <- grep("Ulcerative colitis", tmpP$characteristics_ch1)
+## tmpP$status[ucLoc] <- "uc"
+## pData(cucData) <- tmpP
+## 
+## .sessionInfo <- sessionInfo()
+## .timeDate <- Sys.time()
+## save(cucData, .sessionInfo, .timeDate, file="inst/data/uc_crohns_rawData.RData")
+
+
+## ----rankProbes----------------------------------------------------------
+library(limma)
+library(hgu133plus2.db)
+data(uc_crohns_rawData)
+
+cucComps <- c("uc - control", "crohns - control")
+geneID <- unlist(mget(featureNames(cucData), hgu133plus2ENTREZID))
+
+cucExpr <- exprs(cucData)
+cucCollapse <- collapseProbes(cucExpr, geneID) # collapse to single genes using median of expression
+cucCharacter <- pData(cucData)
+
+cucFC <- rankGenes(cucCollapse, cucCharacter$status, cucComps, doAggregation=FALSE, aggregateIndex=NA)
+names(cucFC) <- c("UC", "CROHNS")
+
+cucDiff <- lapply(cucFC, getDiffGenes, id="id")
+# this object can be run using base ccEnrich, or can be easily coerced
+
+
+## ----ORA-----------------------------------------------------------------
+library(GO.db)
+cucGeneList <- list(UC_up=list(genes=cucDiff$UC$up, universe=cucDiff$UC$universe, annotation="org.Hs.eg.db"),
[…]

### 2014-01-31 R/datasets.R
adding the GO group results in crohns vs uc, both the saving, and the actual data itself.

@@ -58,3 +58,11 @@ NULL
 #' @source running \code{mmGeneSetTest} on muscle denervated over innervated
 #' @details did one contrast, denervated vs innervated using the \code{mmGeneSetTest} function
 NULL
+
+
+#' @name cucCCOutPut.RData
+#' @title cucCCOutPut.RData
+#' @docType data
+#' @source interacting with GO nodes in Cytoscape to partition into groups
+#' @details After performing \code{ccCompare} on \code{cucRomer}, any edges =< 0.8 similarity were removed, and the graph loaded into Cytoscape using \code{ccOutCyt}. Individual groups of GO nodes were determined by Robert M Flight visualy and exploratively, and labels assigned based on biological knowledge of Robert M Flight. Each GO group is defined in \code{cucNodeGroups}, with the list of GO terms that are part of that group, along with their attributes.
+NULL

### 2014-02-05 R/datasets.R
changed cucCCOutput to the correct file name

@@ -60,8 +60,8 @@ NULL
 NULL
 
 
-#' @name cucCCOutPut.RData
-#' @title cucCCOutPut.RData
+#' @name cucCCOutput.RData
+#' @title cucCCOutput.RData
 #' @docType data
 #' @source interacting with GO nodes in Cytoscape to partition into groups
 #' @details After performing \code{ccCompare} on \code{cucRomer}, any edges =< 0.8 similarity were removed, and the graph loaded into Cytoscape using \code{ccOutCyt}. Individual groups of GO nodes were determined by Robert M Flight visualy and exploratively, and labels assigned based on biological knowledge of Robert M Flight. Each GO group is defined in \code{cucNodeGroups}, with the list of GO terms that are part of that group, along with their attributes.

### 2014-02-05 R/datasets.R
added added skinMuscleCCOutput

@@ -66,3 +66,10 @@ NULL
 #' @source interacting with GO nodes in Cytoscape to partition into groups
 #' @details After performing \code{ccCompare} on \code{cucRomer}, any edges =< 0.8 similarity were removed, and the graph loaded into Cytoscape using \code{ccOutCyt}. Individual groups of GO nodes were determined by Robert M Flight visualy and exploratively, and labels assigned based on biological knowledge of Robert M Flight. Each GO group is defined in \code{cucNodeGroups}, with the list of GO terms that are part of that group, along with their attributes.
 NULL
+
+#' @name skinMuscleCCOutput.RData
+#' @title skinMuscleCCOutput.RData
+#' @docType data
+#' @source interacting with GO nodes in Cytoscape to partition into groups
+#' @details After performing \code{ccCompare} on \code{allCCRomer}, any edges =< 0.8 similarity were removed, and the graph loaded into Cytoscape using \code{ccOutCyt}. Individual groups of GO nodes were determined by Robert M Flight visualy and exploratively, and labels assigned based on biological knowledge of Robert M Flight. Each GO group is defined in \code{smNodes}, with the list of GO terms that are part of that group, along with their attributes.
+NULL

### 2014-02-10 R/hypotheticalDataGeneration.R
changed package names to work with new ccPaper and categoryComparePaper

@@ -117,8 +117,8 @@ calcFraction <- function(go2gene, geneList){
 #' @param universe \code{character} vector of the background (assumed to be common across all)
 #' @param ontology which GO ontology to use (default is "BP")
 #' @param annotation what is the source of annotation (default is "org.Hs.eg.db")
-#' @importClassesFrom categoryComparePaperRev GOHyperGParamsCC
-#' @importFrom categoryComparePaperRev hyperGTestCC
+#' @importClassesFrom categoryComparePaper GOHyperGParamsCC
+#' @importFrom categoryComparePaper hyperGTestCC
 #' @return \code{list}, see Details
 #' @details the object returned will have an object for each sample list, as well as the one generated from the intersection
 #' @export

### 2014-02-11 inst/doc/responses2reviews.Rmd
regenerated vignettes based on the new package names

@@
 Set 1500 diff expressed genes from the genes annotated with GO Terms chosen above, and apply the two methods, and see how they are different.
+
+# Major Things to Address after Round #2
+
+## Reviewer #1 is happy
+
+pending GEO submission and revised discussion. Yay! Note: we still need to write a kick-ass discussion.
+
+* GEO submission - Jeff and Ben are working on it
+* Discussion - Eric and I are working on it
+
+## Reviewer #3 still seems to feel that the major contribution is the visualization, and is rather confused by the simulation. This needs to be addressed in some detail.
+
+Doesn't want so much emphasis that categoryCompare is better, but that the methods may be complementary.
+  * This is actually borne out by the simulations: In some cases feature combining worked better, in other cases annotation combining worked better.
+So it is probably not a bad idea to use both, as was advocated by Shen and Tseng.
+
+Currently our major contribution is showing using simulations *why* you might want to combine datasets using annotations instead of features, and implementing a method that is rather generic in `R`.
+
+### Specific Points:
+
+(316) "Traditional methods for comparing high-throughput data sets depend on finding shared features between significant feature lists for each data set and then performing some form of enrichment analysis on the overlapping feature sets to determine biological themes among the shared features"; (329) "Figure 1 illustrates this contrast.
+As a simple example, consider two feature or annotation lists from two different data sets (“Start”, middle box of Figure 1).
+In the “Traditional Analysis”, the intersection and set-difference of the features are determined, and biological interpretation provided by annotation enrichment of each of the resulting lists."
+
+* It is also common to test the genomic features from the experiments independently, and then compare the results in a table or heat map, without extra visualization support for the similarity relations among annotations.
+You should probably state this explicitly, and extend figure 1.
+
+(321) "However, there are many different scenarios that complicate this analysis, including cases when
+different types of features are measured (i.e. two studies using different microarray platforms), or
+comparing studies that examine different types of features that may share the same annotations
+(transcriptomics vs. metabolomics for example)."
+
+* Typically, two studies using different microarray platforms measure the same *type* of features, but the actual feature instances may be only partially (or not) overlapping.
+Also: if different types of features do not share any annotation, it is not possible to compare them anyway.
+
+(339) "When no feature-annotation relationships are available, the network considers the lists themselves as nodes, with edges to annotations that were derived from that list."
+
+* not sure this functionality is helpful at all: if there are no relation, a graph offers no visualization support to offer relations in a more intuitive / intelligible form
+
+Regarding Results and Methods:
+
+These are major additions.
+
+I understand the simulations were added to demonstrate that considering all annotations significant in at least one condition, rather than annotations significant only for the overlapping genomic features, does not inflate the number of noisy annotations.
+This is a legitimate concern, but as presented now it look as the main point of the paper.
+This should be addressed in a subsection of the results, and part of the details should be put in the methods section; the figure should also be modified to convey this message more clearly and concisely.
+
+That said, I am skeptical about proving the superiority of categorycompare mainly by simulations.
+This is a visualization method, the performance must be measured by how it helps a researcher to productively interpret data.
+This cannot be measured by simulation.
+
+Right now, I find the results partially overwhelming, confusing and unconvincing.
+I would like to see two or three examples where a multi-condition publicly-available data-set is presented, the biological question is concisely explained, and what can be gleaned by the "traditional analysis" as well as "category compare" are presented, with corresponding graphics.
+If a researcher reads this, he should agree the method is useful for his research.
+Mind the two approaches may as well be complementary, showing different aspects of the data; this conclusion would not weaken at all this article.
+Conclusions from simulations could be discussed, but, in the current form, the amount of work done on simulations and details presented in the result section does not translate to a sufficiently convincing message about the method's performance.
+
+I encourage the authors to do some more work to address these concerns as they fell appropriate.
+
+Ability to import GSEA type results:
+
+I think what I suggested is very easy to implement: a user should just export his enrichment results in a simple and standardized format and then upload in categorycompare.

### 2014-02-11 inst/doc/hypotheticalExample_vignette.Rmd
regenerated vignettes based on the new package names

@@
   %\VignetteIndexEntry{hypothetical example}
-  %\VignetteDepends{categoryComparePaperRev, ggplot2, ALL, GO.db, hgu95av.db, hgu133plus2.db, limma}
+  %\VignetteDepends{categoryComparePaper, ggplot2, ALL, GO.db, hgu95av.db, hgu133plus2.db, limma}
 -->
@@
 ```{r customCSS, include=FALSE}
-cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+cssFile <- system.file("extdata", "style.css", package="ccPaper")
 options(markdown.HTML.stylesheet = cssFile)
 ```
+
+Note that running this analysis will require at least 4 GB of RAM unless objects are deleted when they are no longer required.
 
@@
 library(ggplot2)
-library(ggthemes)
 useTheme <- theme_bw() + theme(legend.key=element_blank(), legend.title = element_text(size=25), axis.title = element_text(size=25), axis.text = element_text(size=15), legend.text = element_text(size=20), strip.text.y = element_text(size=20), strip.text.x = element_text(size=20))
 
-savePlot <- function(plotObj, plotFile, figPath="/mlab/data/rmflight/Documents/projects/work/categoryComparePaperRev/savedPlots"){
+savePlot <- function(plotObj, plotFile, figPath="/mlab/data/rmflight/Documents/projects/work/ccPaper/savedPlots"){
   if (file.exists(figPath)){
@@
 library(GO.db)
-library(ccPaperRev)
+library(ccPaper)
 hsGO <- as.list(org.Hs.egGO2ALLEGS)
@@
 sfExport("useGOSizeClass", "universeGenes", "useGO")
-sfLibrary(ccPaperRev)
+sfLibrary(ccPaper)
 sfLibrary(GO.db)
@@
 sfInit(parallel=TRUE, cpus=10)
-sfLibrary(ccPaperRev)
+sfLibrary(ccPaper)
 sfExport("hsGO", "nGO", "goLimits")
@@
 sfInit(parallel=TRUE, cpus=10)
-sfLibrary(ccPaperRev)
+sfLibrary(ccPaper)
 sfLibrary(org.Hs.eg.db)

### 2014-02-11 inst/doc/uc_crohns_analysis.Rmd
regenerated vignettes based on the new package names

@@
 %\VignetteIndexEntry{UC vs Crohn's}
-%\VignetteDepends{ccPaperRev, GEOquery, limma, hgu133plus2.db, GO.db, org.Hs.eg.db}
+%\VignetteDepends{ccPaper, GEOquery, limma, hgu133plus2.db, GO.db, org.Hs.eg.db}
 -->
@@
 ```{r customCSS, include=FALSE}
-cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+cssFile <- system.file("extdata", "style.css", package="ccPaper")
 options(markdown.HTML.stylesheet = cssFile)
@@
 ```{r loadLibrary, message=FALSE}
-library(ccPaperRev)
+library(ccPaper)
 ```
@@
 ```
+
+Now we will save groups so that they can be queried and examined later.
+
+```{r saveNodeGroups, eval=FALSE}
+cucNodeGroups <- cytOutNodes("response to lipopolysaccharide and bacterial - UC.Up", cucRomerCW)
+cucNodeGroups <- cytOutNodes("regulation of inflammatory response - UC.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("hydrogen peroxide metabolism - UC.Up,CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("regulation of cell cycle and DNA damage response - UC.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("regulation of ubiquitination and ligase activity - UC.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("nucleotide and nucleoside metabolism - UC.Up,CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("amine metabolism - UC.Up,CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("glandular cell differentiation - UC.Down", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("oligodendrocyte differentiation - UC.Down,CROHNS.Down", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("cellular pattern specification - UC.Down,CROHNS.Down", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("NAD biosynthesis - CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("hormone metabolism - CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("extrinsic signal transduction - UC.Up,CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("regulation of nitric-oxide synthase - UC.Up,CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("fatty-acyl-CoA biosynthesis - UC.Up,CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("response to growth hormone - CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("melanin metabolism - CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("protein dephosphorylation - CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("chemokine and cytokine production - UC.Up,CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("membrane biogenesis and assembly - UC.Down", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("activin receptor signaling - UC.Down", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("nik/nk-kappab cascade - UC.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("er unfolded protein response - UC.Up, CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("regulation of ras/rac/rho gtpase activity - UC.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("antigen processing and presentation - UC.UP,CROHNS.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("COPII vesicle coating and targeting - UC.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("negative regulation of peptidase activity - UC.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("response to type 1 interferon - UC.Up", cucRomerCW, cucNodeGroups)
+cucNodeGroups <- cytOutNodes("protein N-linked glycosylation - UC.Up", cucRomerCW, cucNodeGroups)
+.sessionInfo <- sessionInfo()
+.timeDate <- Sys.time()
+save(cucRomerCW, cucNodeGroups, compareCUCRomer, cucRomerOpts, .sessionInfo, .timeDate, file="inst/data/cucCCOutput.RData")
+```
+
+We can then write out the results into a nice table.
+
+```{r writeTableResults}
+data(cucCCOutput)
+allDescStrings <- sapply(cucNodeGroups, function(x){x$descStr})
+string2List <- strsplit(allDescStrings, " - ", fixed=TRUE)
+justDesc <- sapply(string2List, function(x){x[1]})
+listMem <- sapply(string2List, function(x){x[2]})
+listMemSplit <- strsplit(listMem, ",", fixed=TRUE)
+
+descMembershipTable <- matrix("", nrow=length(allDescStrings), ncol=5)
+colnames(descMembershipTable) <- c("Description", "UC.Down", "UC.Up", "CROHNS.Down", "CROHNS.Up")
+
+descMembershipTable[,"Description"] <- justDesc
+
+for (inRow in seq(1, nrow(descMembershipTable))){
+  useSplit <- listMemSplit[[inRow]]
+  trimSplit <- gsub(" ", "", useSplit)
+  useLocs <- sapply(trimSplit, grep, colnames(descMembershipTable), ignore.case=TRUE)
+  descMembershipTable[inRow, useLocs] <- "X"
+}
+
+orderBy <- c("UC.Up,CROHNS.Up",
+             "UC.UP,CROHNS.Up",
+             "UC.Up",
+             "UC.Down",
+             "UC.Down,CROHNS.Down",
+             "CROHNS.Up",
+             "CROHNS.Down")
+
+listMem <- gsub(" ", "", listMem)
+newOrder <- unlist(lapply(orderBy, function(x){
+  which(listMem %in% x)
+}))
+
+descMembershipTable <- descMembershipTable[newOrder,]
+require(xtable)
+
+# add an html link to each entry in the table
+useLink <- paste('<a href="#loc', seq(1, nrow(descMembershipTable)), '">', descMembershipTable[,"Description"], '</a>', sep="")
+descMembershipTable[, "Description"] <- useLink
+```
+
+Print out the table.
+
+```{r printTable, echo=FALSE, results='asis'}
+cat('<a name="tableLink"></a>\n') # a link back to the table
+print(xtable(descMembershipTable), type = 'html', html.table.attributes = 'style="border-spacing:20px 5px;"', include.rownames=FALSE, sanitize.text.function=function(x){x})
+```
+
+We also want the set of GO terms, their descriptions, and list memberships so that the inquisitive mind can see how we generated the groups.
+
+```{r detailTable, echo=FALSE, results='asis'}
+cucNodeGroups <- cucNodeGroups[newOrder]
+
+for (iNode in seq(1:length(cucNodeGroups))){
+  nodeSet <- cucNodeGroups[[iNode]]
+  toLink <- paste('<a name="loc', iNode, '"></a>', sep="")
+  cat(toLink, "\n")
+  cat("####", nodeSet$descStr, sep=" ")
+  cat("\n")
+
+  nGO <- length(nodeSet$nodes)
+  outMatrix <- matrix("", nrow=nGO, ncol=6)
+  colnames(outMatrix) <- c("GOID", "Description", "UC.Down", "UC.Up", "CROHNS.Down", "CROHNS.Up")
+  rownames(outMatrix) <- nodeSet$nodes
+  outMatrix[,"GOID"] <- nodeSet$nodes
+
+  for (useGO in nodeSet$nodes){
+    outMatrix[useGO, "Description"] <- nodeSet$nodeData[[useGO]]$Desc
+    listMems <- strsplit(nodeSet$nodeData[[useGO]]$listMembership, ",")[[1]]
+    listMems <- listMems[(nchar(listMems) > 0)]
+    outMatrix[useGO, listMems] <- "X"
+  }
+  allMems <- sapply(nodeSet$nodeData, function(x){x$listMembership})
+  reOrder <- order(allMems, decreasing=TRUE)
+  outMatrix <- outMatrix[reOrder,]
+  print(xtable(outMatrix), type = 'html', html.table.attributes = 'style="border-spacing:20px 5px;"', include.rownames=FALSE)
+  cat('\n<a href="#tableLink">back</a>\n')
+}
+```

### 2014-02-11 inst/doc/hypotheticalExample_vignette.R

@@ -1,16 +1,15 @@
 
 ## ----customCSS, include=FALSE--------------------------------------------
-cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+cssFile <- system.file("extdata", "style.css", package="ccPaper")
 options(markdown.HTML.stylesheet = cssFile)
 
 
 ## ----setupPlot-----------------------------------------------------------
 library(ggplot2)
-library(ggthemes)
 useTheme <- theme_bw() + theme(legend.key=element_blank(), legend.title = element_text(size=25), axis.title = element_text(size=25), axis.text = element_text(size=15), legend.text = element_text(size=20), strip.text.y = element_text(size=20), strip.text.x = element_text(size=20))
 
 
-savePlot <- function(plotObj, plotFile, figPath="/mlab/data/rmflight/Documents/projects/work/categoryComparePaperRev/savedPlots"){
+savePlot <- function(plotObj, plotFile, figPath="/mlab/data/rmflight/Documents/projects/work/ccPaper/savedPlots"){
   if (file.exists(figPath)){
     plotObj <- plotObj + ggtitle(NULL)
     png(filename=file.path(figPath, plotFile), width=854, height=628)
@@ -23,7 +22,7 @@ savePlot <- function(plotObj, plotFile, figPath="/mlab/data/rmflight/Documents/p
 ## ----goAnnotations, message=FALSE----------------------------------------
 library(org.Hs.eg.db)
 library(GO.db)
-library(ccPaperRev)
+library(ccPaper)
 hsGO <- as.list(org.Hs.egGO2ALLEGS)
 
 goOntology <- Ontology(names(hsGO))
@@ -247,7 +246,7 @@ tapply(diff_pvalues$diff, diff_pvalues$sizeClass, mean)
 ## }
 ## sfInit(parallel=TRUE, cpus=10)
 ## sfExport("useGOSizeClass", "universeGenes", "useGO")
-## sfLibrary(ccPaperRev)
+## sfLibrary(ccPaper)
 ## sfLibrary(GO.db)
 ## sfLibrary(org.Hs.eg.db)
 ## testRes <- sfLapply(testSamples, testFun)
[…]

### 2014-02-11 inst/doc/process_lung_data.R

@@ -1,11 +1,11 @@
 
 ## ----customCSS, include=FALSE--------------------------------------------
-cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+cssFile <- system.file("extdata", "style.css", package="ccPaper")
 options(markdown.HTML.stylesheet = cssFile)
 
 
 ## ----downloadData, eval=FALSE--------------------------------------------
-## download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE18nnn/GSE18842/matrix/GSE18842_series_matrix.txt.gz", "/mlab/data/rmflight/Documents/projects/work/categoryComparePaperRev/otherData/GSE18842_series_matrix.txt.gz")
+## download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE18nnn/GSE18842/matrix/GSE18842_series_matrix.txt.gz", "/mlab/data/rmflight/Documents/projects/work/ccPaper/otherData/GSE18842_series_matrix.txt.gz")
 
 
 ## ----loadResave, eval=FALSE----------------------------------------------

### 2014-02-11 inst/doc/rcy_pieNodes.R

@@ -1,6 +1,6 @@
 
 ## ----customCSS, include=FALSE--------------------------------------------
-cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+cssFile <- system.file("extdata", "style.css", package="ccPaper")
 options(markdown.HTML.stylesheet = cssFile)
 
 
@@ -88,7 +88,7 @@ pie(c(0.5, 0.5), col=twoColors)
 
 
 ## ----genMultiplePies-----------------------------------------------------
-library(categoryComparePaperRev)
+library(categoryComparePaper)
 
 ccOpts <- new("ccOptions", listNames=c("T10_UP", "T10_DN", "T48_UP", "T48_DN", "T1_X", "T1_Y"), colorType="pie")
 

### 2014-02-11 inst/doc/skin_muscle_preprocessing.R

@@ -1,11 +1,11 @@
 
 ## ----customCSS, include=FALSE--------------------------------------------
-cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+cssFile <- system.file("extdata", "style.css", package="ccPaper")
 options(markdown.HTML.stylesheet = cssFile)
 
 
 ## ----loadLibrary, message=FALSE------------------------------------------
-library(ccPaperRev)
+library(ccPaper)
 
 
 ## ----loadLibraries-------------------------------------------------------
@@ -229,6 +229,187 @@ compareSMRomer
 ## ----out2cytoscape, eval=FALSE-------------------------------------------
 ## compareSMRomer <- breakEdges(compareSMRomer, 0.8)
 ## cwSM <- ccOutCyt(compareSMRomer, allRomerOpts, postText="GSEAGO", rpcPort=9001)
+
+
+## ----saveCytResults, eval=FALSE------------------------------------------
+## .sessionInfo <- sessionInfo()
+## .timeDate <- Sys.time()
+## save(cwSM, compareSMRomer, allRomerOpts, .sessionInfo, .timeDate, file="inst/data/skinMuscleCompareOutput.RData")
+
+
+## ----outGOTerms, eval=FALSE----------------------------------------------
+## smNodes <- cytOutNodes("typeII hypersensitivity - T7.Up", cwSM)
+## smNodes <- cytOutNodes("pH and lysosome regulation - Skin.T7.Down,Skin.T14.Up,Muscle.Up", cwSM, smNodes)
+## smNodes <- cytOutNodes("Neurotransmitter secretion and transport - Skin.T7.Down,Skin.T14.Up", cwSM, smNodes)
+## smNodes <- cytOutNodes("entry into host and movement - Skin.T7.Up", cwSM, smNodes)
+## smNodes <- cytOutNodes("steriod biosynthesis - Skin.T7.Down", cwSM, smNodes)
+## smNodes <- cytOutNodes("microtubule organization - Skin.T7.Up,Skin.T14.Down", cwSM, smNodes)
+## smNodes <- cytOutNodes("cellular respiration - Skin.T14.Down,Muscle.Down", cwSM, smNodes)
[…]

### 2014-02-11 inst/doc/uc_crohns_analysis.R

@@ -1,11 +1,11 @@
 
 ## ----customCSS, include=FALSE--------------------------------------------
-cssFile <- system.file("extdata", "style.css", package="ccPaperRev")
+cssFile <- system.file("extdata", "style.css", package="ccPaper")
 options(markdown.HTML.stylesheet = cssFile)
 
 
 ## ----loadLibrary, message=FALSE------------------------------------------
-library(ccPaperRev)
+library(ccPaper)
 
 
 ## ----getData, eval=FALSE-------------------------------------------------
@@ -137,5 +137,116 @@ compareCUCRomer
 ## compareCUCRomer <- breakEdges(compareCUCRomer, 0.8)
 ## cucRomerCW <- ccOutCyt(compareCUCRomer, cucRomerOpts, postText="romer", rpcPort=9001)
 ## generateLegend(cucRomerOpts)
+
+
+## ----saveNodeGroups, eval=FALSE------------------------------------------
+## cucNodeGroups <- cytOutNodes("response to lipopolysaccharide and bacterial - UC.Up", cucRomerCW)
+## cucNodeGroups <- cytOutNodes("regulation of inflammatory response - UC.Up", cucRomerCW, cucNodeGroups)
+## cucNodeGroups <- cytOutNodes("hydrogen peroxide metabolism - UC.Up,CROHNS.Up", cucRomerCW, cucNodeGroups)
+## cucNodeGroups <- cytOutNodes("regulation of cell cycle and DNA damage response - UC.Up", cucRomerCW, cucNodeGroups)
+## cucNodeGroups <- cytOutNodes("regulation of ubiquitination and ligase activity - UC.Up", cucRomerCW, cucNodeGroups)
+## cucNodeGroups <- cytOutNodes("nucleotide and nucleoside metabolism - UC.Up,CROHNS.Up", cucRomerCW, cucNodeGroups)
+## cucNodeGroups <- cytOutNodes("amine metabolism - UC.Up,CROHNS.Up", cucRomerCW, cucNodeGroups)
+## cucNodeGroups <- cytOutNodes("glandular cell differentiation - UC.Down", cucRomerCW, cucNodeGroups)
+## cucNodeGroups <- cytOutNodes("oligodendrocyte differentiation - UC.Down,CROHNS.Down", cucRomerCW, cucNodeGroups)
[…]

### 2014-02-15 inst/doc/hypotheticalExample_vignette.R
updated vignettes, correct packages, and dates and package versions

@@ -811,8 +811,8 @@ tmpP$frac <- useSig
 ggplot(tmpP, aes(x=frac, y=diff)) + geom_point() + ggtitle("Sample - Combined Fishers Ranks") + useTheme + xlab("Fraction") + ylab("Diff. log p-values")
 
 
-## ----gseaProportionMax---------------------------------------------------
-?gseaProportionMax
+## ----gseaProportionMax, eval=FALSE---------------------------------------
+## ?gseaProportionMax
 
 
 ## ----checkGSEAMax--------------------------------------------------------
@@ -919,5 +919,10 @@ tmpPRank$frac <- gseaFrac
 tmpPRank$class <- gseaSizeClass
 tmpPRank$class <- factor(tmpPRank$class, c("low", "med", "hi"), ordered=TRUE)
 ggplot(tmpPRank, aes(x=frac, y=diff)) + geom_point() + facet_grid(. ~ class) + ggtitle("Sample - Combined Max-P Ranks") + xlab("Fraction") + ylab("Diff. log p-values")
+
+
+## ------------------------------------------------------------------------
+Sys.time()
+sessionInfo()
 
 

### 2014-02-15 inst/doc/skin_muscle_preprocessing.R

@@ -9,7 +9,7 @@ library(ccPaper)
 
 
 ## ----loadLibraries-------------------------------------------------------
-options(stringsAsFactors=TRUE)
+options(stringsAsFactors=FALSE)
 library(affy)
 
 
@@ -320,6 +320,8 @@ compareSMRomer
 ## smNodes <- cytOutNodes("aminoglycan metabolism - Skin.T7.Up", cwSM, smNodes)
 ## smNodes <- cytOutNodes("epithelial to mesenchymal transition, endocardial cushion formation - Skin.T7.Up", cwSM, smNodes)
 ## smNodes <- cytOutNodes("pulmonary valve dev. and morphogenesis - Skin.T7.Up", cwSM, smNodes)
+## .sessionInfo <- sessionInfo()
+## .timeDate <- Sys.time()
 ## save(smNodes, cwSM, allCCRomer, allRomerOpts, compareSMRomer, file="inst/data/skinMuscleCCOutput.RData")
 
 

### 2014-02-15 inst/doc/uc_crohns_analysis.R

@@ -250,3 +250,41 @@ for (iNode in seq(1:length(cucNodeGroups))){
 }
 
 
+## ----getAllData----------------------------------------------------------
+library(graph)
+data(cucRomer)
+data(cucCCOutput)
+
+getTable <- function(inName){
+  tmpDat <- cucRomer[[inName]]
+  colnames(tmpDat) <- paste(inName, colnames(tmpDat), sep=".")
+  tmpDat
+}
+
+cucTables <- lapply(names(cucRomer), getTable)
+all.equal(rownames(cucTables[[1]]), rownames(cucTables[[2]]))
+
+cucTables <- do.call(cbind, cucTables)
+cucTables <- as.data.frame(cucTables, stringsAsFactors=FALSE)
+cucTables$ID <- rownames(cucTables)
+
+sigNodes <- nodes(compareCUCRomer@mainGraph)
+nodeMembership <- unlist(nodeData(compareCUCRomer@mainGraph, sigNodes, "listMembership"))
+
+cucTables <- cucTables[sigNodes,]
+cucTables$membership <- nodeMembership
+cucTables$description <- unlist(nodeData(compareCUCRomer@mainGraph, sigNodes, "Desc"))
+
+groupLabel <- "nucleotide and nucleoside metabolism - UC.Up,CROHNS.Up"
+allLabel <- sapply(cucNodeGroups, function(x){x$descStr})
+whichLabel <- which(allLabel %in% groupLabel)
+groupNodes <- cucNodeGroups[[whichLabel]]$nodes
+cucTables[groupNodes,]
+
+
+## ----checkRanks----------------------------------------------------------
+newOrder <- order(cucTables[, 'UC.Up'], cucTables[, 'CROHNS.Up'], cucTables[, 'membership'])
+cucTables <- cucTables[newOrder,]
+which(cucTables$ID %in% groupNodes)
+
+

### 2014-02-15 inst/doc/skin_muscle_preprocessing.R
actually moved updated vignettes

@@ -427,3 +427,8 @@ for (iNode in seq(1:length(smNodes))){
 ## save(muscleGeneSetTest, t1, t2, file="inst/data/muscleGeneSetTest.RData")
 
 
+## ------------------------------------------------------------------------
+Sys.time()
+sessionInfo()
+
+

### 2014-02-15 inst/doc/uc_crohns_analysis.R

@@ -288,3 +288,8 @@ cucTables <- cucTables[newOrder,]
 which(cucTables$ID %in% groupNodes)
 
 
+## ------------------------------------------------------------------------
+Sys.time()
+sessionInfo()
+
+

### 2014-02-19 inst/doc/skin_muscle_preprocessing.Rmd
added a check on collateral sprouting to the muscle skin vignette, and regenerated all vignettes

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Skin vs Muscle}
%\VignetteDepends{ccPaper, affy, GEOquery, limma, rat2302.db, org.Rn.eg.db, mouse4302.db, org.Mm.eg.db, GO.db}
-->

```{r customCSS, include=FALSE}
cssFile <- system.file("extdata", "style.css", package="ccPaper")
options(markdown.HTML.stylesheet = cssFile)
```

```{r loadLibrary, message=FALSE}
library(ccPaper)
```

# Muscle Skin Data processing

How do we process the data for muscle and skin to something we can start running through `categoryCompare`?

```{r loadLibraries}
options(stringsAsFactors=FALSE)
library(affy)
```

## Skin

Get the raw data in (this should actually be on GEO, but doesn't seem to be there yet)

```{r skinRaw, eval=FALSE}
currLoc <- getwd()
dataLoc <- "/mlab/data/rmflight/Documents/projects/work/Petruska/drg_sc_skin_denervation/Skin/"
setwd(dataLoc)

attFile <- "Sample information file.txt"
dataAttributes <- read.table(attFile, sep="\t", header=T)
dataAttributes$timePoint <- 0
dataAttributes$timePoint[(dataAttributes$Time == "7 day")] <- 7
dataAttributes$timePoint[(dataAttributes$Time == "14 day")] <- 14

celData <- ReadAffy(phenoData=dataAttributes)
skinData <- rma(celData)

setwd(currLoc)
.sessionInfo <- sessionInfo()
.timeDate <- Sys.time()
save(skinData, .sessionInfo, .timeDate, file="inst/data/skin_rma_data.RData")
```

## Muscle

```{r muscleRaw, eval=FALSE}
library(GEOquery)
muscleData <- getGEO(GEO="GSE4411")[[1]]

controlSamples <- grep("Control", pData(muscleData)$title)
muscleData <- muscleData[,controlSamples]

tmpP <- pData(muscleData)
tmpP$innervation <- "innervated"
denLoc <- grep("Denervated", tmpP$title)
tmpP$innervation[denLoc] <- "denervated"
pData(muscleData) <- tmpP
.sessionInfo <- sessionInfo()
.timeDate <- Sys.time()
save(muscleData, .sessionInfo, .timeDate, file="inst/data/muscle_data.RData")
```

# Gene Lists

For both the skin and muscle data sets, we will calculate the p-values using `limma`. We will probably attempt to do both hypergeometric enrichment and GSEA.

Do not forget that these are from different organisms. **Skin**: rat, **muscle**: mouse.

## Skin Gene Lists and Enrichment

```{r diffSkin}
library(limma)
library(rat2302.db)
data(skin_rma_data)

skinComps <- c("T7 - T0", "T14 - T0")
geneID <- unlist(mget(featureNames(skinData), rat2302ENTREZID))

skinExpr <- exprs(skinData)
skinCollapse <- collapseProbes(skinExpr, geneID) # collapse to single genes using median of expression
skinCharacter <- pData(skinData)
skinCharacter$timePoint2 <- paste("T", skinCharacter$timePoint, sep="")

skinFC <- rankGenes(skinCollapse, skinCharacter$timePoint2, skinComps, doAggregation=FALSE, aggregateIndex=NA)
names(skinFC) <- c("T7", "T14")

skinDiff <- lapply(skinFC, getDiffGenes, id="id")
```

```{r setupGeneSets}
library(GO.db)
library(org.Rn.eg.db)
library(limma)
rnGO <- as.list(org.Rn.egGO2ALLEGS)
rnGO <- rnGO[(Ontology(names(rnGO)) == "BP")]
rnGO <- lapply(rnGO, unique)

skinSet <- symbols2indices(rnGO, rownames(skinCollapse))
```

```{r skinContrasts, eval=FALSE}
skinStatus <- skinCharacter$timePoint2
f <- factor(skinStatus)
skinDesign <- model.matrix(~0 + f)
colnames(skinDesign) <- levels(f)

skinContrast <- makeContrasts(contrasts=skinComps, levels=skinDesign)

save(list=ls(), file="runSkinRomer.RData")
```

```{r skinRomer, eval=FALSE}
load("runSkinRomer.RData")
options(mc.cores=12) # at least if we are on hera
t1 <- Sys.time()
skinRomer <- multicontrastRomer(skinSet, skinCollapse, skinDesign, skinContrast, nrot=10000)
names(skinRomer) <- c("T7", "T14")
save(skinRomer, file="inst/data/skinRomer.RData")
t2 <- Sys.time()
difftime(t2, t1)
```

Run using `geneSetTest`

```{r runGeneSetTest, eval=FALSE}
load("runSkinRomer.RData")
options(mc.cores=12) # at least if we are on hera
t1 <- Sys.time()
skinGeneSetTest <- lapply(skinFC, function(x){
  mmGeneSetTest(skinSet, x$t)
})

t2 <- Sys.time()
difftime(t2, t1)
save(skinGeneSetTest, t1, t2, file="inst/data/skinGeneSetTest.RData")
```

## Muscle Gene Lists and Enrichment

```{r diffMuscle}
library(limma)
library(mouse4302.db)
data(muscle_data)

muscleComps <- c("denervated - innervated")
geneID <- unlist(mget(featureNames(muscleData), mouse4302ENTREZID))

muscleExpr <- exprs(muscleData)
muscleCollapse <- collapseProbes(muscleExpr, geneID) # collapse to single genes using median of expression
muscleCharacter <- pData(muscleData)

muscleFC <- rankGenes(muscleCollapse, muscleCharacter$innervation, muscleComps, doAggregation=FALSE, aggregateIndex=NA)
names(muscleFC) <- "denervation"

muscleDiff <- lapply(muscleFC, getDiffGenes, id="id")
```

```{r setupGeneSetsMuscle}
library(GO.db)
library(org.Mm.eg.db)
library(limma)
mmGO <- as.list(org.Mm.egGO2ALLEGS)
mmGO <- mmGO[(Ontology(names(mmGO)) == "BP")]
mmGO <- lapply(mmGO, unique)

muscleSet <- symbols2indices(mmGO, rownames(muscleCollapse))
```

```{r muscleContrasts, eval=FALSE}
muscleStatus <- muscleCharacter$innervation
f <- factor(muscleStatus)
muscleDesign <- model.matrix(~0 + f)
colnames(muscleDesign) <- levels(f)

muscleContrast <- makeContrasts(contrasts=muscleComps, levels=muscleDesign)

save(muscleSet, muscleCollapse, muscleDesign, muscleContrast, mmGO, muscleFC, file="runMuscleRomer.RData")
```

```{r muscleRomer, eval=FALSE}
load("runMuscleRomer.RData")
options(mc.cores=12) # at least if we are on hera
t1 <- Sys.time()
muscleRomer <- multicontrastRomer(muscleSet, muscleCollapse, muscleDesign, muscleContrast, nrot=10000)
names(muscleRomer) <- "denervation"
save(muscleRomer, file="inst/data/muscleRomer.RData")
t2 <- Sys.time()
difftime(t2, t1)
```

## Compare Skin and Muscle

```{r getDataOut}
data(muscleRomer)
data(skinRomer)

bothGO <- union(names(rnGO), names(mmGO))
combMapping <- lapply(bothGO, function(inName){
  isMM <- inName %in% names(mmGO)
  isRN <- inName %in% names(rnGO)

  tmpGene <- character(0)

  if (isMM){
    tmpGene <- c(tmpGene, mmGO[[inName]])
  }
  if (isRN){
    tmpGene <- c(tmpGene, rnGO[[inName]])
  }
  tmpGene <- unique(tmpGene)
  return(tmpGene)
})

names(combMapping) <- bothGO

geneAnnMapping <- new("namedList", .Data=combMapping, names=names(combMapping))

getSigID <- function(inRomer, pCut=0.05, whichCol=c("Up", "Down")){
  sigID <- lapply(whichCol, function(inCol){
    rownames(inRomer[(inRomer[, inCol] <= pCut),])
  })
  names(sigID) <- whichCol
  return(sigID)
}

muscleSigRomer <- lapply(muscleRomer, getSigID, pCut=0.01)
muscleSigRomer <- unlist(muscleSigRomer, recursive=FALSE)
names(muscleSigRomer) <- c("muscle.Up", "muscle.Down")

skinSigRomer <- lapply(skinRomer, getSigID, pCut = 0.01)
skinSigRomer <- unlist(skinSigRomer, recursive=FALSE)

allSigRomer <- c(skinSigRomer, muscleSigRomer)

genCCSigList <- function(inSig){
  tmp <- new("ccSigList", sigID=inSig)
}

allCCSig <- lapply(allSigRomer, genCCSigList)

allCCRomer <- new("GENccEnrichResult", allCCSig, categoryName="GSEAGO", geneAnnMapping=geneAnnMapping, overlapType="overlap", annDescription=Term(names(geneAnnMapping)))

allRomerOpts <- new("ccOptions", listNames=names(allCCRomer), colorType="pie")
compareSMRomer <- ccCompare(allCCRomer, allRomerOpts)
compareSMRomer
```

```{r saveData, eval=FALSE}
save(compareSMRomer, allRomerOpts, file="skinMuscleCompare.RData")
```

```{r out2cytoscape, eval=FALSE}
compareSMRomer <- breakEdges(compareSMRomer, 0.8)
cwSM <- ccOutCyt(compareSMRomer, allRomerOpts, postText="GSEAGO", rpcPort=9001)
```

```{r saveCytResults, eval=FALSE}
.sessionInfo <- sessionInfo()
.timeDate <- Sys.time()
save(cwSM, compareSMRomer, allRomerOpts, .sessionInfo, .timeDate, file="inst/data/skinMuscleCompareOutput.RData")
```

Get out the groups of related GO terms:

```{r outGOTerms, eval=FALSE}
smNodes <- cytOutNodes("typeII hypersensitivity - T7.Up", cwSM)
smNodes <- cytOutNodes("pH and lysosome regulation - Skin.T7.Down,Skin.T14.Up,Muscle.Up", cwSM, smNodes)
smNodes <- cytOutNodes("Neurotransmitter secretion and transport - Skin.T7.Down,Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("entry into host and movement - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("steriod biosynthesis - Skin.T7.Down", cwSM, smNodes)
smNodes <- cytOutNodes("microtubule organization - Skin.T7.Up,Skin.T14.Down", cwSM, smNodes)
smNodes <- cytOutNodes("cellular respiration - Skin.T14.Down,Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("epithelial tube branching - Skin.T7.Down,Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("rRNA processing - Muscle.Up", cwSM, smNodes)
smNodes <- cytOutNodes("digestion - Skin.T14.Up", cwWM, smNodes)
smNodes <- cytOutNodes("digestion - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("response to virus - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("purine metabolism - Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("protein localization in mitochondrion - Skin.T14.Down", cwSM, smNodes)
smNodes <- cytOutNodes("synapse assembly and nervous system development - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("folic acid compound metabolism - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("retinoic acid biosynthesis - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("mitotic spindle assembly - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("regulation of phospholipase C - Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("acetyl CoA biosynthesis - Skin.T14.Down,Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("response to glucose - Skin.T7.Down, Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("response to VEGF - Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("mitotic cell cycle checkpoint - Skin.T7.Down,Skin.T14.Down,Muscle.Up", cwSM, smNodes)
smNodes <- cytOutNodes("urea cycle - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("response to prostaglandin - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("keratinocyte migration - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("atp biosynthesis - Skin.T7.Down,Skin.T14.Down,Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("membrane repolarization - Muscle.Down", cwSM, smNodes)
smNOdes <- cytOutNodes("muscle contraction and development - Skin.T7.Down,Skin.T14.Down,Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("muscle contraction and development - Skin.T7.Down,Skin.T14.Down,Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("muscle contraction and development - Skin.T7.Down,Skin.T14.Down,Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("blood vessel endothelial cell differentiation - Skin.T7.Up,Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("endocrine regulation of blood pressure - Skin.T7.Down,Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("tube development - Skin.T7.Down,Muscle.Up", cwSM, smNodes)
smNodes <- cytOutNodes("amino acid transport - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("cation channel activity - Skin.T14.Down", cwSM, smNodes)
smNodes <- cytOutNodes("interferon gamma response - Skin.T14.Down", cwSM, smNodes)
smNodes <- cytOutNodes("lung cell differentiation - Skin.T7.Down", cwSM, smNodes)
smNodes <- cytOutNodes("neg regulation of astrocyte diff. - Skin.T7.Down", cwSM, smNodes)
smNodes <- cytOutNodes("fear response - Skin.T7.Down", cwSM, smNodes)
smNodes <- cytOutNodes("N acetylglucosamine metabolism - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("proteoglycan biosynthesis - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("carbohydrate catabolism - Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("response to muscle activity - Skin.T14.Down,Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("response to indole 3 methanol - Muscle.Up", cwSM, smNodes)
smNodes <- cytOutNodes("negative regulation of response to granulocyte / myeloid cell diff. - Skin.T7.Up,Muscle.Up", cwSM, smNodes)
smNodes <- cytOutNodes("vitamin metabolism - Skin.T14.Up,Muscle.Up", cwSM, smNodes)
smNodes <- cytOutNodes("DNA break repair - Muscle.Up", cwSM, smNodes)
smNodes <- cytOutNodes("calcineurin NFAT signaling cascade - Skin.T14.Down", cwSM, smNodes)
smNodes <- cytOutNodes("histone demethylation - Skin.T14.Down", cwSM, smNodes)
smNodes <- cytOutNodes("negative regulation of Ras signal transduction - Skin.T14.Down", cwSM, smNodes)
smNodes <- cytOutNodes("positive regulation of metalloenzyme activity - Skin.T14.Down", cwSM, smNodes)
smNodes <- cytOutNodes("glucocorticoid receptor signaling pathway - Skin.T14.Down", cwSM, smNodes)
smNodes <- cytOutNodes("mesenchymal cell diff in kidney / renal system dev. - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("axon extension - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("lung lobe dev. and morphogenesis - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("embryonic digestive tract dev. and morphogenesis - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("negative regulation of protein transport - Skin.T7.Up,Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("phosphate ion transport - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("fluid transport - Skin.T7.Down,Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("fluid secretion - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("interleukin 13 production - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("glia guided migration - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("central nervous system maturation - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("bile acid transport - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("positive regulation of muscle cell apoptosis - Skin.T14.Up", cwSM, smNodes)
smNodes <- cytOutNodes("atrioventricular valve dev. and morphogenesis - Skin.T14.Up,Muscle.Down", cwSM, smNodes)
smNodes <- cytOutNodes("dopamine transport - Skin.T7.Down", cwSM, smNodes)
smNodes <- cytOutNodes("collateral sprouting - Skin.T7.Down", cwSM, smNodes)
smNodes <- cytOutNodes("response to osmotic stress - Skin.T7.Down", cwSM, smNodes)
smNodes <- cytOutNodes("positive regulation of epidermal growth factor signaling - Skin.T7.Down", cwSM, smNodes)
smNodes <- cytOutNodes("response to leptin - Skin.T7.Down", cwSM, smNodes)
smNodes <- cytOutNodes("catecholamin biosynthesis - Skin.T7.Down", cwSM, smNodes)
smNodes <- cytOutNodes("negative regulation of cell junction assembly - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("endocardial cell differentiation - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("sequestering of actin monomers - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("beta amyloid formation - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("response to platelet derived growth factor stimulus - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("lipopolysaccharid biosynthesis - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("aminoglycan metabolism - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("epithelial to mesenchymal transition, endocardial cushion formation - Skin.T7.Up", cwSM, smNodes)
smNodes <- cytOutNodes("pulmonary valve dev. and morphogenesis - Skin.T7.Up", cwSM, smNodes)
.sessionInfo <- sessionInfo()
.timeDate <- Sys.time()
save(smNodes, cwSM, allCCRomer, allRomerOpts, compareSMRomer, file="inst/data/skinMuscleCCOutput.RData")
```

Lets generate a couple of different tables for output.

```{r genTableOutput}
data(skinMuscleCCOutput)

allDescStrings <- sapply(smNodes, function(x){x$descStr})
string2List <- strsplit(allDescStrings, " - ", fixed=TRUE)
justDesc <- sapply(string2List, function(x){x[1]})
listMem <- sapply(string2List, function(x){x[2]})
listMem[1] <- "Skin.T7.Up" # fix an error in the initial setup
listMemSplit <- strsplit(listMem, ",", fixed=TRUE)

descMembershipTable <- matrix("", nrow=length(allDescStrings), ncol=7)
colnames(descMembershipTable) <- c("Description", "Skin.T7.Down", "Skin.T7.Up", "Skin.T14.Down", "Skin.T14.Up", "Muscle.Down", "Muscle.Up")

descMembershipTable[,"Description"] <- justDesc

for (inRow in seq(1, nrow(descMembershipTable))){
  useSplit <- listMemSplit[[inRow]]
  trimSplit <- gsub(" ", "", useSplit)
  descMembershipTable[inRow, trimSplit] <- "X"
}

orderBy <- c("Skin.T7.Down,Skin.T14.Down,Muscle.Down",
             "Skin.T7.Down,Skin.T14.Down,Muscle.Up",
             "Skin.T7.Down,Skin.T14.Down",
             "Skin.T7.Down,Skin.T14.Up,Muscle.Up",
             "Skin.T7.Down,Skin.T14.Up",
             "Skin.T7.Down,Muscle.Up",
             "Skin.T7.Down,Muscle.Down",
             "Skin.T7.Down",
             "Skin.T7.Up,Skin.T14.Down",
             "Skin.T7.Up,Skin.T14.Up",
             "Skin.T7.Up,Muscle.Up",
             "Skin.T7.Up",
             "Skin.T14.Down,Muscle.Down",
             "Skin.T14.Down",
             "Skin.T14.Up,Muscle.Down",
             "Skin.T14.Up,Muscle.Up",
             "Skin.T14.Up",
             "Muscle.Down",
             "Muscle.Up")

listMem <- gsub(" ", "", listMem)
newOrder <- unlist(lapply(orderBy, function(x){
  which(listMem %in% x)
}))

descMembershipTable <- descMembershipTable[newOrder,]
require(xtable)

# add an html link to each entry in the table
useLink <- paste('<a href="#loc', seq(1, nrow(descMembershipTable)), '">', descMembershipTable[,"Description"], '</a>', sep="")
descMembershipTable[, "Description"] <- useLink
```

Print out the table.

```{r printTable, echo=FALSE, results='asis'}
cat('<a name="tableLink"></a>\n') # a link back to the table
print(xtable(descMembershipTable), type = 'html', html.table.attributes = 'style="border-spacing:20px 5px;"', include.rownames=FALSE, sanitize.text.function=function(x){x})
```

We also want the set of GO terms, their descriptions, and list memberships so that the inquisitive mind can see how we generated the groups.

```{r detailTable, echo=FALSE, results='asis'}
smNodes <- smNodes[newOrder]

for (iNode in seq(1:length(smNodes))){
  nodeSet <- smNodes[[iNode]]
  toLink <- paste('<a name="loc', iNode, '"></a>', sep="")
  cat(toLink, "\n")
  cat("####", nodeSet$descStr, sep=" ")
  cat("\n")

  nGO <- length(nodeSet$nodes)
  outMatrix <- matrix("", nrow=nGO, ncol=8)
  colnames(outMatrix) <- c("GOID", "Description", "T7.Down", "T7.Up", "T14.Down", "T14.Up", "muscle.Down", "muscle.Up")
  rownames(outMatrix) <- nodeSet$nodes
  outMatrix[,"GOID"] <- nodeSet$nodes

  for (useGO in nodeSet$nodes){
    outMatrix[useGO, "Description"] <- nodeSet$nodeData[[useGO]]$Desc
    listMems <- strsplit(nodeSet$nodeData[[useGO]]$listMembership, ",")[[1]]
    listMems <- listMems[(nchar(listMems) > 0)]
    outMatrix[useGO, listMems] <- "X"
  }
  allMems <- sapply(nodeSet$nodeData, function(x){x$listMembership})
  reOrder <- order(allMems, decreasing=TRUE)
  outMatrix <- outMatrix[reOrder,]
  print(xtable(outMatrix), type = 'html', html.table.attributes = 'style="border-spacing:20px 5px;"', include.rownames=FALSE)
  cat('\n<a href="#tableLink">back</a>\n')
}
```

### Collateral Sprouting

One of our groups above seems to be related to **collateral sprouting**, which is the process the skin data was actually seeking to measure, but it is in an unexpected direction. We should look at the genes involved and see what is actually going on.

```{r checkCollateralSprouting}
allDescStrings <- sapply(smNodes, function(x){x$descStr})
isCollateral <- grep("collateral sprouting", allDescStrings, ignore.case=TRUE)

collateralGO <- smNodes[[isCollateral]]$nodes

skinRomer$T7[collateralGO,]
skinRomer$T14[collateralGO,]

rnGenes <- unique(unlist(rnGO[collateralGO]))

rnFCGenes <- rownames(skinFC$T7)
rnGenes <- rnGenes[(rnGenes %in% rnFCGenes)]
rnExtra <- cbind(unlist(mget(rnGenes, org.Rn.egSYMBOL)), unlist(mget(rnGenes, org.Rn.egGENENAME)))

rnData <- lapply(skinFC, function(inFC){
  tmp <- cbind(inFC[rnGenes,], rnExtra)
  names(tmp) <- c(names(tmp)[1:7], "Symbol", "Name")
  return(tmp)
})
lapply(rnData, dim)
```

Lets actually generate tables for T7 and T14.

```{r generateTables, echo=FALSE, results='asis'}
for (iTable in 1:length(rnData)){
  tmpData <- rnData[[iTable]]
  negEntries <- which(tmpData$logFC <= 0)
  htmlOut <- capture.output(print(xtable(tmpData), type = 'html', html.table.attributes = 'style="border-spacing:20px 5px;"', include.rownames=FALSE))
  negEntries <- negEntries + 4
  for (iNeg in negEntries){
    htmlOut[iNeg] <- sub("<TR>", '<TR style="color: red;">', htmlOut[iNeg])
  }

  cat('###', names(rnData)[iTable], sep=' ')
  cat('\n\n')
  cat(htmlOut, sep="\n")

  cat('\n')
}
```

Run using `geneSetTest`

```{r runGeneSetTestMuscle, eval=FALSE}
load("runMuscleRomer.RData")
options(mc.cores=12) # at least if we are on hera
t1 <- Sys.time()
muscleGeneSetTest <- lapply(muscleFC, function(x){
  mmGeneSetTest(muscleSet, x$t)
})

t2 <- Sys.time()
difftime(t2, t1)
save(muscleGeneSetTest, t1, t2, file="inst/data/muscleGeneSetTest.RData")
```

### Date and System information

```{r}
Sys.time()
sessionInfo()
```

### 2014-02-19 inst/doc/skin_muscle_preprocessing.R

@@ -414,6 +414,47 @@ for (iNode in seq(1:length(smNodes))){
 }
 
 
+## ----checkCollateralSprouting--------------------------------------------
+allDescStrings <- sapply(smNodes, function(x){x$descStr})
+isCollateral <- grep("collateral sprouting", allDescStrings, ignore.case=TRUE)
+
+collateralGO <- smNodes[[isCollateral]]$nodes
+
+skinRomer$T7[collateralGO,]
+skinRomer$T14[collateralGO,]
+
+rnGenes <- unique(unlist(rnGO[collateralGO]))
+
+rnFCGenes <- rownames(skinFC$T7)
+rnGenes <- rnGenes[(rnGenes %in% rnFCGenes)]
+rnExtra <- cbind(unlist(mget(rnGenes, org.Rn.egSYMBOL)), unlist(mget(rnGenes, org.Rn.egGENENAME)))
+
+rnData <- lapply(skinFC, function(inFC){
+  tmp <- cbind(inFC[rnGenes,], rnExtra)
+  names(tmp) <- c(names(tmp)[1:7], "Symbol", "Name")
+  return(tmp)
+})
+lapply(rnData, dim)
+
+
+## ----generateTables, echo=FALSE, results='asis'--------------------------
+for (iTable in 1:length(rnData)){
+  tmpData <- rnData[[iTable]]
+  negEntries <- which(tmpData$logFC <= 0)
+  htmlOut <- capture.output(print(xtable(tmpData), type = 'html', html.table.attributes = 'style="border-spacing:20px 5px;"', include.rownames=FALSE))
+  negEntries <- negEntries + 4
+  for (iNeg in negEntries){
+    htmlOut[iNeg] <- sub("<TR>", '<TR style="color: red;">', htmlOut[iNeg])
+  }
+  
+  cat('###', names(rnData)[iTable], sep=' ')
+  cat('\n\n')
+  cat(htmlOut, sep="\n")
+  
+  cat('\n')
+}
+
+
 ## ----runGeneSetTestMuscle, eval=FALSE------------------------------------
 ## load("runMuscleRomer.RData")
 ## options(mc.cores=12) # at least if we are on hera
